---
title: "1. Feature counts quantiles Normalization"
author: "xue"
date: "2024-02-27"
output: html_document
---

This document 1) normalize the raw bulk-ATAC-seq reads, 2) generate the normalization factor used in downstream DESeq2 analysis, and 3) generate some diagnostic plots.

# 1a. Setup environment

## Load packages

```{r}
library(ChIPQC)
library(rtracklayer)
library(preprocessCore)
library(Rsubread)
library(tidyverse)
library(reshape2)
library(ggpubr)
library(genomation)
```

## Set working directory

```{r, setup, include=FALSE}
wd = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/differential/geniml_univ_cc_073126"
knitr::opts_knit$set(root.dir = wd)
```

## Define file paths

### Set input directory

```{r}
# maxinum likelihood universe
univ = "/home/bx2ur/code/reninness_score/data/universe_073126/universe_cc_073126.bed"

# for renin vs non-renin
meta = "~/code/renin_atac/metadata/renin_encode_atac.csv"

# for WT renin vs recruited renin 
# meta = "~/code/renin_atac/metadata/native_recruited.csv"

# mm10 black list
blklist = "~/code/renin_atac/data/mm10.blacklist.bed.gz"
```

### Set output directory

```{r}
primary_dir = "primary"
tumoral_dir = "tumoral"

recruited_dir = "recruited"
```

## Load input files

```{r}
# meta data
samples <- read.table(meta, header = TRUE, sep = ",")
samples <- samples[(samples$Group != "Renin_Low"), ] # remove Renin_Low samples

# for renin primary kidney cell vs encode
 samples <- samples[(samples$Tissue != "As4.1"), ]
# for renin as4.1 vs encode
# samples <- samples[(samples$Tissue == "As4.1" | samples$Group == "Non-Renin_Control"), ]

# get sample names
sample_name <- list(samples$SampleID)[[1]]

# list of peak files
peaks <- list(samples$Peaks)
peak_samples <- list(samples$SampleID)[[1]]

# separate single and paired reads bam files
bams.files <- list(samples[(samples$read_type == "single"), "bamReads"])
bamp.files <- list(samples[(samples$read_type == "paired"), "bamReads"])
```

# 1b. Get consensus peaks and count features

# Define functions

```{r}
# GET CONSENSSUS PEAKS WITH GIVEN UNIVERSE
getConsensusPeaks <- function(peaks, univ) {
  myPeaks <- lapply(peaks[[1]], ChIPQC:::GetGRanges, simple = TRUE)
  names(myPeaks) <- peak_samples
  myGRangesList <- GRangesList(myPeaks)
  
  reduced <- readGeneric(univ, chr = 1, start = 2, end = 3,
                         keep.all.metadata = TRUE, zero.based = FALSE,
                         remove.unusual = FALSE, header = TRUE, 
                         skip = 0, sep = "\t")
  
  consensusIDs <- paste0("consensus_", seq(1, length(reduced)))
  mcols(reduced) <- do.call(cbind, lapply(myGRangesList, 
                                          function(x) (reduced %over% x) + 0))
  
  reducedConsensus <- reduced
  mcols(reducedConsensus) <- cbind(as.data.frame(mcols(reducedConsensus), 
                                                 optional = TRUE, 
                                                 make.names = FALSE),
                                   consensusIDs)
  consensusIDs <- paste0("consensus_", seq(1, length(reducedConsensus)))
  
  blklist <- import.bed(blklist)
  # remove blk list
  reducedConsensus <- reducedConsensus[!reducedConsensus %over% blklist & 
                                         !seqnames(reducedConsensus) %in% "chrM"]
  
  regionsToCount <- data.frame(
    GeneID = paste(seqnames(reducedConsensus),
                   start(reducedConsensus),
                   end(reducedConsensus), sep = "_"),
    Chr    = seqnames(reducedConsensus),
    Start  = start(reducedConsensus),
    End    = end(reducedConsensus),
    Strand = strand(reducedConsensus)
  )
  
  return(list(reducedConsensus, regionsToCount))
}

# FEATURE COUNTS & QUANTILES NORMALIZATION

countFeatures <- function(bams, bamp, regionsToCount, nthreads = 2){
  # feature counts for singleton
  fcs <- featureCounts(bams[[1]],
                       annot.ext = regionsToCount,
                       nthreads = nthreads)
  # feature counts for paired end 
  fcp <- featureCounts(bamp[[1]], 
                       annot.ext = regionsToCount, 
                       isPairedEnd = TRUE,  
                       nthreads = nthreads)
  # joint matrix
  fc <- cbind(fcp$counts, fcs$counts)
  #rename columns
  colnames(fc) <- sub("_sort_dedup.bam", "", colnames(fc))
  fc <- fc[, samples$SampleID]
  
  # quantiles normalization
  fc_norm <- normalize.quantiles(fc, copy = TRUE)
  colnames(fc_norm) <- colnames(fc)
  rownames(fc_norm) <- rownames(fc)
  
  # get normalization factors
  nfactors <- fc / fc_norm
  nfactors[nfactors == 0] <- 1
  nfactors[is.na(nfactors)] <- 1
  
  return(list(fc, fc_norm, nfactors))
}

```

## Run consensus peaks & feature counting

```{r}
# get consensus peaks
output = getConsensusPeaks(peaks, univ)
reducedConsensus = output[[1]]
reducedConsensus = reducedConsensus[, sample_name]
regionsToCount = output[[2]]

# counting features & normalizing counts
output = countFeatures(bams.files, bamp.files, regionsToCount, nthreads = 8 )
fc = output[[1]]           # pre normalized counts
fc_norm = output[[2]]      # normalized counts
nfactors = output[[3]]     # normalization factors

df <- data.frame(rownames(fc_norm))
df %>% 
  separate(rownames.fc_norm., into = c('chr', 'start', 'end'), sep="_(?=[^_]+$)")
df[c('chr', 'start', 'end')] <- str_split_fixed(df$rownames.fc_norm, "_", 3)
df = df[c('chr', 'start', 'end', 'rownames.fc_norm.')]

write.table( x =df,
             file = 'ConsensusPeaks.bed', sep="\t",
             quote=FALSE, col.names = FALSE, row.names = FALSE)

save.image("renin_encode_atac_primary.RData")
# save.image("renin_encode_atac_tumoral.RData")
```
```{r}
# save normalized feature count matrix
bed <- file.path(wd, "bed/renin_specific_regions.bed")

regions <- read.table(bed, sep="\t", header=FALSE,
                      stringsAsFactors = FALSE)$V4
keep_ <- regions[regions %in% rownames(fc_norm)]
fc_sub <- fc_norm[keep_,,drop=FALSE]
fc_t <- t(fc_sub)
out <- data.frame(peak_id = rownames(fc_sub), fc_sub,
                  check.names = FALSE, row.names = NULL)
write.csv(out,
          file=file.path(wd,paste0("fc_norm_renin_specifc_", Sys.Date(),".csv")),
          quote = FALSE, row.names = FALSE)
```

# 1c. Visualization

## Define plotting functions

```{r}
# MELT FEATURE COUNT MATIX
meltFC <- function(fc, peak_range, renin_samples) {
  fc_melt <- melt(fc[peak_range, ])
  names(fc_melt) <- c("consensus_peak_ID", "sample_name", "counts")

  fc_melt <- fc_melt %>%
    mutate(consensus_peak_ID = factor(consensus_peak_ID,
                                      levels = rev(sort(unique(consensus_peak_ID))))) %>%
    mutate(countfactor = cut(counts,
                             breaks = c(-1, 0, 1, 10, 100, 500, 1000, max(counts, na.rm = TRUE)),
                             labels = c("0", "0-1", "1-10", "10-100",
                                        "100-500", "500-1000", ">1000"))) %>%
    mutate(countfactor = factor(as.character(countfactor),
                                levels = rev(levels(countfactor)))) %>%
    mutate(Group = factor(ifelse(sample_name %in% renin_samples, "Renin", "Non-Renin_Control"),
                          levels = c("Renin", "Non-Renin_Control")))

  pval_df <- data.frame(consensus_peak_ID = rownames(fc[peak_range, ]))
  for (row in 1:nrow(pval_df)) {
    pval <- t.test(
      fc_melt[fc_melt$consensus_peak_ID == pval_df$consensus_peak_ID[row] &
                fc_melt$Group == "Renin", "counts"],
      fc_melt[fc_melt$consensus_peak_ID == pval_df$consensus_peak_ID[row] &
                fc_melt$Group == "Non-Renin_Control", "counts"],
      var.equal = TRUE
    )
    pval_df[row, "pvals"] <- pval$p.value
    if (pval$p.value > 0.05) {
      pval_df[row, "sig"] <- "> 0.05"
      pval_df[row, "color"] <- "azure1"
    } else if (pval$p.value >= 0.01) {
      pval_df[row, "sig"] <- "< 0.05"
      pval_df[row, "color"] <- "azure2"
    } else if (pval$p.value >= 0.001) {
      pval_df[row, "sig"] <- "< 0.01"
      pval_df[row, "color"] <- "azure3"
    } else {
      pval_df[row, "sig"] <- "< 0.001"
      pval_df[row, "color"] <- "azure4"
    }
  }

  fc_melt <- merge(fc_melt, pval_df, by = "consensus_peak_ID")
  return(fc_melt)
}

plotHeatmap <- function(fc, peak_range, renin_samples, myDesign, normalization, output_dir) {
  fc_melt <- meltFC(fc, peak_range, renin_samples)
  title <- ifelse(normalization,
                  paste(myDesign, "Normalized Counts"),
                  paste(myDesign, "Non-normalized Counts"))
  textcol <- "black"

  plot <- ggplot(fc_melt, aes(consensus_peak_ID, sample_name)) +
    geom_tile(mapping = aes(fill = countfactor, width = 3, height = 3)) +
    facet_grid(Group ~ ., space = "free", scales = "free_y", switch = "y") +
    labs(title = title) +
    guides(fill = guide_legend(title = "Counts")) +
    scale_x_discrete(limits = rev, guide = guide_axis(check.overlap = TRUE)) +
    scale_y_discrete(guide = guide_axis(check.overlap = TRUE)) +
    ylab("samples") +
    xlab("") +
    scale_fill_manual(
      values = c("#d53e4f", "#f46d43", "#fdae61", "#fee08b",
                 "#e6f598", "#abdda4", "#ddf1da"),
      na.value = "grey90"
    ) +
    theme_grey(base_size = 10) +
    theme(
      legend.position = "right",
      legend.direction = "vertical",
      legend.title = element_text(colour = textcol),
      legend.margin = margin(grid::unit(0, "cm")),
      legend.text = element_text(colour = textcol, size = 6),
      legend.key.height = grid::unit(0.5, "cm"),
      legend.key.width = grid::unit(0.2, "cm"),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5,
                                 size = 8, colour = textcol),
      axis.ticks = element_line(size = 0.1),
      plot.background = element_blank(),
      panel.border = element_blank(),
      plot.margin = margin(0.7, 0.4, 0.1, 0.2, "cm"),
      plot.title = element_text(colour = textcol, hjust = 0, size = 10)
    )

  ggsave(paste0(output_dir, "/fig/", title, ".svg"),
         plot = plot, width = 174, height = 174, units = "mm")

  # plot
}

plotPeaks <- function(fc, peak_range, renin_samples, myDesign, normalization, output_dir) {
  fc_melt <- meltFC(fc, peak_range, renin_samples)
  title <- ifelse(normalization,
                  paste(myDesign, "Normalized Counts"),
                  paste(myDesign, "Non-normalized Counts"))

  fc_melt$consensus_peak_ID <- sub("ID_", "", fc_melt$consensus_peak_ID)
  fc_melt <- cbind(fc_melt, read.csv(text = fc_melt$consensus_peak_ID,
                                     header = FALSE, sep = "_",
                                     col.names = c("Chrom", "start", "end")))

  df <- fc_melt %>%
    group_by(Chrom, start, end, Group) %>%
    summarise_at(vars(counts), list(avg_counts = mean))

  pval_df <- unique(fc_melt[, c("Chrom", "start", "end", "pvals", "sig", "color")])
  fc_melt <- merge(df, pval_df, by = c("Chrom", "start", "end"))
  fc_melt <- fc_melt[order(fc_melt$start), ]

  options(repr.plot.width = 150, repr.plot.height = 50)

  p_peaks <- ggplot() +
    labs(title = title) +
    geom_rect(data = fc_melt,
              mapping = aes(xmin = start, xmax = end,
                            ymin = avg_counts - 25, ymax = avg_counts + 25,
                            fill = Group)) +
    geom_rect(data = fc_melt,
              mapping = aes(xmin = start, xmax = end,
                            ymin = -100, ymax = -50,
                            fill = sig),
              fill = fc_melt$color) +
    scale_x_continuous(limits = c(min(fc_melt$start), max(fc_melt$end)),
                       breaks = scales::pretty_breaks(n = 45)) +
    scale_y_continuous(expand = c(0, 0), limits = c(-100, 1800),
                       breaks = scales::pretty_breaks(n = 10)) +
    xlab(paste0("coordinates: ", fc_melt$Chrom[1])) +
    ylab("Counts") +
    scale_fill_manual(values = c("Renin" = "#FF7F0E", "Non-Renin_Control" = "#1F77B4"))

  plot <- p_peaks +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5, size = 8))

  ggsave(paste0(output_dir, "/fig/", title, " peaks.svg"),
         plot = plot, width = 174, height = 174, units = "mm")

  # plot
}

plotLine <- function(fc, peak_range, renin_samples, myDesign, normalization, output_dir) {
  fc_melt <- meltFC(fc, peak_range, renin_samples)
  title <- ifelse(normalization,
                  paste(myDesign, "Normalized Counts"),
                  paste(myDesign, "Non-normalized Counts"))

  fc_melt$consensus_peak_ID <- sub("ID_", "", fc_melt$consensus_peak_ID)
  fc_melt <- cbind(fc_melt, read.csv(text = fc_melt$consensus_peak_ID,
                                     header = FALSE, sep = "_",
                                     col.names = c("Chrom", "start", "end")))

  df1 <- fc_melt[, c("Chrom", "start", "sample_name", "counts",
                      "countfactor", "Group", "pvals", "sig", "color")]
  df2 <- fc_melt[, c("Chrom", "end", "sample_name", "counts",
                      "countfactor", "Group", "pvals", "sig", "color")]
  df1 <- rename(df1, coordinates = start)
  df2 <- rename(df2, coordinates = end)
  fc_melt <- rbind(df1, df2)

  df <- fc_melt %>%
    group_by(Chrom, coordinates, Group) %>%
    summarise_at(vars(counts), list(avg_counts = mean))

  pval_df <- unique(fc_melt[, c("Chrom", "coordinates", "pvals", "sig", "color")])
  fc_melt <- merge(df, pval_df, by = c("Chrom", "coordinates"))
  fc_melt <- fc_melt[order(fc_melt$coordinates), ]

  options(repr.plot.width = 150, repr.plot.height = 50)

  p_line <- ggplot(fc_melt, aes(color = Group, y = avg_counts, x = coordinates)) +
    labs(title = title) +
    geom_line() +
    scale_x_continuous(limits = c(min(fc_melt$coordinates), max(fc_melt$coordinates)),
                       breaks = scales::pretty_breaks(n = 45)) +
    xlab(paste0("coordinates: ", fc_melt$Chrom[1])) +
    ylab("Counts") +
    scale_color_manual(values = c("Renin" = "#FF7F0E", "Non-Renin_Control" = "#1F77B4"))

  plot <- p_line +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5, size = 8)) +
    geom_rug(color = fc_melt$color, sides = "b", size = 1)

  ggsave(paste0(output_dir, "/fig/", title, " line.svg"),
         plot = plot, width = 174, height = 174, units = "mm")

  # plot
}

plotBox <- function(fc, peak_range, renin_samples, myDesign, normalization, output_dir) {
  fc_melt <- meltFC(fc, peak_range, renin_samples)
  fc_melt$consensus_peak_ID <- sub("ID_", "", fc_melt$consensus_peak_ID)
  title <- ifelse(normalization,
                  paste(myDesign, "Normalized Counts"),
                  paste(myDesign, "Non-normalized Counts"))

  options(repr.plot.width = 150, repr.plot.height = 50)

  p_box <- ggplot(fc_melt, aes(fill = Group, y = counts, x = consensus_peak_ID)) +
    labs(title = title) +
    geom_boxplot(outlier.size = 0.1, lwd = 0.2) +
    scale_x_discrete(limits = rev) +
    xlab("Consensus peaks") +
    ylab("Counts") +
    scale_fill_manual(values = c("Renin" = "#FF7F0E", "Non-Renin_Control" = "#1F77B4"))

  plot <- p_box +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5, size = 8)) +
    geom_rug(color = fc_melt$color, sides = "b", size = 1)

  ggsave(paste0(output_dir, "/fig/", title, " box.svg"),
         plot = plot, width = 174, height = 174, units = "mm")

  # plot
}
```

```{r}
# define my design
# myDesign <- "Renin (primary kidney cell) vs Non-renin control"
myDesign <- "Renin (As4.1 cell line) vs Non-renin control"

# set contract group
Group <- list(samples$Group)[[1]]

# renin_samples <- sample_name[1:13] # for renin primary kidney vs encode
renin_samples <- sample_name[1:6] # for renin tumoral vs encode
# renin_samples <- sample_name[1:19] # for all renin vs encode

# set plot range
# +/- 25 peaks around the ren1 SE 
ren1_chr <- "chr1"
ren1_start <- 133350674
# find the closest consensus peak to the ren1 start site
chr1_idx <- which(regionsToCount$Chr == ren1_chr)
closest_idx <- chr1_idx[which.min(abs(regionsToCount$Start[chr1_idx] - ren1_start))]   
flanking <- 25                                            
peak_range <- seq(closest_idx - flanking, closest_idx + flanking)

# create output fig directory
output_dir <- primary_dir
dir.create(paste0(output_dir, "/fig"), recursive = TRUE, showWarnings = FALSE)

# plot pre-normalized counts
plotHeatmap(fc, peak_range, renin_samples, myDesign, normalization = FALSE, output_dir)
plotBox(fc, peak_range, renin_samples, myDesign, normalization = FALSE, output_dir)
plotLine(fc, peak_range, renin_samples, myDesign, normalization = FALSE, output_dir)
plotPeaks(fc, peak_range, renin_samples, myDesign, normalization = FALSE, output_dir)

# plot normalized counts
plotHeatmap(fc_norm, peak_range, renin_samples, myDesign, normalization = TRUE, output_dir)
plotBox(fc_norm, peak_range, renin_samples, myDesign, normalization = TRUE, output_dir)
plotLine(fc_norm, peak_range, renin_samples, myDesign, normalization = TRUE, output_dir)
plotPeaks(fc_norm, peak_range, renin_samples, myDesign, normalization = TRUE, output_dir)

```
