# ---
title: "4. GO Enrichment Analysis with rGREAT"
author: "xue"
date: "2024-02-27"
output: html_document
---

This document 1) performs the GO enrichment analysis on differentially accessible regions in renin-expressing samples, 2) calculates the weighted rank of renin vs non-renin related regions on renin related GO terms, and 3) generate some diagnostic plots.

# 4a. Setup environment

## Load packages

```{r}
library(rGREAT)
library(ggplot2)
library(data.table)
library(genomation)
# GO.db attaches AnnotationDbi, which masks dplyr::select -- load it before
# dplyr so dplyr's verbs win the masking order
library(GO.db)
library(dplyr)
library(tidyr)   # pivot_wider, for the directional significance marks
```

## Ontology / database versions

Two different GO releases are in play in this document and they do NOT match:

* **GREAT** (queried over the network by `submitGreatJob`) does the enrichment
  test against its own, older, bundled GO release. It returns 13,090 GO
  Biological Process terms for mm10.
* **GO.db** is used locally, in `getReninGOTerms()`, only to expand the anchor
  GO IDs to their descendants.

`1694 / 13090` = **12.9%** of the GO IDs GREAT reports no longer exist in the
installed GO.db. Those terms have no ancestry in GO.db, so anchor expansion
cannot reach them; they are only matched if their ID is listed explicitly in
the alias vector in `getReninGOTerms()`. This is why the obsolete anchor
`GO:0019933` silently emptied the whole cAMP branch. Record the versions below
with any figure generated from this document, and re-check the anchor list
after any Bioconductor upgrade.

```{r}
# versions actually used for the results in this document
cat("R            :", R.version.string, "\n")
for (p in c("rGREAT", "GO.db", "AnnotationDbi", "genomation", "dplyr", "ggplot2")) {
  cat(sprintf("%-13s: %s\n", p, as.character(packageVersion(p))))
}
# GREAT server version used by submitGreatJob(); mm10 is served by GREAT v4
cat("GREAT server : v4 (mm10), ontology release bundled server-side\n")
```

<!--
Versions used for the figures committed alongside this document (2026-08-31):
  R            : 4.6.0 (2026-04-24)
  rGREAT       : 2.14.0        GREAT server: v4, genome mm10
  GO.db        : 3.23.1
  AnnotationDbi: 1.74.0
  genomation   : 1.44.0
  dplyr        : 1.2.1
  ggplot2      : 4.0.3
  GREAT returned 13090 GO BP terms; 1694 (12.9%) are absent from GO.db 3.23.1.
-->

## Set working directory

```{r setup, include=FALSE}
wd = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/differential/geniml_univ_cc_073126"
knitr::opts_knit$set(root.dir = wd)
```

## Define file paths

### Set input path

```{r}
bed_dir = "bed"

bed_paths = list(
  all_up = file.path(bed_dir, "deseq_all_diff_up.bed"),        # all inc accessibility region
  all_down = file.path(bed_dir, "deseq_all_diff_down.bed"),    # all dec accessibility region
  non_differential = file.path(bed_dir, "deseq_non_diff.bed"),          # all non-differential region
  renin = file.path(bed_dir, "renin_specific_regions.bed") # renin specific regions
)
```

### Set output path

```{r}
output_plot = file.path("plot", "GO_enrichment")
dir.create(output_plot, showWarnings = FALSE)
```

## Load input files

```{r}
# Load input BED files
region_set = GRangesList(lapply(bed_paths[1:3], function(f) readGeneric(f)))
 
# Background for enrichment
# (ex. any enrichment specific to WT in reference to all renin-expressing samples)
bg = c(region_set$all_up, region_set$all_down, region_set$non_differential)

# set up grangelist for core renin specific vs other increasing accessibility regions
renin_specific = readGeneric(bed_paths$renin)
other_regions = subsetByOverlaps(region_set$all_up, renin_specific, invert = TRUE)

query = GRangesList(renin_specific, other_regions)
names(query) = c("renin_specific", "other")
```

# 4b. Run GO enrichment analysis

```{r}
# define Functions

getBioEnrichment = function(gr, use_bg="FALSE", bg) {
  job = if (use_bg) submitGreatJob(gr, bg, species = "mm10") else submitGreatJob(gr, species = "mm10")
  tb = getEnrichmentTables(job)
  return(tb$`GO Biological Process`)
}

```

## GO enrichment in increasing, decreasing, and non-differential region sets

```{r}
# run go enrichment analysis
differential_enrichment_results = lapply(names(region_set), function(set) {
  getBioEnrichment(region_set[[set]], use_bg = TRUE, bg = bg)
})
# name the result list
names(differential_enrichment_results) = names(region_set)
```

## GO enrichment in core renin-specific and other increasing accessibiloty regions

```{r}
# run go enrichment analysis
renin_enrichment_results = lapply(names(query), function(set) {
  getBioEnrichment(query[[set]], use_bg = TRUE, bg=bg)
})
# name the result list
names(renin_enrichment_results) = names(query)

```

# 4c. Plot GO enrichment results of differentially accessible regions

```{r}
# define function
plotTopEnrichedGO = function(dfs, n = 20, output) {
  dir.create(output, showWarnings = FALSE, recursive = TRUE)

  lapply(names(dfs), function(name) {
    df = dfs[[name]]
    
    # Filter and rank
    go_bio = subset(df, Hyper_Fold_Enrichment >= 2)
    plot_this = head(go_bio[order(go_bio$Hyper_Adjp_BH), ], n)
    
    if (nrow(plot_this) == 0) {
      message(sprintf("Skipping %s: no terms with Fold Enrichment >= 2", name))
      return(NULL)
    }

    # Shorten overly long term names (optional)
    plot_this$name <- sapply(plot_this$name, function(s) {
      if (nchar(s) > 60) paste0(substr(s, 1, 57), "...") else s
    })

    # Prepare data for plotting
    x = factor(plot_this$name, levels = rev(plot_this$name))  # reverse for top-down
    y = -log10(plot_this$Hyper_Adjp_BH)
    data = data.frame(x, y)

    # Estimate width: 0.25 per bar + padding
    plot_width = max(6, 0.25 * n + 2)

    # Build plot
    p = ggplot(data, aes(x = x, y = y)) +
      geom_bar(stat = "identity", fill = "lightgrey", width = 0.8) +
      coord_flip() +
      labs(title = name, y = "-log10(FDR)", x = "") +
      theme_bw() +
      theme(
        aspect.ratio = 1,
        axis.text.y = element_text(size = 8, colour = "black"),
        axis.text.x = element_text(size = 10, colour = "black"),
        axis.line = element_line(colour = "black"),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        plot.title = element_text(hjust = 1)
      )

    # Save plot
    file_name = file.path(output, paste0("top_enriched_go_", name, ".svg"))
    ggsave(file_name, plot = p, width = plot_width, height = 6)
    message("Saved: ", file_name)

    return(p)
  })
}
```

```{r}
plotTopEnrichedGO(differential_enrichment_results, n = 20, output = output_plot)
```

# 4d. Calculate weighted rank of the renin related GO terms for Renin vs other regions

```{r}
# define function

# filterEnrichment / getTopGOTerms / getReninGOTerms / getRank /
# getRankWeight now live in go_enrichment_functions.R, shared with
# 8_deseq2_vs_starspace_regions.Rmd (8j) so the two cannot drift apart
# again. Edit them there.
source("go_enrichment_functions.R")
```

```{r}
# flag significance on the FULL tables; every term is kept so that ranking in
# getRankWeight sees both region sets
go_results_all = lapply(renin_enrichment_results, function(df) {
  df$sig = !is.na(df$Hyper_Adjp_BH) &
    df$Hyper_Adjp_BH <= 0.1 &
    df$Hyper_Foreground_Gene_Hits >= 3
  df
})

# significance-filtered tables, used only to choose which terms to plot
go_results = lapply(go_results_all, filterEnrichment)

# renin-related GO terms that are significant in at least one region set
my_go_terms = getReninGOTerms(go_results)
```

# 4e. Plot weighted rank of the renin related GO terms: core renin specific region vs other increasing accessibility regions

```{r}
rank_dfs = lapply(names(go_results_all), function(set) {
  getRank(go_results_all[[set]], group = set)
})

names(rank_dfs) = names(go_results_all)

rank_dfs = getRankWeight(rank_dfs, my_go_terms)

plot_this = do.call(rbind, rank_dfs)
plot_this = plot_this[!is.na(plot_this$name), ]
rownames(plot_this) = NULL

# WHICH region set called each term significant -- not just how many. The old
# code counted (`n_sig = sum(sig)`) and marked every count-of-1 with the same
# "*", so a term significant only in the core renin set and a term significant
# only in the other set carried an identical flag. On the 50-term figure that
# was 7 core-only vs 34 other-only under one symbol, i.e. the mark meant the
# OPPOSITE of the intended reading 83% of the time. Direction is now recorded.
sig_by_set = plot_this %>%
  dplyr::select(name, Group, sig) %>%
  tidyr::pivot_wider(names_from = Group, values_from = sig,
                     names_prefix = "sig_") %>%
  dplyr::mutate(
    n_sig = as.integer(sig_renin_specific) + as.integer(sig_other),
    sig_class = dplyr::case_when(
      sig_renin_specific &  sig_other ~ "both",
      sig_renin_specific & !sig_other ~ "core_only",
      !sig_renin_specific & sig_other ~ "other_only",
      TRUE                            ~ "neither"))

# directional marks. dagger = significant in the core renin-specific set only,
# double dagger = significant in the other increased-accessibility set only,
# unmarked = significant in both. Keep these in sync with the plot caption.
MARK_CORE  = "\u2020"   # dagger
MARK_OTHER = "\u2021"   # double dagger
markTerm = function(nm, cls) {
  ifelse(cls == "core_only",  paste0(nm, " ", MARK_CORE),
  ifelse(cls == "other_only", paste0(nm, " ", MARK_OTHER), nm))
}

# NOTE: GO.db attaches AnnotationDbi, which masks dplyr::select with an S4
# generic that has no data.frame method -- qualify the verb explicitly.
data = plot_this %>%
  dplyr::left_join(sig_by_set, by = "name") %>%
  dplyr::select(name, Group, n_sig, sig_class, Rank_weight = rank_weight,
                Fold_rank_weight = rank_fold_weight)

data$Group = factor(data$Group, levels = c("renin_specific", "other"))

# named term_order, not `order`: a variable called `order` shadows base::order,
# which getRank() calls, so re-running these chunks in one session breaks it.
# Keep every term whose weight favours the core renin-specific set, rather than
# a fixed top-N. head(50) was arbitrary and cut terms mid-way through the
# near-tie band; > 0.5 is the point the metric actually means something ("ranked
# higher by fold enrichment in the core set"). Currently yields 37 terms.
WEIGHT_CUTOFF = 0.5
term_order = data %>%
  dplyr::filter(Group == "renin_specific", Rank_weight > WEIGHT_CUTOFF) %>%
  dplyr::arrange(desc(Rank_weight)) %>%
  dplyr::mutate(labels = factor(name))

data_new = dplyr::filter(data, name %in% term_order$name)
new_order = data_new %>%
  dplyr::filter(Group == "renin_specific") %>%
  dplyr::arrange(desc(Rank_weight)) %>%
  dplyr::mutate(labels = markTerm(name, sig_class))

message(sprintf("Fig 2D: %d terms with weight > %.2f | %d significant in the core set (%d of them core-exclusive) | %d significant only in the other set",
        nrow(new_order), WEIGHT_CUTOFF,
        sum(new_order$sig_class %in% c("both", "core_only")),
        sum(new_order$sig_class == "core_only"),
        sum(new_order$sig_class == "other_only")))
```

```{r}
# plot
output_file <- file.path(output_plot, "rank_weight_stacked_bar.svg")
n_terms <- length(unique(data_new$name))
plot_width <- max(6, 0.25 * n_terms + 2)

# Build and save the plot
p <- data_new %>%
  dplyr::mutate(labels = factor(markTerm(name, sig_class), levels = new_order$labels)) %>%
  ggplot(aes(fill = Group, y = Rank_weight, x = labels)) + 
  geom_bar(position = "fill", stat = "identity", width = 0.9) +
  scale_fill_manual(values = c("renin_specific" = "#e1812c", "other" = "lightgrey")) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6, color = "black"),
    axis.text.y = element_text(size = 6, color = "black"),
    axis.title = element_text(size = 8, color = "black"),
    panel.grid = element_blank(),
  ) +
  labs(y = "Enrichment Weight", x = "GO Terms",
       caption = paste0(MARK_CORE, " significant in the core renin-specific regions only    ",
                        MARK_OTHER, " significant in the other increased-accessibility regions only    ",
                        "(unmarked: significant in both)"))

ggsave(output_file,
       plot = p,
       device = svglite::svglite,
       width = max(4, n_terms * 0.1),
       height = 7,
       units = "in")

```
