---
title: "4. De Novo Motif Enrichment Analyses"
author: "xue"
date: "2024-02-27"
output: html_document
---

# 4a. Setup environment

## Load packages

```{r}
library(genomation)
library(GenomicFeatures)
library (BSgenome.Mmusculus.UCSC.mm10)
library(ggplot2)
library(patchwork)
```

## Set working directory

```{r, setup, include=FALSE}
wd = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/differential/geniml_univ_cc"
knitr::opts_knit$set(root.dir = wd)
```

## Define file paths

### Set input directory

```{r}
# non-renin consensus peaks
bg_bed = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/encode_ATAC/summary/ENCODE_ATAC_mm10_consensusPeaks.narrowPeak"

# region set dir 
bed_dir = "bed"

# motif enrichment result
meme_dir =  "meme_res"
```

### Set output directory

```{r}
fa_dir = "fasta"
plot_dir = "plot/motif"

```

# 4b. Generate fasta file from regions sets

```{r}
# define reference genome
genome = BSgenome.Mmusculus.UCSC.mm10
chroms = c("chr1", "chr2","chr3","chr4","chr5",
          "chr6","chr7","chr8", "chr9", "chr10",
          "chr11","chr12","chr13","chr14","chr15",
          "chr16","chr17","chr18","chr19","chrX",
          "chrY")

# define function to loop through all region sets
getFasta = function(input_bed){
  # Modify the file name to change the extension
  name_bed = basename(input_bed)
  name_fa <- sub("\\.(bed|narrowPeak)$", ".fa", name_bed) 
  
  # Construct the output file path
  if (!dir.exists(fa_dir)) {
    dir.create(fa_dir, showWarnings = FALSE, recursive = TRUE)
  }
  output <- file.path(fa_dir, name_fa)
  
  # load peaks
  df <- read.table(input_bed, header = FALSE, sep = "\t")
  colnames(df)[1:3] <- c("chr", "start", "end")

  # Remove rows with missing coordinates
  df <- df[!is.na(df$start) & !is.na(df$end), ]
  
  # Convert to GRanges
  peaks <- makeGRangesFromDataFrame(df, keep.extra.columns = TRUE,
                                    seqnames.field = "chr",
                                    start.field = "start",
                                    end.field = "end",
                                    starts.in.df.are.0based = FALSE)
  peaks = subset(peaks, seqnames %in% chroms )

  ## extract sequence
  seq <- getSeq(genome, peaks)

  ## add names
  df = as.data.frame(peaks)
  name = paste(df$seqnames,":",df$start,"-",df$end, sep="")
  names(seq) <- name
  writeXStringSet(seq, file=output )
}
```

## Generate fasta file for consensus peaks of non-renin background

```{r}
getFasta(bg_bed)
```

## Generate fasta files for region sets of interest in renin samples

```{r}
# Get a list of all files in the input directory
region_sets = list.files(file.path(bed_dir), full.names = TRUE)

# Use lapply to apply the function to each file in motif_sets
lapply(region_sets, getFasta)

```

# 4c. Run de novo motif enrichment analyses with SEA from the MEME suite

```{bash}

bash 6_motif_enrichment.sh
```

# 4d. Plot enriched motifs

```{r}
plotMotifEnrichment = function(set){
  input <- file.path(meme_dir, set, "sea.tsv")
  output <- file.path(plot_dir, paste0(set, "_top20.pdf"))
  tryCatch(
    expr = {
      # load data
      df = read.csv(input, sep="\t", check.names = FALSE)
      df = df[1:(nrow(df) - 3), ]
      df$ID = sub("\\..*", "", df$ID)
      df$ID = factor(df$ID, levels = unique(df$ID))
      df = df[!duplicated(df$ID), ]
       
      # subset only the top 20 motifs
      if (nrow(df) > 20) {
        df = df[1:20, ]
      } 
      
      df = df[, c("RANK",	"ID", "TP%",	"LOG_PVALUE")]
      df$`-LOG_PVALUE` = -df$LOG_PVALUE
      df$`-LOG_PVALUE` = as.numeric(as.character(df$`-LOG_PVALUE`))
      df$`TP%` = as.numeric(as.character(df$`TP%`))
      # print(df)
      # plot the data
      scaleFactor <- max(df$`-LOG_PVALUE`) / 100
      # Create plot
      p = ggplot()+
        geom_bar(data=df, 
                 aes(x = ID, y = `-LOG_PVALUE`), 
                 stat="identity", 
                 fill = "grey", 
                 width = 0.8) +
        geom_line(data=df, 
                  aes(x=ID, y=`TP%`* scaleFactor, group = 1), 
                  color = "black", 
                  size = 0.25) +
        geom_point(data=df, 
                   aes(x=ID, y = `TP%`* scaleFactor, group = 1), 
                   shape = 21, 
                   color="black", 
                   fill = "white", 
                   size = 2, 
                   stroke = 0.5) +
        geom_text(data = df, 
                  aes(x = ID, 
                      y = `TP%` * scaleFactor, 
                      label = paste(round(`TP%`, 2), "%")
                      ), 
                  angle = 90,
                  hjust = -0.2, 
                  size = 2.5, 
                  color = "black") +  # Add this geom_text layer
        scale_y_continuous(name="-log10(p-value)", 
                           expand = c(0,0.1, 0.1, 0)) +
        ggtitle(set) +
        theme(axis.text.x = element_text(angle = 90, 
                                         vjust = 0.5, 
                                         hjust = 1, 
                                         size = 12, 
                                         color = "black"),
              axis.text.y = element_text(angle = 90, 
                                         vjust = 0.5, 
                                         hjust = 0.5, 
                                         size = 10,
                                         color = "black"),
              panel.grid = element_blank(), 
              panel.background = element_rect(color = "black", fill = NA))
      
      ggsave(output, p, width = 3.5, height = 2.25)
    },
  error = function(e){
    message(paste('Caught an error for set:', set))
    print(e)
  }
  )
}
```

```{r}
# Get a list of all files in the input directory
motif_sets = list.files(file.path(meme_dir))

# Use lapply to apply the function to each file in motif_sets
lapply(motif_sets, plotMotifEnrichment)

```

# 4e. JG cell scRNA-seq expression of enriched-motif TFs

## Load JG cell scRNA-seq data

```{r}
# paths relative to src/
meta_dir <- "~/code/renin_atac/metadata"

# score matrix: one row per gene, one column per JG cell
df_score <- read.csv(file.path(meta_dir, "GIM_JG-cells_matrix.csv"),
                     check.names = FALSE)

# row data: gene annotation, row-parallel to df_score
df_gene <- read.csv(file.path(meta_dir, "GIM_JG-cells_row-data.csv"))
```

## Average expression per gene across all JG cells

```{r}
df_gene$average <- rowMeans(as.matrix(df_score))
```

## Define motif classes

```{r}
# HOCOMOCO motif names (motif) mapped to the mm10 gene symbols (gene)
# used in the scRNA-seq row data
motif_classes <- data.frame(
  class = c(rep("MADS box", 5), rep("RHR", 3), rep("Homeo", 6),
            rep("C4 zinc fingers", 8), rep("bZIP", 18)),
  motif = c("MEF2A", "MEF2B", "MEF2C", "MEF2D", "SRF",
            "COE1", "COE2", "COE3",
            "HXC10", "HXA9", "HXC9", "HXB7", "MEIS1", "MEIS2",
            "RXRA", "COT1", "RARG", "NR2F6", "RARA", "NR4A2", "RARB", "SATB1",
            "JUNB", "JUND", "FOSL1", "FOSL2", "FOS", "FOSB", "JUN", "CREB1",
            "ATF3", "BATF", "BATF2", "BACH2", "NFIL3", "NFE2", "NF2L1",
            "NF2L2", "MAFG", "MAFB"),
  gene = c("Mef2a", "Mef2b", "Mef2c", "Mef2d", "Srf",
           "Ebf1", "Ebf2", "Ebf3",
           "Hoxc10", "Hoxa9", "Hoxc9", "Hoxb7", "Meis1", "Meis2",
           "Rxra", "Nr2f1", "Rarg", "Nr2f6", "Rara", "Nr4a2", "Rarb", "Satb1",
           "Junb", "Jund", "Fosl1", "Fosl2", "Fos", "Fosb", "Jun", "Creb1",
           "Atf3", "Batf", "Batf2", "Bach2", "Nfil3", "Nfe2", "Nfe2l1",
           "Nfe2l2", "Mafg", "Mafb"),
  stringsAsFactors = FALSE
)
```

## Grouped bar plot of TF expression by motif class

```{r}
plot_df <- merge(motif_classes, df_gene[, c("name", "average")],
                 by.x = "gene", by.y = "name", all.x = TRUE)

class_levels <- c("MADS box", "RHR", "Homeo", "C4 zinc fingers", "bZIP")
plot_df$class <- factor(plot_df$class, levels = class_levels)

# one fixed colour per class
class_colors <- setNames(scales::hue_pal()(length(class_levels)), class_levels)

# build one bar plot per class, bars sorted high -> low
make_class_plot <- function(cls) {
  d <- plot_df[plot_df$class == cls, ]
  d <- d[order(-d$average), ]
  d$motif <- factor(d$motif, levels = d$motif)
  ylab <- if (cls == class_levels[1]) "Average expression score (JG cells)" else NULL
  ggplot(d, aes(x = motif, y = average)) +
    geom_bar(stat = "identity", width = 0.8, fill = class_colors[cls]) +
    labs(x = NULL, y = ylab, title = cls) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1,
                                     size = 8, color = "black"),
          plot.title = element_text(size = 9, hjust = 0.5),
          panel.grid = element_blank())
}

plots <- lapply(class_levels, make_class_plot)

# panel widths proportional to bar count -> uniform bar width across classes
bar_counts <- as.integer(table(plot_df$class)[class_levels])
p <- patchwork::wrap_plots(plots, nrow = 1, widths = bar_counts)

if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)
ggsave(file.path(plot_dir, "tf_expression_by_motif_class.pdf"),
       p, width = 11, height = 3.5)
p
```
