---
title: "4. De Novo Motif Enrichment Analyses"
author: "xue"
date: "2024-02-27"
output: html_document
---

# 4a. Setup environment

## Load packages

```{r}
library(genomation)
library(GenomicFeatures)
library (BSgenome.Mmusculus.UCSC.mm10)
library(ggplot2)
library(patchwork)
```

## Set working directory

```{r, setup, include=FALSE}
wd = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/differential/geniml_univ_cc_073126"
knitr::opts_knit$set(root.dir = wd)
```

## Define file paths

### Set input directory

```{r}
# universe
univ_bed = "/home/bx2ur/code/reninness_score/data/universe_073126/universe_cc_073126.bed"

# region set dir 
bed_dir = "bed"

# motif enrichment result
meme_dir =  "meme_res"
```

### Set output directory

```{r}
fa_dir = "fasta"
plot_dir = "plot/motif"

```

# 4b. Generate fasta file from regions sets

```{r}
# define reference genome
genome = BSgenome.Mmusculus.UCSC.mm10
chroms = c("chr1", "chr2","chr3","chr4","chr5",
          "chr6","chr7","chr8", "chr9", "chr10",
          "chr11","chr12","chr13","chr14","chr15",
          "chr16","chr17","chr18","chr19","chrX",
          "chrY")

# define function to loop through all region sets
getFasta = function(input_bed){
  # Modify the file name to change the extension
  name_bed = basename(input_bed)
  name_fa <- sub("\\.(bed|narrowPeak)$", ".fa", name_bed) 
  
  # Construct the output file path
  if (!dir.exists(fa_dir)) {
    dir.create(fa_dir, showWarnings = FALSE, recursive = TRUE)
  }
  output <- file.path(fa_dir, name_fa)
  
  # load peaks
  df <- read.table(input_bed, header = FALSE, sep = "\t")
  colnames(df)[1:3] <- c("chr", "start", "end")

  # Remove rows with missing coordinates
  df <- df[!is.na(df$start) & !is.na(df$end), ]
  
  # Convert to GRanges
  peaks <- makeGRangesFromDataFrame(df, keep.extra.columns = TRUE,
                                    seqnames.field = "chr",
                                    start.field = "start",
                                    end.field = "end",
                                    starts.in.df.are.0based = FALSE)
  peaks = subset(peaks, seqnames %in% chroms )

  ## extract sequence
  seq <- getSeq(genome, peaks)

  ## add names
  df = as.data.frame(peaks)
  name = paste(df$seqnames,":",df$start,"-",df$end, sep="")
  names(seq) <- name
  writeXStringSet(seq, file=output )
}
```

## Generate fasta file for the background

The background for the increased-accessibility sets is the universe with every
increased-accessibility region removed.

Why this one rather than the universe itself, or a per-subset background:

- SEA compares `--p` against `--n` with a Fisher test on a 2x2 table, which needs
  the two sets to be **disjoint**. Every region set here is a strict subset of the
  universe, so using the raw universe would put the foreground inside its own
  background and shrink the measured effect by the subset's share of the universe.
  Negligible for the small sets, fatal for large ones.
- The union of all `*_diff_up` sets is exactly `deseq_all_diff_up`, so removing
  them all leaves a background disjoint from *every* increased set at once. One
  shared background keeps the FP% denominator identical across runs, so TP%/FP%
  and E-values stay comparable between renin types. A per-subset background would
  break that comparability, and would still leave the other, heavily overlapping
  increased sets in the background diluting the shared renin signal.
- `renin_specific_regions` sits entirely inside `deseq_all_diff_up`, so it is
  covered by the same removal.

Naming: the output lands in `fa_dir` as `universe_minus_diff_up.fa`. The
`universe_*` prefix matters -- `6_motif_enrichment.sh` excludes that pattern from
its `*.fa` glob, so the background is not picked up and scored as a region set.

```{r}
bg_dir = "background"
if (!dir.exists(bg_dir)) dir.create(bg_dir, recursive = TRUE)
bg_bed = file.path(bg_dir, "universe_minus_diff_up.bed")

# exact-coordinate key; the increased sets are drawn from the universe, so whole
# regions are dropped. An interval subtraction (setdiff) would instead trim and
# split universe regions, changing their length distribution.
bedKey = function(d) paste(d$chr, d$start, d$end, sep = "_")

readBed3 = function(f) {
  d = read.table(f, header = FALSE, sep = "\t")[, 1:3]
  colnames(d) = c("chr", "start", "end")
  d
}

univ = readBed3(univ_bed)

# every increased-accessibility region set
up_beds = list.files(bed_dir, pattern = "_diff_up\\.bed$", full.names = TRUE)
up = do.call(rbind, lapply(up_beds, readBed3))

bg = univ[!(bedKey(univ) %in% bedKey(up)), ]

# must hold, or the Fisher test above is invalid
stopifnot(!any(bedKey(bg) %in% bedKey(up)))
message(sprintf("background: %d universe - %d increased = %d regions (from %d up sets)",
                nrow(univ), length(unique(bedKey(up))), nrow(bg), length(up_beds)))

write.table(bg, bg_bed, sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = FALSE)
getFasta(bg_bed)
```

## Generate fasta files for region sets of interest in renin samples

```{r}
# Get a list of all files in the input directory
region_sets = list.files(file.path(bed_dir), full.names = TRUE)

# Use lapply to apply the function to each file in motif_sets
lapply(region_sets, getFasta)

```

# 4c. Run de novo motif enrichment analyses with SEA from the MEME suite

```{bash}

bash 6_motif_enrichment.sh
```

# 4d. Plot enriched motifs

```{r}
plotMotifEnrichment = function(set){
  input <- file.path(meme_dir, set, "sea.tsv")
  output <- file.path(plot_dir, paste0(set, "_top20.pdf"))
  tryCatch(
    expr = {
      # load data
      df = read.csv(input, sep="\t", check.names = FALSE)
      df = df[1:(nrow(df) - 3), ]
      df$ID = sub("\\..*", "", df$ID)
      df$ID = factor(df$ID, levels = unique(df$ID))
      df = df[!duplicated(df$ID), ]
       
      # subset only the top 20 motifs
      if (nrow(df) > 20) {
        df = df[1:20, ]
      } 
      
      df = df[, c("RANK",	"ID", "TP%",	"LOG_PVALUE")]
      df$`-LOG_PVALUE` = -df$LOG_PVALUE
      df$`-LOG_PVALUE` = as.numeric(as.character(df$`-LOG_PVALUE`))
      df$`TP%` = as.numeric(as.character(df$`TP%`))
      # print(df)
      # plot the data
      scaleFactor <- max(df$`-LOG_PVALUE`) / 100
      # Create plot
      p = ggplot()+
        geom_bar(data=df, 
                 aes(x = ID, y = `-LOG_PVALUE`), 
                 stat="identity", 
                 fill = "grey", 
                 width = 0.8) +
        geom_line(data=df, 
                  aes(x=ID, y=`TP%`* scaleFactor, group = 1), 
                  color = "black", 
                  size = 0.25) +
        geom_point(data=df, 
                   aes(x=ID, y = `TP%`* scaleFactor, group = 1), 
                   shape = 21, 
                   color="black", 
                   fill = "white", 
                   size = 2, 
                   stroke = 0.5) +
        geom_text(data = df, 
                  aes(x = ID, 
                      y = `TP%` * scaleFactor, 
                      label = paste(round(`TP%`, 2), "%")
                      ), 
                  angle = 90,
                  hjust = -0.2, 
                  size = 2.5, 
                  color = "black") +  # Add this geom_text layer
        scale_y_continuous(name="-log10(p-value)", 
                           expand = c(0,0.1, 0.1, 0)) +
        ggtitle(set) +
        theme(axis.text.x = element_text(angle = 90, 
                                         vjust = 0.5, 
                                         hjust = 1, 
                                         size = 12, 
                                         color = "black"),
              axis.text.y = element_text(angle = 90, 
                                         vjust = 0.5, 
                                         hjust = 0.5, 
                                         size = 10,
                                         color = "black"),
              panel.grid = element_blank(), 
              panel.background = element_rect(color = "black", fill = NA))
      
      ggsave(output, p, width = 3.5, height = 2.25)
    },
  error = function(e){
    message(paste('Caught an error for set:', set))
    print(e)
  }
  )
}
```

```{r}
# Get a list of all files in the input directory
motif_sets = list.files(file.path(meme_dir))

# Use lapply to apply the function to each file in motif_sets
lapply(motif_sets, plotMotifEnrichment)

```

# 4e. JG cell scRNA-seq expression of enriched-motif TFs

## Load JG cell scRNA-seq data

```{r}
# paths relative to src/
meta_dir <- "~/code/renin_atac/metadata"

# score matrix: one row per gene, one column per JG cell
df_score <- read.csv(file.path(meta_dir, "GIM_JG-cells_matrix.csv"),
                     check.names = FALSE)

# row data: gene annotation, row-parallel to df_score
df_gene <- read.csv(file.path(meta_dir, "GIM_JG-cells_row-data.csv"))
```

## Average expression per gene across all JG cells

```{r}
df_gene$average <- rowMeans(as.matrix(df_score))
```

## Define motif classes

```{r}
# HOCOMOCO motif names (motif) mapped to the mm10 gene symbols (gene)
# used in the scRNA-seq row data.
#
# Membership is the union of the top 20 SEA hits (by E-value) from the four
# increased-accessibility sets of run 18495723 -- renin_specific_regions,
# deseq_wt_all_diff_up, deseq_tumoral_all_diff_up, deseq_recruited_all_diff_up
# -- deduplicated across sets and across H14CORE variants of the same TF
# (e.g. BATF3.H14CORE.0/.1 collapse to one BATF3). 49 motifs.
#
# Classes follow TFClass. Four are small because the top-20 union pulled in
# single non-canonical factors; collapse them into one "Other" level if the
# one-bar panels are not wanted in the plot below.
motif_classes <- data.frame(
  class = c(rep("MADS box", 5), rep("RHR", 3), rep("Homeo", 5),
            rep("C4 zinc fingers", 12), rep("bZIP", 19),
            rep("C2H2 zinc fingers", 2), "STAT", "SAND", "AT-hook"),
  motif = c("MEF2A", "MEF2B", "MEF2C", "MEF2D", "SRF",
            "COE1", "COE2", "COE3",
            "HXA9", "HXC9", "HXC10", "MEIS1", "MEIS2",
            "RXRA", "COT1", "COT2", "NR2F6", "NR2C1", "RARA", "RARB", "RARG",
            "THB", "NR4A1", "ESR1", "ESR2",
            "JUN", "JUNB", "JUND", "FOS", "FOSB", "FOSL1", "FOSL2", "ATF3",
            "ATF6A", "ATF6B", "BATF", "BATF2", "BATF3", "BACH1", "BACH2",
            "NFE2", "NF2L2", "CREB1", "CREM",
            "PRD16", "SALL3",
            "STA5A", "GMEB1", "ELYS"),
  gene = c("Mef2a", "Mef2b", "Mef2c", "Mef2d", "Srf",
           "Ebf1", "Ebf2", "Ebf3",
           "Hoxa9", "Hoxc9", "Hoxc10", "Meis1", "Meis2",
           "Rxra", "Nr2f1", "Nr2f2", "Nr2f6", "Nr2c1", "Rara", "Rarb", "Rarg",
           "Thrb", "Nr4a1", "Esr1", "Esr2",
           "Jun", "Junb", "Jund", "Fos", "Fosb", "Fosl1", "Fosl2", "Atf3",
           "Atf6", "Atf6b", "Batf", "Batf2", "Batf3", "Bach1", "Bach2",
           "Nfe2", "Nfe2l2", "Creb1", "Crem",
           "Prdm16", "Sall3",
           "Stat5a", "Gmeb1", "Ahctf1"),
  stringsAsFactors = FALSE
)

stopifnot(nrow(motif_classes) == 49,
          !anyDuplicated(motif_classes$motif))
```

## Grouped bar plot of TF expression by motif class

```{r}
plot_df <- merge(motif_classes, df_gene[, c("name", "average")],
                 by.x = "gene", by.y = "name", all.x = TRUE)

class_levels <- c("MADS box", "RHR", "Homeo", "C4 zinc fingers", "bZIP",
                  "C2H2 zinc fingers", "STAT", "SAND", "AT-hook")
plot_df$class <- factor(plot_df$class, levels = class_levels)

# one fixed colour per class
class_colors <- setNames(scales::hue_pal()(length(class_levels)), class_levels)

# build one bar plot per class, bars sorted high -> low
make_class_plot <- function(cls) {
  d <- plot_df[plot_df$class == cls, ]
  d <- d[order(-d$average), ]
  d$motif <- factor(d$motif, levels = d$motif)
  ylab <- if (cls == class_levels[1]) "Average expression score (JG cells)" else NULL
  ggplot(d, aes(x = motif, y = average)) +
    geom_bar(stat = "identity", width = 0.8, fill = class_colors[cls]) +
    labs(x = NULL, y = ylab, title = cls) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1,
                                     size = 8, color = "black"),
          plot.title = element_text(size = 9, hjust = 0.5),
          panel.grid = element_blank())
}

plots <- lapply(class_levels, make_class_plot)

# panel widths proportional to bar count -> uniform bar width across classes
bar_counts <- as.integer(table(plot_df$class)[class_levels])
p <- patchwork::wrap_plots(plots, nrow = 1, widths = bar_counts)

if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)
ggsave(file.path(plot_dir, "tf_expression_by_motif_class.pdf"),
       p, width = 11, height = 3.5)

```
