---
title: "8. DESeq2 vs. StarSpace differentially-accessible regions"
author: "xue"
date: "2026-05-22"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

This document 1) loads region-class assignments made from a trained
StarSpace model (nearest-label cosine clustering, computed in
`src/7_Evaluating_reninness_scoring_tool.ipynb`), 2) compares them to DESeq2
differentially-accessible region sets, and 3) plots tornado signal
profiles of motif-bound regions stratified by both axes.

# 8a. Setup environment

## Load packages

```{r}
# GenomicDistributions / GenomicDistributionsData / venn were attached here
# but no symbol from any of them is referenced in this document -- the
# genomic-distribution panels and the venn()-based set partition live in
# 3_plot_genomic_distributions.Rmd. Dropped 2026-09-05.
# stringr and ggplot2 are attached by tidyverse; listed explicitly because
# str_match() and ggplot() are used directly throughout.
library(genomation)
library(UpSetR)
library(tidyverse)
library(EnrichedHeatmap)
library(rtracklayer)
library(stringr)
library(ggplot2)
```

## Set working directory

```{r, setup, include=FALSE}
wd = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/differential/geniml_univ_cc_073126"
knitr::opts_knit$set(root.dir = wd)
```

## Define file paths

### DESeq2 input

```{r}
# DESeq2 differential-accessibility reports
primary_normal_file = "deseq/deseq_normal_primary_report_shrink.csv"
as_normal_file      = "deseq/deseq_normal_tumoral_report_shrink.csv"
primary_high_file   = "deseq/deseq_high_primary_report_shrink.csv"
```

### StarSpace input

```{r}
# Trained StarSpace model
# n_starspace_labels = how many "__label__" rows the model was trained with;
# StarSpace writes them as the last rows of model.tsv. The joint_l3 model has
# three (Control, Tumoral, Renin); older 2-class models have two.
n_starspace_labels = 3
# focus_label = the cluster 8d-8j treat as the signal side. Every other label
# is pooled into the "everything else" contrast that feeds the `_nr` sets, so
# the renin-vs-rest framing of the paper survives the move from 2 to 3 labels.
focus_label = "Renin"
starspace_model_path = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/reninness_score_l3_upsampled/sweep/dim100_ep50_mar0.4_neg50/model_dim100_ep50_mar0.4_neg50.tsv"
```

### Universe

```{r}
universe_file = "/home/bx2ur/code/reninness_score/data/universe_073126/universe_cc_073126.bed"
```

### Output

```{r}
bed_dir = "bed"
plot_dir = "plot"
```

# 8b. Load StarSpace region clusters

Region-to-label assignment is done in
`src/7_Evaluating_reninness_scoring_tool.ipynb` ("Region embedding colored by
nearest-label cluster"), which writes `bed/cluster_regions.csv`. That notebook
already reads `model.tsv` to draw the region embedding, so clustering there
keeps one implementation instead of two that can drift; it also puts the
cluster figure and the log2FoldChange figure on the same fitted embedding.

Each region is assigned to the label vector it is closest to by cosine
similarity, giving one cluster per trained label — `Renin` / `Non_Renin` for a
2-label model, `Control` / `Tumoral` / `Renin` for a 3-label one. Re-run that
notebook section to refresh the CSV after training a new model.

```{r}
cluster_csv <- file.path(bed_dir, "cluster_regions.csv")
if (!file.exists(cluster_csv)) {
  stop(cluster_csv, " not found. Run the 'Region embedding colored by ",
       "nearest-label cluster' section of ",
       "src/7_Evaluating_reninness_scoring_tool.ipynb, which writes it.")
}
starspace_regions <- read.csv(cluster_csv, stringsAsFactors = FALSE)

stopifnot(all(c("title", "id", "start", "end") %in% names(starspace_regions)))
message(sprintf("Loaded %s regions from %s | clusters: %s",
                format(nrow(starspace_regions), big.mark = ","), cluster_csv,
                paste(sprintf("%s=%s", names(table(starspace_regions$title)),
                              format(as.numeric(table(starspace_regions$title)),
                                     big.mark = ",")),
                      collapse = ", ")))
```


# 8c. Process DESeq2 differential regions

## Load DESeq2 reports

```{r}
primary_normal <- readGeneric(primary_normal_file, chr = 1, start = 2, end = 3,
                              keep.all.metadata = TRUE, zero.based = FALSE,
                              remove.unusual = FALSE, header = TRUE,
                              skip = 0, sep = ",")
primary_high <- readGeneric(primary_high_file, chr = 1, start = 2, end = 3,
                            keep.all.metadata = TRUE, zero.based = FALSE,
                            remove.unusual = FALSE, header = TRUE,
                            skip = 0, sep = ",")
as_normal <- readGeneric(as_normal_file, chr = 1, start = 2, end = 3,
                         keep.all.metadata = TRUE, zero.based = FALSE,
                         remove.unusual = FALSE, header = TRUE,
                         skip = 0, sep = ",")
absFoldChange = 2
fdr = 0.01
```

## Load universe

```{r}
universe <- read.table(universe_file, sep = "\t", header = FALSE,
                       stringsAsFactors = FALSE)
universe_keys <- paste(universe[, 1], universe[, 2], universe[, 3], sep = "_")
length(universe_keys)
```

## Filter into up / down sets

```{r}
primary_normal <- subset(primary_normal, pvalue < fdr)
primary_normal_up <- subset(primary_normal, log2FoldChange >=  absFoldChange)
primary_normal_down <- subset(primary_normal, log2FoldChange <= -absFoldChange)

primary_high <- subset(primary_high, pvalue < fdr)
primary_high_up <- subset(primary_high, log2FoldChange >=  absFoldChange)
primary_high_down <- subset(primary_high, log2FoldChange <= -absFoldChange)

as_normal <- subset(as_normal, pvalue < fdr)
as_normal_up <- subset(as_normal, log2FoldChange >=  absFoldChange)
as_normal_down <- subset(as_normal, log2FoldChange <= -absFoldChange)
```

## Build cross-contrast intersection sets

```{r}
lst_up <- list(
  paste(as.data.frame(as_normal_up)$seqnames,        as.data.frame(as_normal_up)$start,        as.data.frame(as_normal_up)$end,        sep = "_"),
  paste(as.data.frame(primary_normal_up)$seqnames,   as.data.frame(primary_normal_up)$start,   as.data.frame(primary_normal_up)$end,   sep = "_"),
  paste(as.data.frame(primary_high_up)$seqnames,     as.data.frame(primary_high_up)$start,     as.data.frame(primary_high_up)$end,     sep = "_")
)

lst_down <- list(
  paste(as.data.frame(as_normal_down)$seqnames,      as.data.frame(as_normal_down)$start,      as.data.frame(as_normal_down)$end,      sep = "_"),
  paste(as.data.frame(primary_normal_down)$seqnames, as.data.frame(primary_normal_down)$start, as.data.frame(primary_normal_down)$end, sep = "_"),
  paste(as.data.frame(primary_high_down)$seqnames,   as.data.frame(primary_high_down)$start,   as.data.frame(primary_high_down)$end,   sep = "_")
)

# Count DISTINCT contrasts per region, matching venn()/fromList() in
# 3_plot_genomic_distributions.Rmd: a key duplicated inside one contrast must
# not reach 2 on its own.
counts_up   <- table(unlist(lapply(lst_up,   unique)))
counts_down <- table(unlist(lapply(lst_down, unique)))

inc <- names(counts_up)  [counts_up   >= 2]
dec <- names(counts_down)[counts_down >= 2]

```

# 8d. Overlap analysis between DESeq2 and StarSpace

## Build region categories

One region-key set per trained StarSpace label + DESeq2 calls (`inc`,
`dec`, `nd`). `nd` is just `universe - inc - dec`, drawn from the BED
loaded in 8c.

`r` is the focal cluster (`focus_label`) and `nr` pools every other
label, which is what keeps the rest of the document a two-sided
renin-vs-rest comparison however many labels the model was trained with.
The ambiguous `other` band that `margin_threshold` can produce is a
non-answer, not a cluster, so it is excluded from `nr` rather than
pooled into it.

```{r}
ss_key  <- paste(starspace_regions$id, starspace_regions$start,
                 starspace_regions$end, sep = "_")
ss_sets <- split(ss_key, factor(starspace_regions$title,
                                levels = unique(starspace_regions$title)))

ambiguous_label <- "other"   # the notebook's margin band, if it was used
cluster_labels  <- setdiff(names(ss_sets), ambiguous_label)
stopifnot(focus_label %in% cluster_labels)
# Focus first, so set order in the UpSet plot and in `sets` below is stable
# across models rather than following model.tsv's vocabulary order.
cluster_labels  <- c(focus_label, setdiff(cluster_labels, focus_label))

r  <- ss_sets[[focus_label]]
nr <- unlist(ss_sets[setdiff(cluster_labels, focus_label)], use.names = FALSE)

nd <- setdiff(c(r, nr), c(inc, dec))

message(sprintf("StarSpace clusters: %s | pooled non-%s = %s",
                paste(sprintf("%s=%s", names(ss_sets),
                              format(lengths(ss_sets), big.mark = ",")),
                      collapse = ", "),
                focus_label, format(length(nr), big.mark = ",")))
```

## UpSet plot

```{r}
# UpSetR's fromList() runs the names through data.frame(), which would
# mangle "Non-renin" into "Non.renin" and break the `sets =` lookup below.
# Slugging to "Non_renin" here reproduces the historical 2-label naming.
disp        <- gsub("[^A-Za-z0-9]+", "_", cluster_labels)
focus_disp  <- gsub("[^A-Za-z0-9]+", "_", focus_label)
deseq_sets  <- list(Increase = inc, Decrease = dec, Non_differential = nd)

listInput_all <- c(deseq_sets, setNames(ss_sets[cluster_labels], disp))

svg(file.path(plot_dir, "ss_deseq_upset.svg"), width = 10, height = 6, onefile = FALSE)
upset(fromList(listInput_all),
      mainbar.y.label = "overlapped regions",
      sets.x.label = "regions per group",
      sets = c(names(deseq_sets), disp),
      keep.order = TRUE, order.by = "freq",
      show.numbers = "yes",
      set_size.show = TRUE,
      set_size.scale_max = max(lengths(listInput_all)) * 1.2,
      text.scale = c(1.5,1.5,1.5,1.5,1.75,1.5))
dev.off()

# Every DESeq2 call x every StarSpace label. With 3+ labels the pooled
# "non-<focus>" complement is appended as well, since it is the set the
# tornado / GO steps downstream actually consume; with 2 labels the single
# other label already IS that complement, so it is not duplicated.
sets <- list()
for (d in names(deseq_sets)) {
  for (i in seq_along(cluster_labels)) {
    sets[[paste0(d, ":", disp[i])]] <-
      intersect(deseq_sets[[d]], ss_sets[[cluster_labels[i]]])
  }
  if (length(cluster_labels) > 2) {
    sets[[paste0(d, ":non_", focus_disp)]] <- intersect(deseq_sets[[d]], nr)
  }
}
lengths(sets)
```

# 8e. Write intersection BED files

```{r}
write_set_bed <- function(peaks, out_path) {
  df <- data.frame(peaks = peaks)
  df <- df %>%
    extract(peaks, c("seqname", "start", "end"), regex = "(.*)_([^_]+)_([^_]+)$")
  df <- df[order(df$seqname, df$start), ]
  write.table(df, out_path, row.names = FALSE, quote = FALSE,
              sep = "\t", col.names = FALSE)
  GenomicRanges::makeGRangesFromDataFrame(df)
}

# Filename convention, unchanged for 2-label models: the focal cluster is
# "_r" and the pooled complement is "_nr", giving exactly the historical six
# (inc_r, inc_nr, dec_r, dec_nr, nd_r, nd_nr) that 8g reloads and 8j reuses.
# With 3+ labels each individual non-focal label additionally gets its own
# file (inc_control.bed, inc_tumoral.bed, ...) next to the pooled "_nr" one.
bed_prefix <- c(Increase = "inc", Decrease = "dec", Non_differential = "nd")
bed_suffix <- function(label_part) {
  if (label_part == focus_disp)                            "r"
  else if (label_part == paste0("non_", focus_disp))       "nr"
  else if (length(cluster_labels) == 2)                    "nr"
  else                                        tolower(label_part)
}

gr_sets <- list()
for (nm in names(sets)) {
  parts <- strsplit(nm, ":", fixed = TRUE)[[1]]
  var   <- paste(bed_prefix[[parts[1]]], bed_suffix(parts[2]), sep = "_")
  gr_sets[[var]] <- write_set_bed(sets[[nm]],
                                  file.path(bed_dir, paste0(var, ".bed")))
  message(sprintf("  %-16s -> %-14s %s regions", nm, paste0(var, ".bed"),
                  format(length(sets[[nm]]), big.mark = ",")))
}

# Named variables the rest of the document (8g-8j) refers to directly.
#
# Every set written above is bound to a variable, not just the pooled six.
# With a 3-label model that is the full 3 x 3 grid -- inc_r/inc_control/
# inc_tumoral, dec_*, nd_* -- plus the pooled inc_nr/dec_nr/nd_nr that 8j
# still consumes. 8g-8i treat Control and Tumoral as SEPARATE feature sets,
# so binding only the pooled names here (as this chunk used to) left the
# in-memory path out of sync with the reload path in 8g.
#
# Short aliases mirror the reload block in 8g: _c = Control, _t = Tumoral.
for (v in names(gr_sets)) assign(v, gr_sets[[v]])

alias <- c(control = "c", tumoral = "t")
for (cat in c("inc", "dec", "nd")) {
  for (long in names(alias)) {
    from <- paste(cat, long, sep = "_")
    if (!is.null(gr_sets[[from]]))
      assign(paste(cat, alias[[long]], sep = "_"), gr_sets[[from]])
  }
}

message("Bound ", length(gr_sets), " set variables: ",
        paste(sort(names(gr_sets)), collapse = ", "))
```

# 8f. StarSpace embedding vs DESeq2 differential accessibility

Two diagnostics tying the trained StarSpace embedding back to the DESeq2
calls. This section runs before the tornado plots (8g–8i) so the
embedding ↔ DE relationship is validated first, and the cos-sim values
are available to inform tornado feature-set selection downstream.

1.  **Scatter** — cosine similarity of each region embedding to the
    trained `Renin` label embedding vs DESeq2 log2 fold change, faceted
    by contrast. If the StarSpace model learned "renin-like
    accessibility," cos-sim should track log2FC; per-facet Spearman ρ
    quantifies that.
2.  **Binned line plot** — % of regions called differential by DESeq2 as
    a function of cos-sim, in quantile bins. This is the figure that
    shows how the DESeq2 call rate rises with embedding proximity.
    Pooled-across-contrasts and per-contrast views are both produced.
    It is read, not thresholded: nothing downstream cuts regions at a
    cos-sim value.

Cost: cos-sim is one matrix–vector product (O(n·d), \<1 s for n=10⁶,
d=50). The bottleneck is re-reading the three DESeq2 CSVs (8c
subset-overwrote the unfiltered objects). The merged region-level table
is built once and reused for both panels.

## Read embeddings, compute cos-sim to every label

```{r}
read_starspace_embeddings <- function(model_path, n_labels) {
  # StarSpace's model.tsv is one row per token, no header:
  #   <token>\t<v1>\t...\t<vd>
  # The trained label embeddings are written as the LAST n_labels rows, so
  # n_labels is the single knob that adapts every downstream step to a
  # 2-label model (Renin / Non_Renin) or a 3-label one
  # (Control / Tumoral / Renin).
  #
  # fread is ~10x faster than read.table on the 200 MB+ files this produces.
  if (requireNamespace("data.table", quietly = TRUE)) {
    embed <- data.table::fread(model_path, sep = "\t", header = FALSE,
                               showProgress = FALSE, data.table = FALSE)
  } else {
    embed <- read.table(model_path, sep = "\t", header = FALSE,
                        comment.char = "", quote = "",
                        stringsAsFactors = FALSE, fill = TRUE)
  }
  embed <- embed[!is.na(embed[, 1]) & embed[, 1] != "", , drop = FALSE]

  n_row <- nrow(embed)
  if (n_labels < 2 || n_labels >= n_row) {
    stop(sprintf("n_labels = %s is not usable with %s rows in model.tsv.",
                 n_labels, format(n_row, big.mark = ",")))
  }
  label_idx   <- seq.int(n_row - n_labels + 1L, n_row)
  label_rows  <- embed[ label_idx, , drop = FALSE]
  region_rows <- embed[-label_idx, , drop = FALSE]

  # Positional split, verified by the "__label__" prefix: if the model was
  # trained with a different number of classes than n_labels claims, fail
  # here rather than silently treating a region vector as a label.
  if (!all(grepl("^__label__", label_rows[, 1]))) {
    stop(sprintf(
      "The last %d row(s) of model.tsv are not all '__label__' rows (got: %s). Set n_labels to the number of trained classes.",
      n_labels, paste(label_rows[, 1], collapse = ", ")))
  }
  if (any(grepl("^__label__", region_rows[, 1]))) {
    stop(sprintf(
      "model.tsv holds more than %d '__label__' rows; raise n_labels.", n_labels))
  }

  region_emb <- as.matrix(region_rows[, -1]); storage.mode(region_emb) <- "double"
  label_emb  <- as.matrix(label_rows[,  -1]); storage.mode(label_emb)  <- "double"

  list(
    region_names = region_rows[, 1],
    region_emb   = region_emb,
    label_emb    = label_emb,
    label_names  = gsub("^__label__", "", label_rows[, 1])
  )
}

# Row-wise L2 normalization — shared by every cosine computation below.
unit_rows <- function(M) {
  u <- M / pmax(sqrt(rowSums(M * M)), 1e-12)
  u[!is.finite(u)] <- 0
  u
}

emb <- read_starspace_embeddings(starspace_model_path, n_starspace_labels)

# Cosine sim of every region to every label — one BLAS call, n_regions x n_labels.
cos_mat <- unit_rows(emb$region_emb) %*% t(unit_rows(emb$label_emb))
colnames(cos_mat) <- emb$label_names
stopifnot(focus_label %in% colnames(cos_mat))

# Everything from here to 8j is written around a focal label and its
# complement:
#   cos_renin     = cos sim to focus_label
#   cos_non_renin = the region's BEST cos sim among all the other labels, i.e.
#                   "how strongly does this region belong to something that is
#                   not focus_label".
# With 2 labels cos_non_renin is exactly the old cos sim to __label__Non_Renin.
# With 3 it pools Control and Tumoral, which keeps the downstream top-N
# and GO steps a two-sided renin-vs-rest comparison. The per-label columns
# are carried alongside so a specific label can be inspected directly.
other_labels <- setdiff(colnames(cos_mat), focus_label)
cos_renin     <- as.numeric(cos_mat[, focus_label])
cos_non_renin <- if (length(other_labels) == 1) {
  as.numeric(cos_mat[, other_labels])
} else {
  do.call(pmax, as.data.frame(cos_mat[, other_labels, drop = FALSE]))
}

emb_df <- data.frame(
  region        = emb$region_names,
  cos_renin     = cos_renin,
  cos_non_renin = cos_non_renin,
  stringsAsFactors = FALSE
)
for (lb in colnames(cos_mat)) {
  emb_df[[paste0("cos_", gsub("[^A-Za-z0-9]+", "_", lb))]] <- as.numeric(cos_mat[, lb])
}
```

## Join with the 8b cluster labels

```{r}
# 8b loaded starspace_regions from bed/cluster_regions.csv; re-read it here
# so 8f can be run on its own without re-running 8b.
if (!exists("starspace_regions")) {
  cluster_csv <- file.path(bed_dir, "cluster_regions.csv")
  if (!file.exists(cluster_csv)) {
    stop(cluster_csv, " not found — run 8b (or the clustering section of ",
         "src/7_Evaluating_reninness_scoring_tool.ipynb that writes it).")
  }
  starspace_regions <- read.csv(cluster_csv, stringsAsFactors = FALSE)
  message("Loaded starspace_regions from ", cluster_csv,
          " (", nrow(starspace_regions), " regions)")
}

ss_dt <- data.frame(
  region  = paste(starspace_regions$id, starspace_regions$start,
                  starspace_regions$end, sep = "_"),
  cluster = starspace_regions$title,
  stringsAsFactors = FALSE
)
emb_df <- merge(emb_df, ss_dt, by = "region", all.x = TRUE)

```

## Re-read DESeq2 reports (unfiltered) and build the long-format table

```{r}
# Use data.table::fread if available (~5× faster on 100k+ rows); else read.csv.
read_deseq <- function(path) {
  df <- if (requireNamespace("data.table", quietly = TRUE)) {
    as.data.frame(data.table::fread(path))
  } else {
    read.csv(path, stringsAsFactors = FALSE)
  }
  df$region <- paste(df$seqnames, df$start, df$end, sep = "_")
  df
}

dq_pn <- read_deseq(primary_normal_file); dq_pn$contrast <- "primary_normal"
dq_ph <- read_deseq(primary_high_file);   dq_ph$contrast <- "primary_high"
dq_an <- read_deseq(as_normal_file);      dq_an$contrast <- "as_normal"

# Per-contrast differential flag — same filter logic as 8c: raw pvalue for
# all three contrasts, matching filterRegions() in
# 3_plot_genomic_distributions.Rmd.
dq_pn$is_diff <- !is.na(dq_pn$pvalue) & dq_pn$pvalue < fdr &
                 abs(dq_pn$log2FoldChange) >= absFoldChange
dq_ph$is_diff <- !is.na(dq_ph$pvalue) & dq_ph$pvalue < fdr &
                 abs(dq_ph$log2FoldChange) >= absFoldChange
dq_an$is_diff <- !is.na(dq_an$pvalue) & dq_an$pvalue < fdr &
                 abs(dq_an$log2FoldChange) >= absFoldChange

keep_cols   <- c("region", "log2FoldChange", "contrast", "is_diff")
all_dq      <- rbind(dq_pn[, keep_cols], dq_ph[, keep_cols], dq_an[, keep_cols])
scatter_df  <- merge(all_dq, emb_df, by = "region")
scatter_df$contrast <- factor(scatter_df$contrast,
                              levels = c("primary_normal", "primary_high", "as_normal"))

# Per-contrast Spearman correlation, computed on ALL rows (not the
# plotting subsample). Annotated onto each facet below.
cor_df <- scatter_df %>%
  group_by(contrast) %>%
  summarise(rho = cor(log2FoldChange, cos_renin,
                      method = "spearman", use = "complete.obs"),
            n   = n(),
            .groups = "drop") %>%
  mutate(label = sprintf("rho = %.2f  (n = %s)",
                         rho, format(n, big.mark = ",")))
print(cor_df)
```

## (1) Scatter: cosine similarity to Renin label vs log2 fold change

```{r, fig.width = 12, fig.height = 5}
plot_cos_vs_log2fc <- function(df, non_diff_cap = 20000, seed = 42) {
  has_ggrastr <- requireNamespace("ggrastr", quietly = TRUE)
  if (!has_ggrastr) {
    set.seed(seed)
    diff_df <- df[df$is_diff, ]
    nd_df   <- df[!df$is_diff, ]
    nd_df   <- nd_df[
      if (nrow(nd_df) > non_diff_cap) sort(sample.int(nrow(nd_df), non_diff_cap))
      else seq_len(nrow(nd_df)),
    ]
    df <- rbind(diff_df, nd_df)
    message(sprintf(
      "ggrastr not installed; plotting %s differential + %s non-differential points",
      format(nrow(diff_df), big.mark = ","),
      format(nrow(nd_df),   big.mark = ",")
    ))
  }

  # Status: Increase / Decrease / Non-differential. Levels ordered so grey
  # draws first, colored on top.
  df$status <- factor(
    ifelse(df$is_diff & df$log2FoldChange > 0, "Increase",
    ifelse(df$is_diff & df$log2FoldChange < 0, "Decrease", "Non-differential")),
    levels = c("Non-differential", "Decrease", "Increase")
  )
  df <- df[order(df$status), ]   # draw grey first, colored on top

  # Pivot to long: each region appears twice — once with cos_sim = cos_renin
  # (top row), once with cos_sim = cos_non_renin (bottom row). Keep only the
  # columns the plot needs to bound memory at ~half scatter_df's footprint.
  df_long <- bind_rows(
    df %>% transmute(contrast, log2FoldChange, status,
                     cos_sim = cos_renin,     label_target = "Renin"),
    df %>% transmute(contrast, log2FoldChange, status,
                     cos_sim = cos_non_renin, label_target = "Non-renin")
  ) %>%
    mutate(label_target = factor(label_target, levels = c("Renin", "Non-renin")))

  pt <- if (has_ggrastr) {
    ggrastr::rasterise(geom_point(alpha = 0.3, size = 0.3), dpi = 200)
  } else {
    geom_point(alpha = 0.3, size = 0.3)
  }

  ggplot(df_long, aes(log2FoldChange, cos_sim, colour = status)) +
    pt +
    facet_grid(label_target ~ contrast, switch = "y") +
    geom_hline(yintercept = 0, colour = "black", linewidth = 0.3) +
    geom_vline(xintercept = c(-absFoldChange, absFoldChange),
               colour = "black", linewidth = 0.3, linetype = "dashed") +
    geom_text(
      data = cor_df, aes(x = -Inf, y = Inf, label = label),
      hjust = -0.05, vjust = 1.3, size = 3.2, inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c(Increase           = "#FF7F0E",   # tab:orange
                 Decrease           = "#1F77B4",   # tab:blue
                 `Non-differential` = "grey70"),
      breaks = c("Increase", "Decrease", "Non-differential"),
      name   = NULL
    ) +
    labs(x = "DESeq2 log2 fold change",
         y = "cosine similarity to label") +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          legend.position  = "bottom",
          strip.placement  = "outside")
}

p_scatter <- plot_cos_vs_log2fc(scatter_df)

ggsave(file.path(plot_dir, "starspace_cos_vs_log2fc.svg"),
       p_scatter, width = 12, height = 9, units = "in", dpi = 200)

```

## (2) P(differential) vs cosine similarity — binned line plot

Replaces the cluster stacked bar. The nearest-label split is a hard cut
that throws away the continuous cos-sim information, which is precisely
what we want to inspect here. Binning regions by cos-sim quantile and
plotting the per-bin DE rate directly shows whether embedding distance
and DESeq2 calls co-vary — a rising curve means regions the embedding
places nearer the label are more often called differential.

This is a **descriptive** figure. Region sets are defined by the DESeq2
call and ranked by cosine (8g); nothing is cut at a cos-sim value, so
the curve can be read as evidence rather than as the thing that produced
the sets.

Bin edges come from quantiles of the full pooled cos-sim distribution,
so the same edges apply to every panel — the averaged and faceted views
are directly comparable.

```{r}
plot_pde_vs_cos <- function(df,
                            facet_by_contrast = TRUE,
                            n_bins = 20,
                            label_cols = NULL) {
  # ONE ROW PER LABEL (2026-09-06). `label_cols` is a named vector: the name
  # is the row strip, the value is that label's cosine column in `df`
  # (`cos_Renin`, `cos_Control`, `cos_Tumoral` -- written by the emb_df loop
  # in 8f). NULL keeps the historical single-row, Renin-only figure, so the
  # (2a) dual-axis caller and any older call site are unchanged.
  #
  # WHY EACH ROW GETS ITS OWN BIN EDGES. Edges are quantiles of that label's
  # OWN cosine distribution. Binning the Control row on Renin quantiles would
  # put most of its regions in one or two bins -- the three labels do not
  # share a cosine scale. The comparability the single-row version had
  # (same edges across contrast panels) is preserved WITHIN each row, which
  # is the comparison the figure is actually read for.
  if (is.null(label_cols)) label_cols <- c(Renin = "cos_renin")
  missing_cols <- setdiff(unname(label_cols), names(df))
  if (length(missing_cols))
    stop("plot_pde_vs_cos: column(s) not in df: ",
         paste(missing_cols, collapse = ", "))

  bin_one <- function(lab_name, cos_col) {
    x <- df[[cos_col]]
    brks <- quantile(x, probs = seq(0, 1, length.out = n_bins + 1),
                     na.rm = TRUE)
    d <- df
    d$cos_val <- x
    d$cos_bin <- cut(x, breaks = brks, include.lowest = TRUE)
    group_cols <- if (facet_by_contrast) c("contrast", "cos_bin") else "cos_bin"
    d %>%
      group_by(across(all_of(group_cols))) %>%
      summarise(
        n       = n(),
        cos_mid = median(cos_val),
        pct_inc = 100 * mean(is_diff & log2FoldChange > 0),
        pct_dec = 100 * mean(is_diff & log2FoldChange < 0),
        .groups = "drop"
      ) %>%
      mutate(label = lab_name)
  }

  bin_summary <- purrr::imap_dfr(label_cols, ~ bin_one(.y, .x)) %>%
    mutate(label = factor(label, levels = names(label_cols))) %>%
    tidyr::pivot_longer(c(pct_inc, pct_dec),
                        names_to  = "direction",
                        values_to = "pct") %>%
    mutate(direction = recode(direction,
                              pct_inc = "Increase",
                              pct_dec = "Decrease"))

  multi <- length(label_cols) > 1
  p <- ggplot(bin_summary, aes(cos_mid, pct, colour = direction)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    scale_colour_manual(values = c(Increase = "#FF7F0E", Decrease = "#1F77B4"),
                        name = "DESeq2 direction") +
    labs(x = if (multi) "cosine similarity to the row's label (bin median)"
             else "cosine similarity to Renin label (bin median)",
         y = "% of regions called differential",
         caption = sprintf(
           "Quantile-binned within each row, %d bins, n_rows = %s%s",
           n_bins, format(nrow(df), big.mark = ","),
           if (facet_by_contrast) " (per-contrast)" else " (pooled across contrasts)"
         )) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())

  # free x: each label's cosine occupies a different range, so a shared axis
  # would compress two of the three rows into a sliver
  if (multi && facet_by_contrast) {
    p <- p + facet_grid(label ~ contrast, scales = "free_x")
  } else if (multi) {
    p <- p + facet_wrap(~ label, scales = "free_x")
  } else if (facet_by_contrast) {
    p <- p + facet_wrap(~ contrast)
  }
  p
}
```

### (2a) Pooled across contrasts — dual axis

Each direction gets its own cos-sim axis:

- **Increase (orange)** is binned by `cos_renin` and read against the
  **bottom** x-axis (cos sim to the Renin label).
- **Decrease (blue)** is binned by `cos_non_renin` and read against the
  **top** x-axis (cos sim to the Non-renin label).

Both axes share the same numerical scale (cos-sim is bounded in [-1, 1]
either way), they just measure distance to different label vectors. The
two binnings are independent — each line is in its own native axis
space.

```{r, fig.width = 7, fig.height = 4.5}
plot_pde_vs_cos_dual_axis <- function(df, n_bins = 20) {
  # ---- bin Increase by cos_renin ----
  brks_r <- quantile(df$cos_renin,
                     probs = seq(0, 1, length.out = n_bins + 1),
                     na.rm = TRUE)
  bin_inc <- df %>%
    mutate(cos_bin = cut(cos_renin, breaks = brks_r, include.lowest = TRUE)) %>%
    group_by(cos_bin) %>%
    summarise(cos_mid = median(cos_renin),
              pct     = 100 * mean(is_diff & log2FoldChange > 0),
              n       = n(),
              .groups = "drop") %>%
    mutate(direction = "Increase")

  # ---- bin Decrease by cos_non_renin ----
  brks_nr <- quantile(df$cos_non_renin,
                      probs = seq(0, 1, length.out = n_bins + 1),
                      na.rm = TRUE)
  bin_dec <- df %>%
    mutate(cos_bin = cut(cos_non_renin, breaks = brks_nr, include.lowest = TRUE)) %>%
    group_by(cos_bin) %>%
    summarise(cos_mid = median(cos_non_renin),
              pct     = 100 * mean(is_diff & log2FoldChange < 0),
              n       = n(),
              .groups = "drop") %>%
    mutate(direction = "Decrease")

  binned <- bind_rows(
    bin_inc %>% select(cos_mid, pct, n, direction),
    bin_dec %>% select(cos_mid, pct, n, direction)
  ) %>%
    mutate(direction = factor(direction, levels = c("Increase", "Decrease")))

  p <- ggplot(binned, aes(cos_mid, pct, colour = direction)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    scale_x_continuous(
      name     = "cos sim to Renin label  (Increase line, bottom axis)",
      sec.axis = dup_axis(
        name = "cos sim to Non-renin label  (Decrease line, top axis)"
      )
    ) +
    scale_colour_manual(values = c(Increase = "#FF7F0E", Decrease = "#1F77B4"),
                        name = "DESeq2 direction") +
    labs(y = "% of regions called differential",
         caption = sprintf(
           "Quantile-binned (%d bins each). n_rows = %s (pooled across contrasts)",
           n_bins, format(nrow(df), big.mark = ",")
         )) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())

}

p_pde_avg <- plot_pde_vs_cos_dual_axis(scatter_df, n_bins = 20)

ggsave(file.path(plot_dir, "starspace_pde_vs_cos_avg.svg"),
       p_pde_avg, width = 7, height = 4.5, units = "in")

```

### (2b) Faceted by contrast — primary_normal / primary_high / as_normal

```{r, fig.width = 12, fig.height = 4}
# One row per trained label, in `cluster_labels` order (focus first, so
# Renin / Control / Tumoral). Built from cluster_labels rather than hardcoded
# so a 2-label model still works -- it just draws two rows.
pde_label_cols <- setNames(
  paste0("cos_", gsub("[^A-Za-z0-9]+", "_", cluster_labels)),
  cluster_labels)

p_pde_facet <- plot_pde_vs_cos(scatter_df, facet_by_contrast = TRUE,
                               label_cols = pde_label_cols)

ggsave(file.path(plot_dir, "starspace_pde_vs_cos_by_contrast.svg"),
       p_pde_facet, width = 12,
       height = 3.4 * length(pde_label_cols), units = "in")
```

# 8g. Tornado signal profiles

## Reload intersection BEDs (for resumable execution)

```{r}
load_bed <- function(path) {
  readGeneric(path, chr = 1, start = 2, end = 3, strand = NULL,
              meta.cols = NULL, keep.all.metadata = FALSE, zero.based = FALSE,
              remove.unusual = FALSE, header = FALSE, skip = 0, sep = "\t")
}

# 8e writes one BED per (DESeq2 call x StarSpace label) plus the pooled
# "_nr" complement. 8g-8i use the PER-LABEL files, not the pool, so every
# feature set carries a single cluster identity and the two non-focal
# labels are kept APART: _r = Renin, _c = Control, _t = Tumoral.
#
# The grid is built from what is on disk rather than hand-listed, so a model
# with a different label vocabulary needs no edit here. Sets below
# `min_set_size` are dropped with a reason printed -- previously
# inc_control (30) and dec_r (3) were hardcoded out, which silently removed
# Control from the Increase row of every figure.
min_set_size <- 20

label_slug  <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
# Suffix that 8e's bed_suffix() gave each label, inverted back to the label.
bed_labels  <- if (exists("cluster_labels")) {
  cluster_labels
} else {
  c(focus_label, "Control", "Tumoral")
}
suffix_of   <- function(lab) {
  if (identical(lab, focus_label)) "r" else tolower(label_slug(lab))
}
# Short alias used in variable and panel names: "r" for the focal label,
# first letter otherwise. Guard against two labels colliding on one letter.
short_of    <- function(lab) {
  if (identical(lab, focus_label)) "r" else substr(tolower(label_slug(lab)), 1, 1)
}
stopifnot(!anyDuplicated(vapply(bed_labels, short_of, character(1))))

set_spec  <- list()
gr_list   <- list()
dropped   <- character(0)
for (cat in c("inc", "dec", "nd")) {
  for (lab in bed_labels) {
    bed <- file.path(bed_dir, sprintf("%s_%s.bed", cat, suffix_of(lab)))
    if (!file.exists(bed)) next
    gr <- load_bed(bed)
    nm <- paste(cat, short_of(lab), sep = "_")
    if (length(gr) < min_set_size) {
      dropped <- c(dropped, sprintf("%s (%d < %d)", nm, length(gr), min_set_size))
      next
    }
    assign(nm, gr)
    gr_list[[nm]]  <- gr
    set_spec[[nm]] <- list(
      label = lab,
      split = if (cat == "nd") "nd" else "diff")
  }
}
set_order <- names(set_spec)
stopifnot(length(set_order) > 0)
if (length(dropped))
  message("Feature sets below min_set_size, dropped: ",
          paste(dropped, collapse = ", "))

# Tag for cache + plot filenames. Caches and SVGs are keyed by filename
# alone, so the tag has to change whenever the SET MEMBERSHIP changes or a
# stale .rds comes back with the wrong feature groups. Folding the count in
# makes that automatic.
set_tag <- sprintf("perlabel%d", length(set_order))

# Everything downstream works off `gr_list`, not the individual variables.
gr_list <- gr_list[set_order]
message("Reloaded intersection BEDs:\n",
        paste(sprintf("  %-6s %9s regions  (%s)", set_order,
                      format(vapply(gr_list, length, integer(1)), big.mark = ","),
                      vapply(set_spec, `[[`, character(1), "label")),
              collapse = "\n"))
```

## Subset feature sets by top-N cos-sim ranking

Row count per subset is capped at a fixed top-N, and this is the **only**
selection applied to the tornado feature sets. Membership comes from the
DESeq2 call × StarSpace label intersection; ordering comes from the cosine
to that label. Each DESeq×StarSpace subset in `set_spec` stays, ranked by
the cosine similarity to **its own** StarSpace label:

- `inc_r`, `nd_r`          → keep top N by `cos_Renin`
- `dec_c`, `nd_c`          → keep top N by `cos_Control`
- `inc_t`, `dec_t`, `nd_t` → keep top N by `cos_Tumoral`

This is stricter than the old pooled rule: `cos_non_renin` was the *max*
over Control and Tumoral, so a Control set could be ranked by a region's
Tumoral affinity. Per-label ranking asks each set only about the cluster
it actually came from.

Why top-N: it gives **deterministic, balanced row counts per subset**
(every subset is at most `top_n`), so the heatmaps stop being dominated
by the huge `nd_*` blocks. The DESeq2 categorization is preserved. Note
`dec_c` (69) and `dec_t` (51) start well under `top_n`, so this step is a
no-op for them — they are bounded by their own size, not by N.

Needs `emb_df` from 8f, which writes one `cos_<label>` column per
StarSpace label alongside `cos_renin` / `cos_non_renin`.

```{r}
top_n <- 1000

# 8f slugs label names when it builds the per-label columns; match it.
cos_col <- function(label) paste0("cos_", gsub("[^A-Za-z0-9]+", "_", label))

needed_cols <- unique(vapply(set_spec, function(s) cos_col(s$label), character(1)))
if (!exists("emb_df") || !all(needed_cols %in% names(emb_df))) {
  stop("emb_df is missing column(s): ",
       paste(setdiff(needed_cols, if (exists("emb_df")) names(emb_df) else character(0)),
             collapse = ", "),
       ". Run 8f first — it writes one cos_<label> column per StarSpace label.")
}

# Key lookups: "chr_start_end" -> per-region cosine sim, one per label.
key_cos <- lapply(setNames(nm = needed_cols),
                  function(cn) setNames(emb_df[[cn]], emb_df$region))

gr_to_key <- function(gr) {
  paste(GenomicRanges::seqnames(gr),
        GenomicRanges::start(gr),
        GenomicRanges::end(gr), sep = "_")
}

# Filter a GRanges to its top-N regions by an arbitrary cos-sim lookup.
filter_top_by_cos <- function(gr, cos_lookup, top_n) {
  keys     <- gr_to_key(gr)
  cos_vals <- cos_lookup[keys]
  # Drop NAs (regions missing from emb_df). Then take top-N by cos.
  finite_ix <- which(is.finite(cos_vals))
  ord <- finite_ix[order(cos_vals[finite_ix], decreasing = TRUE)]
  gr[head(ord, top_n)]
}

sizes_before <- vapply(gr_list, length, integer(1))

# Every set ranked by the cosine to its own StarSpace label, and capped at
# top_n. This is the ONLY selection applied to the tornado feature sets:
# membership comes from the DESeq2 call x StarSpace label intersection, and
# ordering comes from the cosine. Nothing is matched, weighted, or
# thresholded on a third variable.
#
# baseMean matching was tried here and REMOVED (2026-09-05). It is not used
# in the differential section either, and measurement showed it does not do
# what it was meant to: baseMean correlates with bigWig coverage at only
# Spearman 0.41-0.68, and matching inc_r's baseMean exactly (2.70 -> 3.67)
# moved the PN bigWig gap just 4.44x -> 3.58x while pushing NR, NRT and TN
# FURTHER from parity. See notes/BASEMEAN_MATCHING_2026-09-04.md.
gr_list <- lapply(set_order, function(nm)
  filter_top_by_cos(gr_list[[nm]],
                    key_cos[[cos_col(set_spec[[nm]]$label)]], top_n))
names(gr_list) <- set_order

sizes_after <- vapply(gr_list, length, integer(1))

# top_n is a COUNT, so the cosine value it cuts at differs per set: for a
# 200,000-region set it is the top 0.5%, for dec_c (69) it is a no-op. Print
# the surviving cosine range so that truncation is visible, not implicit.
message(sprintf("Ranked by cosine, capped at top-%s:", format(top_n, big.mark = ",")))
for (nm in set_order) {
  cvs <- key_cos[[cos_col(set_spec[[nm]]$label)]][gr_to_key(gr_list[[nm]])]
  cvs <- cvs[is.finite(cvs)]
  message(sprintf(
    "  %-7s %9s -> %9s  (%5.1f%% kept; cos %.3f to %.3f, ranked by %s)",
    paste0(nm, ":"), format(sizes_before[[nm]], big.mark = ","),
    format(sizes_after[[nm]], big.mark = ","),
    100 * sizes_after[[nm]] / max(sizes_before[[nm]], 1),
    if (length(cvs)) min(cvs) else NA_real_,
    if (length(cvs)) max(cvs) else NA_real_,
    cos_col(set_spec[[nm]]$label)))
}
```

## Load bigWig signal files

Two bigwig configurations — one for each tornado plot we'll produce:

1.  **`bigwigs_avg`**: five group averages (Non-renin, WT, Recruited,
    Tumoral, Non-renin tumoral).
2.  **`bigwigs_samples`**: hand-picked individual replicate bigwigs.

`EnrichedHeatmap`'s `normalizeToMatrix` takes a GRanges of signal
coverage, which we import on the fly via `rtracklayer::import.bw()`.

PEPATAC output layout (renin + ENCODE pipelines):
`<base>/<sample_id>/aligned_mm10/<sample_id>_smooth.bw`

```{r}
# PEPATAC output roots. Renin and ENCODE samples live in sibling pipeline
# dirs under the same brickyard processed root.
renin_bw_base  <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/renin_ATAC/results_pipeline"
encode_bw_base <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/encode_ATAC/results_pipeline"
susztak_bw_base <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/susztak_ATAC/results_pipeline"
ctrl_tumoral_bw_base <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/ctrl_tumoral_ATAC/results_pipeline"

pepatac_bw <- function(sample_id, base = renin_bw_base) {
  file.path(base, sample_id, "aligned_mm10",
            paste0(sample_id, "_smooth.bw"))
}

# (1) Group-averaged bigwigs — one per sample group, for the high-level
# tornado. The PEPATAC `data/` sibling dir holds NR/WT/PH/PN/TN_average.bw.
# Order here = display order in legends and as columns in the heatmap.
avg_bw_dir <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/data"

# Ordered renin -> non-renin, with the two tumoral tracks adjacent.
bigwigs_avg <- c(
  WT  = file.path(avg_bw_dir, "PN_average.bw"),   # wild-type renin cells (subset of Renin_Normal)
  PH  = file.path(avg_bw_dir, "PH_average.bw"),   # primary high (recruited)
  TN  = file.path(avg_bw_dir, "TN_average.bw"),   # tumoral (As4.1)
  NRT = file.path(avg_bw_dir, "NRT_average.bw"),  # non-renin tumoral (ctrl_tumoral_ATAC, 51 cell-line samples)
  NR  = file.path(avg_bw_dir, "NR_average.bw")    # non-renin (ENCODE controls)
)
stopifnot(all(file.exists(bigwigs_avg)))

# (2) Per-replicate bigwigs — for the per-sample tornado.
#
# Selected from metadata, not hand-listed, so the panel tracks the pipeline
# directories and the annotation sheet rather than drifting from them:
#
#   Renin_WT / Renin_Recruited / Renin_Tumoral : ALL samples
#   Control_Primary / Control_Tumoral          : KIDNEY tissue only
#
# Renin is a kidney gene, so non-kidney controls are the wrong comparison —
# and taking all of them would mean 178 ENCODE + 51 tumour-line columns.
#
# "Kidney" is decided two ways because the two control groups annotate
# tissue differently:
#   * Control_Primary carries the organ directly ("primary_kidney_cells").
#   * Control_Tumoral carries a CELL LINE in `Tissue` ("RENCA", "RAG"); which
#     lines are renal comes from the `tissue` column of
#     renin_annotation_sheet_add_tumoral_l4.csv ("RENCA_renal_carcinoma",
#     "RAG_renal_adenocarcinoma"). That sheet keys by abbreviated names
#     (RENCA_EVP_1) that do not match the pipeline directories, so the join
#     goes directory -> `Tissue` in renin_encode_atac.csv -> renal? in l4.
meta_bw <- read.csv(path.expand(
  "~/code/renin_atac/metadata/renin_encode_atac.csv"), stringsAsFactors = FALSE)
l4_tissue <- read.csv(path.expand(
  "~/code/renin_atac/metadata/renin_annotation_sheet_add_tumoral_l4.csv"),
  stringsAsFactors = FALSE)$tissue

kidney_re    <- "kidney|renal"
kidney_lines <- unique(sub("_.*$", "",
  l4_tissue[grepl(kidney_re, l4_tissue, ignore.case = TRUE)]))
is_kidney <- function(tis) {
  !is.na(tis) & (grepl(kidney_re, tis, ignore.case = TRUE) | tis %in% kidney_lines)
}

# sample id -> pipeline project, by scanning the PEPATAC roots.
#
# `exclude_projects` drops a whole PEPATAC project from the per-sample panel.
# susztak_ATAC is excluded: it is the only pre-pipestat run in the set
# (2021-2024, and the one whose assets.tsv pointed at a retired refgenie),
# so its tracks are the least comparable with the rest. Removing it leaves
# Control_Primary as 8 ENCODE kidney samples from a single processing batch.
exclude_projects <- "susztak_ATAC"

bw_roots <- c(renin_ATAC        = renin_bw_base,
              encode_ATAC       = encode_bw_base,
              susztak_ATAC      = susztak_bw_base,
              ctrl_tumoral_ATAC = ctrl_tumoral_bw_base)
bw_roots <- bw_roots[setdiff(names(bw_roots), exclude_projects)]
if (length(exclude_projects)) {
  message("Excluded PEPATAC project(s) from the per-sample panel: ",
          paste(exclude_projects, collapse = ", "))
}
sample_root <- unlist(lapply(bw_roots, function(r)
  stats::setNames(rep(r, length(list.files(r))), list.files(r))))
names(sample_root) <- sub("^[^.]*\\.", "", names(sample_root))   # drop the root prefix

take_all    <- c("Renin_WT", "Renin_Recruited", "Renin_Tumoral")
take_kidney <- c("Control_Primary", "Control_Tumoral")

sel <- meta_bw[meta_bw$Group_2 %in% c(take_all, take_kidney) &
               meta_bw$SampleID %in% names(sample_root), ]
sel <- sel[sel$Group_2 %in% take_all | is_kidney(sel$Tissue), ]

# Samples that have a pipeline directory but no Group_2 are reported, never
# silently dropped — As4.1_1 is one, and it is a real renin tumoral library.
renin_unassigned <- setdiff(list.files(renin_bw_base),
                            meta_bw$SampleID[nzchar(meta_bw$Group_2)])
if (length(renin_unassigned)) {
  message("NOTE: renin_ATAC sample(s) with no Group_2, EXCLUDED: ",
          paste(renin_unassigned, collapse = ", "))
}

# Long susztak ids are unreadable as a heatmap `name`; panels are titled by
# group anyway, so abbreviate purely for legends and messages. Inert while
# susztak_ATAC is in `exclude_projects`; kept so re-including it stays a
# one-word change.
short_name <- function(x) {
  x <- sub("^([0-9]+)-week_mouse_kidney_bulk_ATAC-seq_sample_", "\\1W_", x)
  sub("^P0_mouse_kidney_bulk_ATAC-seq_sample_", "P0_", x)
}
bigwigs_samples <- stats::setNames(
  pepatac_bw(sel$SampleID, sample_root[sel$SampleID]),
  short_name(sel$SampleID))

stopifnot(!anyDuplicated(names(bigwigs_samples)))
stopifnot(all(file.exists(bigwigs_samples)))

for (g in c(take_all, take_kidney)) {
  ids <- sel$SampleID[sel$Group_2 == g]
  message(sprintf("  %-16s n=%2d  (%s)", g, length(ids),
                  paste(sort(unique(sel$Tissue[sel$Group_2 == g])), collapse = ", ")))
}
excluded <- meta_bw[meta_bw$Group_2 %in% take_kidney &
                    meta_bw$SampleID %in% names(sample_root) &
                    !is_kidney(meta_bw$Tissue), ]
message(sprintf("  excluded %d non-kidney control sample(s) across %d tissue(s)",
                nrow(excluded), length(unique(excluded$Tissue))))

# 4T1_Adherent_2, 4T1_DAC_decitabine_3 and 4T1_MFP_2 never had chrM removed
# (2026-08-17 QC audit). All three are 4T1, so none survive the kidney
# filter — checked rather than assumed, since the filter could change.
chrm_dirty <- c("4T1_Adherent_2", "4T1_DAC_decitabine_3", "4T1_MFP_2")
if (any(chrm_dirty %in% sel$SampleID)) {
  message("NOTE: chrM not removed from ",
          paste(intersect(chrm_dirty, sel$SampleID), collapse = ", "),
          " (<=1.02x low; fine for display, not for FRiP/complexity).")
}
message(sprintf("Per-sample bigWigs: %d total", length(bigwigs_samples)))
```

## Sample -> group map for the per-sample tornado

The per-sample heatmap has 49 columns. Labelling each with its sample id is
unreadable and is not the comparison being made, so panels are ordered by
sample group and each group is named once, over the middle panel of its run.

Groups come from the `Group_2` column of `metadata/renin_encode_atac.csv`,
which already carries exactly these five levels — they are not re-derived
here. All 49 samples resolve; the chunk stops if any does not.

Two things this figure cannot hide, both from the 2026-08-17 QC audit
(`notes/data_quality_ENCODE_comparison_2026-08-17.md`):

* **`Control_Tumoral` is the 10 KIDNEY lines only** — RENCA (8) and RAG (2)
  — not all 51 in `ctrl_tumoral_ATAC`. Renin is a kidney gene, so the breast,
  lung, melanoma, colon and pancreatic lines are the wrong comparison, and 51
  columns would swamp the figure. Note this makes the per-sample
  `Control_Tumoral` a different set from `NRT_average.bw`, which pools all
  51 — the two figures are not showing the same samples for that group.
* **`WT_1` / `WT_2` are the same libraries as `Ren1c-YFP_1` / `Ren1c-YFP_2`**
  under two names. `Renin_WT` therefore shows four panels for two unique
  libraries. Set `drop_duplicate_wt <- TRUE` to plot the two unique ones.

```{r}
drop_duplicate_wt <- FALSE   # TRUE drops Ren1c-YFP_1/_2, the duplicate copies

# meta_bw was read when bigwigs_samples was built, just above.
meta_df <- meta_bw
stopifnot(all(c("SampleID", "Group_2") %in% names(meta_df)))

# bigwigs_samples is keyed by display name; recover the PEPATAC sample id
# from the path, which is the directory the bigwig sits in.
sample_id_of <- function(paths) basename(dirname(dirname(paths)))
sample_groups <- stats::setNames(
  meta_df$Group_2[match(sample_id_of(bigwigs_samples), meta_df$SampleID)],
  names(bigwigs_samples))

if (anyNA(sample_groups)) {
  stop("No Group_2 for: ",
       paste(names(sample_groups)[is.na(sample_groups)], collapse = ", "))
}

if (drop_duplicate_wt) {
  dup <- c("Ren1c-YFP_1", "Ren1c-YFP_2")
  bigwigs_samples <- bigwigs_samples[setdiff(names(bigwigs_samples), dup)]
  sample_groups   <- sample_groups[names(bigwigs_samples)]
  message("Dropped duplicate WT copies: ", paste(dup, collapse = ", "))
}

# Order panels by group so each group is one contiguous run — the column-title
# logic in build_tornado_heatmap_independent() depends on it.
group_levels    <- c("Renin_WT", "Renin_Recruited", "Renin_Tumoral",
                     "Control_Primary", "Control_Tumoral")
present         <- intersect(group_levels, unique(sample_groups))
ord             <- order(match(sample_groups, present))
bigwigs_samples <- bigwigs_samples[ord]
sample_groups   <- factor(sample_groups[ord], levels = present)
stopifnot(!is.unsorted(as.integer(sample_groups)))   # runs are contiguous

message("Per-sample panels, grouped:")
for (g in present) {
  nm <- names(sample_groups)[sample_groups == g]
  message(sprintf("  %-16s n=%2d  %s", g, length(nm), paste(nm, collapse = ", ")))
}
missing_groups <- setdiff(group_levels, present)
if (length(missing_groups)) {
  message("  NOT represented per-sample: ", paste(missing_groups, collapse = ", "),
          " (group-average figure only)")
}
```

## Per-track background scale factors

FRiP differs **17-fold** across these sample groups — 0.030 in sorted renin
primary vs 0.501 in the tumour lines
(`notes/data_quality_ENCODE_comparison_2026-08-17.md`). That is
signal-to-noise, not depth, so PEPATAC's per-sample normalisation does not
touch it and no rescaling can undo it. Measured over 20,000 matched
background bins the tracks agree to within 1.45x, but at the 99th percentile
they differ 4.7x — the whole disparity lives in the peaks.

Consequence for the figures: on one shared colour scale (cap = max q99 =
7.86e-08, set by Ctrl_tumoral) the two renin primary tracks reach only ~21%
of the ramp. They look washed out for reasons that have nothing to do with
chromatin.

`heatmap_norm = "background"` divides each track by its own genomic
background, so the colour becomes **fold enrichment over that track's noise
floor** and means the same thing in every column. Background is estimated
from random 10-kb genomic bins, NOT from peaks and NOT from the consensus
universe — peak-derived factors are biased exactly when composition differs,
which is the situation here. This is the standard signal-over-background
construction (MACS2 `fold_enrichment`, `bigwigCompare --ratio`).

That the result is not itself a FRiP artifact is checkable, and it checks
out: FRiP ranks WT lowest and Ctrl_tumoral highest, while `inc_r` enrichment
ranks them the other way round (7.31 vs 1.18), and `dec_c` flips completely
(0.89 in WT vs 6.54 in Ctrl_tumoral). A quality artifact would push the
high-FRiP tracks up everywhere.

Factors are cached — the bins are random but seeded, so the same factors
come back across runs.

```{r}
bg_bin_n     <- 20000
bg_bin_width <- 10000
bg_seed      <- 3

bw_background_factors <- function(bigwigs, n = bg_bin_n, w = bg_bin_width,
                                  seed = bg_seed, cache = NULL) {
  if (!is.null(cache) && file.exists(cache)) {
    f <- readRDS(cache)
    if (setequal(names(f), names(bigwigs))) {
      message("Background factors from cache: ", cache)
      return(f[names(bigwigs)])
    }
  }
  si <- GenomeInfoDb::seqinfo(rtracklayer::BigWigFile(bigwigs[[1]]))
  keep <- grep("^chr[0-9XY]+$", GenomeInfoDb::seqnames(si), value = TRUE)
  si <- si[keep]
  sl <- GenomeInfoDb::seqlengths(si)

  set.seed(seed)
  ch <- sample(names(sl), n, replace = TRUE, prob = sl / sum(sl))
  st <- as.integer(stats::runif(n, 1, sl[ch] - w))
  bg <- GenomicRanges::GRanges(ch, IRanges::IRanges(st, st + w - 1L))

  f <- vapply(bigwigs, function(path) {
    sig <- rtracklayer::import(path, which = bg, as = "RleList")
    v   <- rep(NA_real_, length(bg)); cc <- as.character(GenomicRanges::seqnames(bg))
    for (c1 in intersect(unique(cc), names(sig))) {
      i <- which(cc == c1)
      v[i] <- as.numeric(mean(IRanges::Views(sig[[c1]],
                GenomicRanges::start(bg)[i], GenomicRanges::end(bg)[i])))
    }
    mean(v, na.rm = TRUE)
  }, numeric(1))

  message(sprintf("Background factors (%s x %s-kb bins):", format(n, big.mark = ","), w/1000))
  for (nm in names(f)) message(sprintf("  %-14s %.4g", nm, f[[nm]]))
  message(sprintf("  spread across tracks: %.2fx", max(f)/min(f)))
  if (!is.null(cache)) saveRDS(f, cache)
  f
}

bg_avg     <- bw_background_factors(bigwigs_avg,
                 cache = file.path(plot_dir, "bg_factors_avg.rds"))
bg_samples <- bw_background_factors(bigwigs_samples,
                 cache = file.path(plot_dir, "bg_factors_samples.rds"))
```

## Define feature set + signal-binning helper

For each (feature, bigwig) pair, resize features to their midpoint, bin
the bigwig coverage into `w`-bp windows spanning ±`extend` around each
midpoint, then average across features per position. Output is a long-
format df with (`feature_set, sample_name, position, mean, sd, n`).

Defaults: **extend = 1000 bp, w = 20 bp** → 100 bins per row (50 per side).

```{r}
flatten_features_eh <- function(features, bigwigs, extend = 1000, w = 20,
                                verbose = TRUE) {
  ts  <- function() format(Sys.time(), "%H:%M:%S")
  log <- function(...) if (verbose) message(sprintf("[%s] %s", ts(), paste0(...)))

  pos_vec <- c(
    seq(-extend + w / 2, -w / 2, by = w),
    seq( w / 2,  extend - w / 2, by = w)
  )
  out <- vector("list", length(features) * length(bigwigs))
  k <- 0L

  n_s <- length(bigwigs)
  n_f <- length(features)
  log(sprintf(">> flatten_features_eh: %d sample(s) × %d feature set(s) = %d calls; ",
              n_s, n_f, n_s * n_f),
      sprintf("extend=%d w=%d", extend, w))

  total_t0 <- Sys.time()
  for (si in seq_along(bigwigs)) {
    sample_name <- names(bigwigs)[si]
    bw_path     <- bigwigs[[sample_name]]

    log(sprintf("[%d/%d] importing %s ...", si, n_s, basename(bw_path)))
    t0 <- Sys.time()
    signal <- rtracklayer::import(bw_path)
    log(sprintf("    imported in %.1fs; %s ranges, %.0f MB in memory",
                as.numeric(Sys.time() - t0, units = "secs"),
                format(length(signal), big.mark = ","),
                object.size(signal) / 1e6))

    for (fi in seq_along(features)) {
      feature_set <- names(features)[fi]
      feat_gr     <- features[[feature_set]]
      if (length(feat_gr) == 0) {
        log(sprintf("    [%d/%d] %s × %s: skipped (0 features)",
                    fi, n_f, sample_name, feature_set))
        next
      }
      log(sprintf("    [%d/%d] %s × %s: %s features, normalizeToMatrix ...",
                  fi, n_f, sample_name, feature_set,
                  format(length(feat_gr), big.mark = ",")))
      t1 <- Sys.time()
      mid <- GenomicRanges::resize(feat_gr, width = 1, fix = "center")
      mat <- EnrichedHeatmap::normalizeToMatrix(
        signal, target = mid,
        value_column = "score",
        extend = extend, mean_mode = "w0", w = w
      )
      log(sprintf("        done in %.1fs; mat dim %d × %d",
                  as.numeric(Sys.time() - t1, units = "secs"),
                  nrow(mat), ncol(mat)))

      k <- k + 1L
      out[[k]] <- data.frame(
        feature_set = feature_set,
        sample_name = sample_name,
        position    = pos_vec,
        mean        = colMeans(mat, na.rm = TRUE),
        sd          = matrixStats::colSds(as.matrix(mat), na.rm = TRUE),
        n           = nrow(mat),
        stringsAsFactors = FALSE
      )
    }
    rm(signal); gc(verbose = FALSE)
  }
  log(sprintf(">> total wall time: %.1f min",
              as.numeric(Sys.time() - total_t0, units = "mins")))
  do.call(rbind, out[seq_len(k)])
}

# DESeq2 × StarSpace intersection feature sets for tornado + heatmap — one
# per (DESeq2 call × StarSpace label), as declared in `set_spec` in 8g.
# The two non-focal labels are kept SEPARATE throughout (_c = Control,
# _t = Tumoral); the pooled "_nr" union is 8j's, not this section's. With a
# 3-label model the grid is inc/dec/nd × Renin/Control/Tumoral, minus any
# set under `min_set_size`.
# Each was ranked top-N by its own label's cosine. The pooled "_nr" sets are
# no longer used here.
feats <- do.call(GRangesList, gr_list)
```

# 8h. Tornado line plots

±1000 bp around each region's midpoint, w = 20 bp bins (100 bins total).
Lines are per-bin mean across regions; **ribbons show mean ± SEM**
(standard error of the per-region values at that bin). Color palette is
shared with the heatmap top-annotation lines in 8i.

## Helpers

```{r}
# Shared discrete color palette for the feature groups. Same colors in
# tornado_facets() lines and build_tornado_heatmap_independent()
# top-annotation curves so the two figures cross-reference visually.
#
# The seven hues below are the historical assignment and are kept so old
# figures stay comparable — inc_r red, nd_r blue/cyan, as in the original
# 4-set palette. `set_spec` is now built from whatever labels the model has
# (§8g), so any set without an entry here — inc_c, dec_r, or a set from a
# different label vocabulary — gets a hue appended from hue_pal(). Never
# reorder the fixed block: the colors are referenced in written-up figures.
feature_group_colors <- c(
  inc_r = "#F8766D",   # red    — Increase × Renin
  inc_t = "#C49A00",   # gold   — Increase × Tumoral
  dec_c = "#53B400",   # green  — Decrease × Control
  dec_t = "#00C094",   # teal   — Decrease × Tumoral
  nd_r  = "#00B6EB",   # blue   — Non-diff × Renin
  nd_c  = "#A58AFF",   # purple — Non-diff × Control
  nd_t  = "#FB61D7"    # pink   — Non-diff × Tumoral
)

missing_cols <- setdiff(set_order, names(feature_group_colors))
if (length(missing_cols)) {
  extra <- scales::hue_pal(h = c(15, 375), l = 45)(length(missing_cols))
  names(extra) <- missing_cols
  feature_group_colors <- c(feature_group_colors, extra)
  message("Palette extended for ", length(missing_cols), " new set(s): ",
          paste(missing_cols, collapse = ", "))
}

# Line + ribbon plot, faceted by sample. `ribbon` controls the band:
#   "sem"  = mean ± SEM (default — narrow at high n; shows mean reliability)
#   "sd"   = mean ± SD (wider — shows region-to-region spread)
#   "ci95" = mean ± 1.96 × SEM
#   "none" = line only
tornado_facets <- function(df, title,
                           y_label      = "mean signal",
                           ribbon       = c("sem", "sd", "ci95", "none"),
                           ribbon_alpha = 0.2,
                           hline_zero   = FALSE,
                           group_colors = feature_group_colors) {
  ribbon <- match.arg(ribbon)

  if (ribbon != "none") {
    df <- df %>%
      dplyr::mutate(
        .half = dplyr::case_when(
          ribbon == "sem"  ~ sd / sqrt(pmax(n, 1)),
          ribbon == "ci95" ~ 1.96 * sd / sqrt(pmax(n, 1)),
          ribbon == "sd"   ~ sd,
          TRUE             ~ 0
        ),
        .ymin = mean - .half,
        .ymax = mean + .half
      )
  }

  p <- ggplot(df, aes(position))
  if (ribbon != "none") {
    p <- p +
      geom_ribbon(aes(ymin = .ymin, ymax = .ymax, fill = feature_set),
                  alpha = ribbon_alpha, colour = NA) +
      scale_fill_manual(values = group_colors, guide = "none")
  }
  p <- p +
    geom_line(aes(y = mean, colour = feature_set), linewidth = 0.5) +
    scale_colour_manual(values = group_colors) +
    facet_wrap(~ sample_name) +
    labs(x = "position relative to region midpoint",
         y = y_label, colour = NULL, title = title) +
    theme_bw(base_size = 11) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position = "right")
  if (hline_zero) {
    p <- p + geom_hline(yintercept = 0, linetype = "dashed",
                        linewidth = 0.3, colour = "grey50")
  }
  p
}
```

## (1) Group-averaged tornado

```{r}
# set_tag keeps the per-label caches separate from the old 4-set ones,
# which were keyed by filename alone.
df_avg_rds <- file.path(plot_dir, sprintf("df_avg_1k_w20_%s.rds", set_tag))
df_avg <- if (file.exists(df_avg_rds)) {
  message("Loading cached df_avg from ", df_avg_rds)
  readRDS(df_avg_rds)
} else {
  message(sprintf("Computing df_avg (%d bigwig imports x %d feature sets)...",
                  length(bigwigs_avg), length(feats)))
  out <- flatten_features_eh(feats, bigwigs_avg, extend = 1000, w = 20)
  saveRDS(out, df_avg_rds)
  out
}

p_avg <- tornado_facets(
  df_avg,
  "ATAC signal at region midpoints (group averages, mean ± SEM)",
  ribbon = "sem"
)
ggsave(file.path(plot_dir,
                 sprintf("tornado_group_averages_1k_w20_%s.svg", set_tag)),
       p_avg, width = 12, height = 8, units = "in", dpi = 300)
```

## (2) Per-sample tornado

To keep memory bounded for the 33-sample run, subsample each feature set
to at most 5,000 regions before normalizeToMatrix. The mean signal curve
is essentially indistinguishable from the full set, since each bin still
averages over thousands of regions.

```{r}
subsample_feats <- function(gr_list, max_n = 5000, seed = 42) {
  set.seed(seed)
  GRangesList(lapply(gr_list, function(g) {
    if (length(g) <= max_n) g
    else g[sort(sample.int(length(g), max_n))]
  }))
}
subsample_n <- 5000   # set to NULL to disable

feats_for_samples <- if (is.null(subsample_n)) {
  feats
} else {
  subsample_feats(feats, max_n = subsample_n)
}
vapply(feats_for_samples, length, integer(1))

df_samples_rds <- file.path(plot_dir,
                            sprintf("df_samples_1k_w20_%s.rds", set_tag))
df_samples <- if (file.exists(df_samples_rds)) {
  message("Loading cached df_samples from ", df_samples_rds)
  readRDS(df_samples_rds)
} else {
  message("Computing df_samples (33 bigwig imports + normalizeToMatrix)...")
  out <- flatten_features_eh(feats_for_samples, bigwigs_samples,
                              extend = 1000, w = 20)
  saveRDS(out, df_samples_rds)
  out
}

p_samples <- tornado_facets(
  df_samples,
  "ATAC signal at region midpoints (selected samples, mean ± SEM)",
  ribbon       = "sem",
  ribbon_alpha = 0.15
) +
  theme(legend.text     = element_text(size = 7),
        legend.key.size = unit(0.6, "lines"))
ggsave(file.path(plot_dir,
                 sprintf("tornado_per_sample_1k_w20_%s.svg", set_tag)),
       p_samples, width = 16, height = 12, units = "in", dpi = 300)
```

# 8i. Per-region tornado heatmaps

±1000 bp / w = 20 bp matrices, same feature sets as 8h. Each row is a
region; columns are bin positions. Color = signal (raw) or fraction of
total per-(sample, bin) (sum-to-one normalization).

## Helpers

```{r}
compute_tornado_matrices <- function(features, bigwigs, extend = 1000, w = 20,
                                     verbose = TRUE) {
  ts  <- function() format(Sys.time(), "%H:%M:%S")
  log <- function(...) if (verbose) message(sprintf("[%s] %s", ts(), paste0(...)))

  group_names <- names(features)
  combined <- do.call(c, lapply(group_names, function(fn) {
    g <- features[[fn]]
    if (length(g) == 0) return(g)
    mcols(g) <- S4Vectors::DataFrame(.group = rep(fn, length(g)))
    g
  }))
  group_factor <- factor(combined$.group, levels = group_names)
  target_mid   <- GenomicRanges::resize(combined, width = 1, fix = "center")
  log(sprintf(">> compute_tornado_matrices: %d sample(s) × %d feature group(s) → ",
              length(bigwigs), length(features)),
      sprintf("%s total regions, %d bins", format(length(target_mid), big.mark = ","),
              2 * extend / w))

  mats <- vector("list", length(bigwigs))
  names(mats) <- names(bigwigs)
  for (i in seq_along(bigwigs)) {
    sn <- names(bigwigs)[i]
    log(sprintf("  [%d/%d] %s: import + normalizeToMatrix ...",
                i, length(bigwigs), sn))
    t0 <- Sys.time()
    sig <- rtracklayer::import(bigwigs[[i]])
    mats[[sn]] <- EnrichedHeatmap::normalizeToMatrix(
      sig, target_mid,
      value_column = "score",
      extend = extend, mean_mode = "w0", w = w
    )
    log(sprintf("      done in %.1fs",
                as.numeric(Sys.time() - t0, units = "secs")))
    rm(sig); gc(verbose = FALSE)
  }

  list(
    mats         = mats,
    group_factor = group_factor,
    extend       = extend,
    w            = w
  )
}

# Apply (or skip) sum_to_one across feature_sets, then compute one shared
# color function spanning every (sample, group, bin) cell. Lets multiple
# build_tornado_heatmap_independent() calls (e.g. diff vs nd splits)
# share an exact color scale.
# Divide every sample matrix by that track's own genomic background, turning
# the plotted value into fold enrichment. Errors rather than silently passing
# raw signal through if a factor is missing — a half-normalised panel set is
# worse than an un-normalised one.
apply_bg_norm <- function(mats, bg_factors) {
  if (is.null(bg_factors)) {
    stop("normalize = \"background\" needs bg_factors; ",
         "run the 'Per-track background scale factors' chunk in 8g.")
  }
  missing <- setdiff(names(mats), names(bg_factors))
  if (length(missing)) {
    stop("No background factor for: ", paste(missing, collapse = ", "))
  }
  for (sn in names(mats)) mats[[sn]] <- mats[[sn]] / bg_factors[[sn]]
  mats
}

compute_unified_col_fun <- function(tornado_data,
                                    normalize = c("none", "sum_to_one", "background"),
                                    col_pal = c("white", "darkblue"),
                                    cap_quantile = 0.99,
                                    bg_factors = NULL) {
  normalize    <- match.arg(normalize)
  mats         <- tornado_data$mats
  group_factor <- tornado_data$group_factor

  if (normalize == "background") mats <- apply_bg_norm(mats, bg_factors)

  if (normalize == "sum_to_one") {
    for (sn in names(mats)) {
      mat <- mats[[sn]]
      group_bin_means <- vapply(levels(group_factor), function(g) {
        colMeans(mat[group_factor == g, , drop = FALSE], na.rm = TRUE)
      }, numeric(ncol(mat)))
      bin_totals <- rowSums(group_bin_means, na.rm = TRUE)
      mats[[sn]] <- sweep(mat, 2, pmax(bin_totals, 1e-12), "/")
    }
  }

  cap <- max(vapply(mats,
                    function(m) stats::quantile(m, cap_quantile, na.rm = TRUE),
                    numeric(1)))
  message(sprintf("compute_unified_col_fun: %d sample(s), %d group(s), cap = %.4f",
                  length(mats), length(levels(group_factor)), cap))
  circlize::colorRamp2(c(0, cap), col_pal)
}

# Per-sample INDEPENDENT row sorting: each sample's heatmap is built
# standalone with its own per-group row order, sorted by that sample's
# enriched_score. Returns a NAMED LIST of EnrichedHeatmap objects.
# Render the list with draw_independent_heatmaps() below.
build_tornado_heatmap_independent <- function(tornado_data,
                                              display_groups = NULL,
                                              normalize = c("none", "sum_to_one", "background"),
                                              col_pal = c("white", "darkblue"),
                                              cap_quantile = 0.99,
                                              col_fun = NULL,
                                              use_raster = TRUE,
                                              group_colors = feature_group_colors,
                                              bg_factors = NULL,
                                              sample_groups = NULL,
                                              verbose = TRUE) {
  normalize    <- match.arg(normalize)
  mats         <- tornado_data$mats
  group_factor <- tornado_data$group_factor

  if (normalize == "background") mats <- apply_bg_norm(mats, bg_factors)

  if (normalize == "sum_to_one") {
    for (sn in names(mats)) {
      mat <- mats[[sn]]
      group_bin_means <- vapply(levels(group_factor), function(g) {
        colMeans(mat[group_factor == g, , drop = FALSE], na.rm = TRUE)
      }, numeric(ncol(mat)))
      bin_totals <- rowSums(group_bin_means, na.rm = TRUE)
      mats[[sn]] <- sweep(mat, 2, pmax(bin_totals, 1e-12), "/")
    }
  }

  if (!is.null(display_groups)) {
    keep_rows <- group_factor %in% display_groups
    mats <- lapply(mats, function(m) m[keep_rows, , drop = FALSE])
    group_factor <- factor(group_factor[keep_rows], levels = display_groups)
  }

  if (is.null(col_fun)) {
    cap <- max(vapply(mats,
                      function(m) stats::quantile(m, cap_quantile, na.rm = TRUE),
                      numeric(1)))
    col_fun <- circlize::colorRamp2(c(0, cap), col_pal)
  } else {
    cap <- environment(col_fun)$breaks[length(environment(col_fun)$breaks)]
  }
  legend_title <- switch(normalize,
    sum_to_one = "fraction",
    background = "fold over\nbackground",
    "signal")

  line_cols <- group_colors[levels(group_factor)]
  if (any(is.na(line_cols))) {
    fallback <- scales::hue_pal()(length(levels(group_factor)))
    line_cols[is.na(line_cols)] <- fallback[is.na(line_cols)]
  }

  if (verbose) {
    message(sprintf("  independent sort: %d sample heatmap(s), %d feature groups, color cap = %.4f",
                    length(mats), length(levels(group_factor)), cap))
  }

  heatmaps <- lapply(seq_along(mats), function(i) {
    sn  <- names(mats)[i]
    mat <- mats[[sn]]

    new_order <- unlist(lapply(levels(group_factor), function(g) {
      gi <- which(group_factor == g)
      gi[order(EnrichedHeatmap::enriched_score(mat[gi, , drop = FALSE]),
               decreasing = TRUE)]
    }), use.names = FALSE)
    mat_sorted <- mat[new_order, , drop = FALSE]
    gf_sorted  <- group_factor[new_order]

    # Row-block titles only on the leftmost panel — keeps right panels compact.
    row_title <- if (i == 1L)
                   levels(group_factor)
                 else
                   rep("", length(levels(group_factor)))

    # Column title. With `sample_groups` the 33 per-sample panels are labelled
    # by SAMPLE GROUP rather than by sample id: the name is drawn once, over
    # the middle panel of each run of same-group samples, and blank elsewhere.
    # `mats` must already be ordered by group for the runs to be contiguous —
    # asserted where sample_groups is built.
    col_title <- if (is.null(sample_groups)) {
      sn
    } else {
      g   <- sample_groups[[sn]]
      idx <- which(sample_groups[names(mats)] == g)
      if (i == idx[ceiling(length(idx) / 2)]) g else ""
    }

    EnrichedHeatmap::EnrichedHeatmap(
      mat_sorted,
      col          = col_fun,
      name         = sn,
      column_title = col_title,
      row_split    = gf_sorted,
      row_title    = row_title,
      cluster_rows = FALSE,
      use_raster   = use_raster,
      show_heatmap_legend  = (i == 1L),
      heatmap_legend_param = list(title = legend_title),
      top_annotation = ComplexHeatmap::HeatmapAnnotation(
        enriched = EnrichedHeatmap::anno_enriched(
          gp = grid::gpar(col = line_cols, lty = 1, lwd = 1.2)
        )
      )
    )
  })
  names(heatmaps) <- names(mats)
  heatmaps
}

# Render a list of standalone heatmaps side-by-side in one SVG. Each
# heatmap is drawn in its own grid viewport, so per-sample row orders
# are preserved (no shared alignment from the `+` operator).
draw_independent_heatmaps <- function(heatmaps,
                                       svg_path,
                                       column_title = NULL,
                                       panel_width_in = 2.5,
                                       height_in = 10) {
  n <- length(heatmaps)
  total_width <- n * panel_width_in + 1.5   # reserve space for legend

  svg(svg_path, width = total_width, height = height_in)
  grid::grid.newpage()

  if (!is.null(column_title)) {
    grid::grid.text(column_title, x = 0.5, y = 0.97,
                    gp = grid::gpar(fontsize = 12, fontface = "bold"))
    grid::pushViewport(grid::viewport(y = 0.48, height = 0.92))
  }

  grid::pushViewport(grid::viewport(
    layout = grid::grid.layout(nrow = 1, ncol = n)
  ))

  for (i in seq_along(heatmaps)) {
    grid::pushViewport(grid::viewport(layout.pos.col = i))
    ComplexHeatmap::draw(
      heatmaps[[i]],
      newpage = FALSE,
      show_heatmap_legend = (i == 1L),
      show_annotation_legend = (i == 1L)
    )
    grid::popViewport()
  }
  grid::popViewport()
  if (!is.null(column_title)) grid::popViewport()
  dev.off()
}
```

## (1b) Group-averaged tornado heatmap (5 columns: WT, PH, TN, NRT, NR)

```{r}
# Normalization: "sum_to_one" or "none"
heatmap_norm <- "background"   # "none" | "sum_to_one" | "background"

td_avg_rds <- file.path(plot_dir, sprintf("td_avg_1k_w20_%s.rds", set_tag))
# The cache is keyed only by filename, so a cache written before a track was
# added to bigwigs_avg — or before a feature set was added to set_spec —
# would silently come back short a column or a row block. Recompute unless
# it holds exactly the tracks AND the feature groups we asked for, in order.
cached_td_avg <- if (file.exists(td_avg_rds)) readRDS(td_avg_rds) else NULL
if (!is.null(cached_td_avg) &&
    !identical(names(cached_td_avg$mats), names(bigwigs_avg))) {
  message("Cached td_avg has tracks [", paste(names(cached_td_avg$mats), collapse = ", "),
          "] but bigwigs_avg is [", paste(names(bigwigs_avg), collapse = ", "),
          "] - recomputing.")
  cached_td_avg <- NULL
}
if (!is.null(cached_td_avg) &&
    !identical(levels(cached_td_avg$group_factor), set_order)) {
  message("Cached td_avg has groups [",
          paste(levels(cached_td_avg$group_factor), collapse = ", "),
          "] but set_order is [", paste(set_order, collapse = ", "),
          "] - recomputing.")
  cached_td_avg <- NULL
}
td_avg <- if (!is.null(cached_td_avg)) {
  message("Loading cached td_avg from ", td_avg_rds)
  cached_td_avg
} else {
  message("Computing td_avg (", length(bigwigs_avg),
          " bigwig imports + normalizeToMatrix)...")
  out <- compute_tornado_matrices(feats, bigwigs_avg, extend = 1000, w = 20)
  saveRDS(out, td_avg_rds)
  out
}

# Unified color scale across BOTH splits (diff and nd) of the avg heatmap.
avg_col_fun <- compute_unified_col_fun(td_avg, normalize = heatmap_norm,
                                       bg_factors = bg_avg)

# Panel membership comes from set_spec$split, so dropping an empty set in
# 8g removes it from the right split automatically.
splits_of <- function(which_split)
  set_order[vapply(set_spec, `[[`, character(1), "split") == which_split]
diff_groups <- splits_of("diff")   # inc_*, dec_* for every surviving label
nd_groups   <- splits_of("nd")     # nd_*  for every surviving label

# Row blocks per split are no longer fixed (Control is now its own set in the
# Increase row), so the canvas has to grow with them or the blocks compress.
# 2.5 in/block reproduces the historical 10 in at 4 blocks exactly, so old
# figures are unchanged and only the larger grids get taller.
split_height_in <- function(groups) max(10, 2.5 * length(groups))

# Split 1: differential — one row block per (call x label) that cleared
# min_set_size, so with a 3-label model this is up to 6, not the old 4.
draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_avg, display_groups = diff_groups,
    normalize = heatmap_norm, col_fun = avg_col_fun, bg_factors = bg_avg
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_averages_diff_%s_1k_w20_%s.svg",
                                heatmap_norm, set_tag)),
  column_title = sprintf("Differential (%s) — group averages [%s]",
                          paste(diff_groups, collapse = ", "), heatmap_norm),
  panel_width_in = 2.5,
  height_in      = split_height_in(diff_groups)
)

# Split 2: non-differential — one row block per surviving label
draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_avg, display_groups = nd_groups,
    normalize = heatmap_norm, col_fun = avg_col_fun, bg_factors = bg_avg
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_averages_nd_%s_1k_w20_%s.svg",
                                heatmap_norm, set_tag)),
  column_title = sprintf("Non-differential (%s) — group averages [%s]",
                          paste(nd_groups, collapse = ", "), heatmap_norm),
  panel_width_in = 2.5,
  height_in      = split_height_in(nd_groups)
)
```

## (2b) Per-sample tornado heatmap (33 columns, grouped by WT/PH/TN/NR)

```{r}
td_samples_rds <- file.path(plot_dir,
                            sprintf("td_samples_1k_w20_%s.rds", set_tag))
cached_td_samples <- if (file.exists(td_samples_rds)) readRDS(td_samples_rds) else NULL
# Same staleness guard as td_avg: a cache built with different tracks or a
# different set_spec must not be reused.
if (!is.null(cached_td_samples) &&
    (!identical(names(cached_td_samples$mats), names(bigwigs_samples)) ||
     !identical(levels(cached_td_samples$group_factor), set_order))) {
  message("Cached td_samples does not match the current tracks/feature groups - recomputing.")
  cached_td_samples <- NULL
}
td_samples <- if (!is.null(cached_td_samples)) {
  message("Loading cached td_samples from ", td_samples_rds)
  cached_td_samples
} else {
  message(sprintf("Computing td_samples (%d bigwig imports x %d feature sets)...",
                  length(bigwigs_samples), length(feats_for_samples)))
  out <- compute_tornado_matrices(feats_for_samples, bigwigs_samples,
                                   extend = 1000, w = 20)
  saveRDS(out, td_samples_rds)
  out
}

samples_col_fun <- compute_unified_col_fun(td_samples, normalize = heatmap_norm,
                                          bg_factors = bg_samples)

draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_samples, display_groups = diff_groups,
    normalize = heatmap_norm, col_fun = samples_col_fun,
    bg_factors = bg_samples, sample_groups = sample_groups
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_per_sample_diff_%s_1k_w20_%s.svg",
                                heatmap_norm, set_tag)),
  column_title = sprintf("Differential (%s) — per-sample [%s]",
                          paste(diff_groups, collapse = ", "), heatmap_norm),
  panel_width_in = 0.7,
  height_in      = split_height_in(diff_groups)
)

draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_samples, display_groups = nd_groups,
    normalize = heatmap_norm, col_fun = samples_col_fun,
    bg_factors = bg_samples, sample_groups = sample_groups
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_per_sample_nd_%s_1k_w20_%s.svg",
                                heatmap_norm, set_tag)),
  column_title = sprintf("Non-differential (%s) — per-sample [%s]",
                          paste(nd_groups, collapse = ", "), heatmap_norm),
  panel_width_in = 0.7,
  height_in      = split_height_in(nd_groups)
)
```
# 8k. Cosine quartiles within a fixed DESeq2 category

Every other figure in this document either selects on cosine (8g-8i) or
lets the DESeq2 call vary with it (8f), so none of them can separate
"the embedding recovers what DESeq2 already found" from "the embedding
adds something DESeq2 missed".

This one holds the DESeq2 category **fixed** and varies cosine inside it.
Take all `inc` regions, split them into quartiles by cosine to the Renin
label, and measure the renin-minus-non-renin signal in each. If Q4 > Q1
the embedding is ordering regions the fold-change call did not, and that
statement is immune to the circularity objection because every region
plotted carries the same DESeq2 label.

Two deliberate departures from 8g:

* **No intersection with a StarSpace label.** `inc_r` is already a
  cosine-based selection; using it would put cosine on both sides again.
  The input here is the raw `inc` / `dec` key set from 8c.
* **No top-N.** Quartiles are defined on the full category, so the cosine
  range is not truncated.

Signal comes from the same five group-average bigWigs as 8h, so the two
figures are directly comparable. Note this is bigWig coverage, not the
quantity DESeq2 modelled — for a number to quote in the text, recompute
the delta from the normalised count matrix that fed `dds`
(`1_featureCounts_quantilesNormalization.R`); the bigWigs are depth-
normalised per sample and then averaged, which is a different
normalisation.

```{r}
q_extend  <- 1000
q_w       <- 20
q_center  <- 200     # bp each side of the midpoint used for the scalar delta
q_renin     <- c("WT", "PH", "TN")     # renin-expressing group averages
q_non_renin <- c("NRT", "NR")          # non-renin group averages
stopifnot(all(c(q_renin, q_non_renin) %in% names(bigwigs_avg)))

# inc / dec as plain key vectors from 8c, turned back into GRanges.
keys_to_gr <- function(keys) {
  # Anchored at the END: 33 chromosomes in this universe contain underscores
  # (chr1_GL456210_random and friends, 270 regions), so splitting on "_" gives
  # a ragged list that do.call(rbind, .) silently recycles into wrong
  # coordinates. Same regex as write_set_bed() in 8e.
  m  <- stringr::str_match(keys, "^(.+)_([0-9]+)_([0-9]+)$")
  ok <- !is.na(m[, 1])
  if (any(!ok)) {
    warning(sum(!ok), " key(s) did not parse as <chrom>_<start>_<end>; dropped.")
  }
  GenomicRanges::GRanges(
    seqnames = m[ok, 2],
    ranges   = IRanges::IRanges(as.integer(m[ok, 3]), as.integer(m[ok, 4])))
}

cos_renin_col <- cos_col(focus_label)
cos_r_lookup  <- key_cos[[cos_renin_col]]

# One feature set per (category, quartile).
build_quartiles <- function(keys, tag, n_q = 4) {
  cv   <- cos_r_lookup[keys]
  keep <- is.finite(cv)
  keys <- keys[keep]; cv <- cv[keep]
  if (!length(keys)) return(list())
  qs <- stats::quantile(cv, probs = seq(0, 1, length.out = n_q + 1), na.rm = TRUE)
  qs[1] <- -Inf; qs[length(qs)] <- Inf
  bin <- cut(cv, unique(qs), include.lowest = TRUE, labels = FALSE)
  out <- lapply(seq_len(n_q), function(q) keys_to_gr(keys[which(bin == q)]))
  names(out) <- sprintf("%s_Q%d", tag, seq_len(n_q))
  out[lengths(out) > 0]
}

q_feats <- c(build_quartiles(inc, "inc"), build_quartiles(dec, "dec"))
message("Cosine-quartile feature sets (ranked by ", cos_renin_col, "):\n",
        paste(sprintf("  %-8s %7s regions", names(q_feats),
                      format(lengths(q_feats), big.mark = ",")),
              collapse = "\n"))

q_rds <- file.path(plot_dir, "df_cos_quartiles_1k_w20.rds")
if (file.exists(q_rds)) {
  df_q <- readRDS(q_rds)
  message("Loaded cached quartile profiles: ", q_rds)
} else {
  df_q <- flatten_features_eh(q_feats, bigwigs_avg,
                              extend = q_extend, w = q_w)
  saveRDS(df_q, q_rds)
}

df_q$category <- sub("_Q[0-9]+$", "", df_q$feature_set)
df_q$quartile <- as.integer(sub("^.*_Q", "", df_q$feature_set))
```

## (1) Profiles by quartile

```{r, fig.width = 11, fig.height = 6}
p_q_prof <- ggplot(df_q,
                   aes(position, mean,
                       colour = factor(quartile),
                       group  = interaction(quartile, sample_name))) +
  geom_line(linewidth = 0.5) +
  facet_grid(category ~ sample_name) +
  scale_colour_viridis_d(name = sprintf("%s quartile\n(Q4 = closest)", focus_label),
                         option = "C", end = 0.9) +
  labs(x = "distance from region midpoint (bp)",
       y = "mean bigWig coverage",
       title = sprintf("Signal by %s cosine quartile, DESeq2 category held fixed",
                       focus_label)) +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

ggsave(file.path(plot_dir, "starspace_cos_quartile_profiles.svg"),
       p_q_prof, width = 11, height = 6)

```

## (2) Renin minus non-renin delta per quartile

The scalar is the mean over the central +/-`q_center` bp. Averaging
positions and then regions is the same number as averaging regions and
then positions, so this is exact for the mean; the ribbon is a
across-region SEM propagated from the per-position `sd` and `n` that
`flatten_features_eh` returns, and it ignores the correlation between
adjacent bins, so read it as indicative rather than as a test.

```{r, fig.width = 7, fig.height = 4}
q_core <- subset(df_q, abs(position) <= q_center)

q_grp <- q_core %>%
  mutate(arm = ifelse(sample_name %in% q_renin, "renin", "non_renin")) %>%
  group_by(category, quartile, arm) %>%
  summarise(sig = mean(mean),
            sem = sqrt(sum((sd / sqrt(n))^2)) / n(),
            .groups = "drop")

q_delta <- q_grp %>%
  tidyr::pivot_wider(names_from = arm, values_from = c(sig, sem)) %>%
  mutate(delta = sig_renin - sig_non_renin,
         delta_sem = sqrt(sem_renin^2 + sem_non_renin^2))

p_q_delta <- ggplot(q_delta, aes(factor(quartile), delta,
                                 group = category, colour = category)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = delta - delta_sem, ymax = delta + delta_sem),
                width = 0.15) +
  geom_line() + geom_point(size = 2.5) +
  labs(x = sprintf("%s cosine quartile (Q4 = closest to label)", focus_label),
       y = sprintf("mean coverage, renin (%s) - non-renin (%s)",
                   paste(q_renin, collapse = "/"),
                   paste(q_non_renin, collapse = "/")),
       title = "Does embedding distance order regions within a fixed DESeq2 call?") +
  theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank())

ggsave(file.path(plot_dir, "starspace_cos_quartile_delta.svg"),
       p_q_delta, width = 7, height = 4)


knitr::kable(q_delta, digits = 4)
```

# 8j. GO enrichment weighted-rank plot — refined Renin vs Non-renin regions

Same GREAT enrichment + renin-term weighted-rank procedure as
`4_greatGOenrichment.Rmd` sections 4d/4e, but applied to the refined
region classification that combines both DESeq2 and StarSpace evidence:

- **Renin** = `inc_r` — regions that (1) showed increased accessibility
  in renin samples in the DESeq2 analysis AND (2) fell in the StarSpace
  Renin cluster from 8b.
- **Non-renin** = everything else in the StarSpace cluster universe,
  i.e. `inc_nr ∪ dec_r ∪ dec_nr ∪ nd_r ∪ nd_nr`. These are regions that
  failed at least one of the two criteria above.

Background: genome-wide null (no explicit `bg` arg to GREAT). Using the
cluster union as background here would push foreground ≈ background for
the Non-renin set (it covers ~99% of the universe), collapsing every
hypergeometric p-value toward 1 and emptying the post-filter table. The
genome-wide null lets both sets produce real enrichment tables that the
weighted-rank step can score against each other.

## Load packages

```{r}
library(rGREAT)
```

## Build refined Renin / Non-renin GRanges from 8e intersection BEDs

Reload from disk so this chunk is independent of the in-memory `gr_list`
/ `inc_r` / etc., which 8g has already narrowed by top-N and by the
cosine ranking. 8j is still a two-sided Renin-vs-rest comparison and so
keeps using the pooled `_nr` BEDs that 8e writes alongside the per-label
ones; `inc_nr ∪ dec_nr ∪ nd_nr` is exactly the Control ∪ Tumoral union
that 8g now handles as separate sets.

```{r}
load_bed_8j <- function(path) {
  readGeneric(path, chr = 1, start = 2, end = 3, strand = NULL,
              meta.cols = NULL, keep.all.metadata = FALSE,
              zero.based = FALSE, remove.unusual = FALSE,
              header = FALSE, skip = 0, sep = "\t")
}

inc_r_full  <- load_bed_8j(file.path(bed_dir, "inc_r.bed"))
inc_nr_full <- load_bed_8j(file.path(bed_dir, "inc_nr.bed"))
dec_r_full  <- load_bed_8j(file.path(bed_dir, "dec_r.bed"))
dec_nr_full <- load_bed_8j(file.path(bed_dir, "dec_nr.bed"))
nd_r_full   <- load_bed_8j(file.path(bed_dir, "nd_r.bed"))
nd_nr_full  <- load_bed_8j(file.path(bed_dir, "nd_nr.bed"))

renin_refined_gr <- inc_r_full
non_renin_gr     <- c(inc_nr_full, dec_r_full, dec_nr_full,
                      nd_r_full,  nd_nr_full)

message(sprintf(
  "Refined sets: Renin = %s regions (inc_r), Non-renin = %s regions (everything else)",
  format(length(renin_refined_gr), big.mark = ","),
  format(length(non_renin_gr),     big.mark = ",")
))

ss_query <- GenomicRanges::GRangesList(
  Renin     = renin_refined_gr,
  Non_renin = non_renin_gr
)
```

## Run GREAT GO Biological Process enrichment

```{r}
ss_getBioEnrichment <- function(gr) {
  # No explicit background: when the Non-renin set covers ~99% of the
  # cluster universe, foreground ≈ background and every hypergeometric
  # p-value collapses toward 1, leaving the filter step with zero rows.
  # Testing each set against the genome-wide null lets both produce real
  # enrichment tables that the weighted-rank step can compare.
  job <- rGREAT::submitGreatJob(gr, species = "mm10")
  tb  <- rGREAT::getEnrichmentTables(job)
  tb$`GO Biological Process`
}

ss_enrichment <- lapply(names(ss_query), function(s) {
  ss_getBioEnrichment(ss_query[[s]])
})
names(ss_enrichment) <- names(ss_query)
```

## Weighted rank of renin-related GO terms

```{r}
# Shared with 4_greatGOenrichment.Rmd. 8j used to carry its own `ss_*` copies
# that had drifted back to the pre-fix versions of all four functions -- see
# the header of go_enrichment_functions.R for what each of them got wrong.
source("go_enrichment_functions.R")
```

```{r}
# Same order of operations as 4_greatGOenrichment.Rmd 4d/4e, which matters:
#
#   go_results_all  FULL tables + a `sig` flag  -> what getRank/getRankWeight see
#   go_results      significance-filtered       -> ONLY used to choose terms
#
# 8j used to rank and weight the FILTERED tables. A term clearing the filter in
# one region set but not the other then contributed 0 for the other set,
# term_sum collapsed to one value, and rank_weight came out exactly 1.0 — a
# spurious 100% bar that looked identical to a genuinely exclusive term.
ss_all <- lapply(ss_enrichment, function(df) {
  df$sig <- !is.na(df$Hyper_Adjp_BH) &
            df$Hyper_Adjp_BH <= 0.1 &
            df$Hyper_Foreground_Gene_Hits >= 3
  df
})

# filterEnrichment() keys on Hyper_Foreground_Gene_Hits, the column
# 4_greatGOenrichment.Rmd has actually run against. Fail with the real column
# list rather than a bare object-not-found if the GREAT server ever renames it.
need <- c("Hyper_Adjp_BH", "Hyper_Foreground_Gene_Hits", "Hyper_Fold_Enrichment",
          "name", "ID")
absent <- setdiff(need, names(ss_enrichment[[1]]))
if (length(absent)) {
  stop("GREAT table is missing: ", paste(absent, collapse = ", "),
       "\n  Columns present: ", paste(names(ss_enrichment[[1]]), collapse = ", "),
       "\n  (older rGREAT called the hits column Hyper_Observed_Gene_Hits)")
}

ss_go <- lapply(ss_all, filterEnrichment)

# Guard: empty enrichment table after filtering usually means foreground ≈
# background (collapsing hypergeometric p-values toward 1). Print sizes so
# this is easy to diagnose if it happens.
n_rows_per_set <- sapply(ss_go, nrow)
message("Filtered enrichment table sizes: ",
        paste(sprintf("%s = %d", names(n_rows_per_set), n_rows_per_set),
              collapse = ", "))
if (any(n_rows_per_set == 0)) {
  stop("Empty enrichment table(s): ",
       paste(names(n_rows_per_set)[n_rows_per_set == 0], collapse = ", "),
       ". Check the foreground vs background sizes in ss_getBioEnrichment().")
}

# Terms from the FILTERED tables (significant in at least one set), by GO graph
# ancestry — not the old keyword regex.
ss_renin_terms <- getReninGOTerms(ss_go)
message(sprintf("Renin-related GO terms (GO-ancestry): %d", length(ss_renin_terms)))

# Rank and weight on the FULL tables.
ss_rank_dfs <- lapply(names(ss_all), function(s) getRank(ss_all[[s]], group = s))
names(ss_rank_dfs) <- names(ss_all)
ss_rank_dfs <- getRankWeight(ss_rank_dfs, ss_renin_terms)

ss_plot_df <- do.call(rbind, ss_rank_dfs)
ss_plot_df <- ss_plot_df[!is.na(ss_plot_df$name), ]
rownames(ss_plot_df) <- NULL

# getRank() is fold-only by design — the two region sets differ greatly in
# size, so their hypergeometric p-values are not comparable across groups.
# The old rank_pvalue / rank_pvalue_weight columns no longer exist.
ss_plot_df <- ss_plot_df %>%
  select(name, Group, sig,
         Rank_weight = rank_weight,
         Rank_fold   = rank_fold,
         Fold        = Hyper_Fold_Enrichment)

ss_plot_df$Group <- factor(ss_plot_df$Group, levels = c("Renin", "Non_renin"))

ss_term_order <- ss_plot_df %>%
  filter(Group == "Renin") %>%
  arrange(desc(Rank_weight)) %>%
  mutate(labels = factor(name)) %>%
  head(50)

ss_plot_df <- filter(ss_plot_df, name %in% ss_term_order$name)
ss_new_order <- ss_plot_df %>%
  filter(Group == "Renin") %>%
  arrange(desc(Rank_weight)) %>%
  mutate(labels = factor(name))
```

## Plot stacked bar of weighted rank

```{r, fig.width = 8, fig.height = 7}
output_file <- file.path(plot_dir, "refined_renin_go_rank_weight_stacked_bar.svg")
dir.create(dirname(output_file), showWarnings = FALSE, recursive = TRUE)

n_terms <- length(unique(ss_plot_df$name))

p_ss_go <- ss_plot_df %>%
  mutate(labels = factor(name, levels = ss_new_order$labels)) %>%
  ggplot(aes(fill = Group, y = Rank_weight, x = labels)) +
  geom_bar(position = "fill", stat = "identity", width = 0.9) +
  scale_fill_manual(values = c(Renin = "#e1812c", Non_renin = "lightgrey")) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6, color = "black"),
    axis.text.y = element_text(size = 6, color = "black"),
    axis.title  = element_text(size = 8, color = "black"),
    panel.grid  = element_blank()
  ) +
  labs(y = "Enrichment Weight", x = "GO Terms",
       title = "Refined Renin (inc_r) vs Non-renin (everything else) — renin-related GO terms")

ggsave(output_file, plot = p_ss_go,
       width = max(4, n_terms * 0.1), height = 7, units = "in")

p_ss_go
```

# 8l. Top GO terms per StarSpace cluster

One panel per StarSpace label cluster — `Renin`, `Control`, `Tumoral` —
showing that cluster's top enriched GO Biological Process terms.

This is a deliberately simpler question than 8j. 8j scores *renin-related*
terms on the **refined** Renin set (`inc_r`, i.e. DESeq2 ∩ StarSpace) against
everything else. 8l drops the DESeq2 axis entirely and asks what each cluster
of the embedding is enriched for on its own terms, using the cluster
membership from 8b (`ss_sets`) with no DESeq2 intersection.

`go_top_n` selects between two foregrounds, and the chunk is written to
produce either from the same code:

* **`NULL` — the whole cluster.** Every region assigned to that label, all
  68,503 / 212,698 / 141,287 of them. This is the honest description of what
  the embedding partitioned the universe into.
* **an integer — the N regions closest to that cluster's own label** by cosine
  similarity to the label embedding (`cos_mat`, computed in 8f). Cluster
  assignment is already nearest-label, so this keeps the same ranking and just
  cuts it off at N: the cluster's core rather than its whole basin. A cluster
  of 212,698 regions is half the universe and its enrichment is diluted by
  everything sitting near the decision boundary; the top N is the sharper
  question, at the cost of no longer describing the full cluster.

The two are written to different files and neither overwrites the other, so
they can be read side by side.

**Background is the cluster universe** (all three clusters pooled = the
consensus peak universe the model was trained on), not the genome-wide null
8j is forced to use. The three clusters partition that universe, so no cluster
is anywhere near 100% of it — `Renin` is 16%, `Tumoral` 33%, `Control` 50% —
and the collapse-toward-p=1 problem that made 8j use the genome-wide null does
not arise here. Using the universe matters: against a genome-wide null every
ATAC peak set enriches for "sits near a gene", and all three panels would
return near-identical developmental terms. Against the universe those cancel
and what is left is what separates one cluster from the other two.

**GREAT runs locally**, via `rGREAT::great()`, and not by choice: the online
server `4_greatGOenrichment.Rmd` submits to with `submitGreatJob()` **no
longer exists**. `great.stanford.edu` returns NXDOMAIN from public DNS
(checked against 8.8.8.8 from a compute node, 2026-09-06), while
`stanford.edu` and `bioconductor.org` resolve normally from the same node —
so this is the hostname being gone, not a cluster firewall. 4c/4d in
`4_greatGOenrichment.Rmd` cannot be re-run as written; its existing outputs
predate the shutdown.

Local mode applies the same `basalPlusExt` regulatory-domain rule (5 kb
upstream / 1 kb downstream, extended to 1 Mb) against
`TxDb.Mmusculus.UCSC.mm10.knownGene` TSSs, so the enrichment model is GREAT's,
but the column names differ from the web server's (`fold_enrichment_hyper` /
`p_adjust_hyper` here vs `Hyper_Fold_Enrichment` / `Hyper_Adjp_BH` there) and
the GO annotation comes from the locally installed `org.Mm.eg.db` rather than
whatever vintage the server held. Numbers are therefore close to, but not
byte-identical with, what 4c produced.

Ranking follows `4_greatGOenrichment.Rmd` 4c (`plotTopEnrichedGO`): keep terms
with fold enrichment ≥ `go_fold_min` (2, as in 4c) and ≥ 3 foreground gene
hits, then take the smallest BH-adjusted p-values.

**That filter empties the table for the whole-cluster foreground, and the
reason is structural.** `fold_enrichment_hyper` is
`observed_gene_hits / (gene_set_size * n_gene_gr / n_gene_total)`. A
foreground covering a large fraction of the background touches nearly every
gene, so `observed_gene_hits` saturates at `gene_set_size` and the ratio is
pinned near 1 no matter how real the biology is. Measured on the whole-cluster
run (job 19384856): the **maximum** fold over ~9,280 terms was 1.24 for Renin
(16% of the universe), 1.05 for Control (50%) and 1.08 for Tumoral (33%) — and
for Control and Tumoral that maximum equals the third quartile, which is the
saturation signature. Nothing anywhere near 2 exists to be kept.

So this is not a threshold to lower; a "fold enrichment" axis spanning
1.00–1.05 measures the foreground's size, not its biology. It is the reason
`go_top_n` exists: at 5,000 regions each cluster is ~1.2% of the universe and
the gene hits stop saturating. The p-values remain informative in both cases
(2,018 / 986 / 489 terms at BH ≤ 0.05 on the whole clusters), which is why the
failure shows up only in the fold channel. The hypergeometric (gene-based) columns are used, which
are the `Hyper_*` columns 4c and 8j rank on; local `great()` calls them
`fold_enrichment_hyper` / `p_adjust_hyper` and reports the binomial
region-based test alongside them in the same table.

## Cluster region sets and background

```{r}
# ss_sets comes from 8b and is keyed by the trained label; keys_to_gr() is the
# underscore-safe key parser defined in 8k (33 chromosomes in this universe
# have underscores in their names, so a plain strsplit gives wrong coordinates).
cluster_go_labels <- intersect(cluster_labels, names(ss_sets))

# NULL = whole cluster; an integer = that cluster's N regions closest to its
# own label. The batch runners in src/cluster_go_run/ pass this through the
# GO_TOP_N environment variable, so both variants come from this one chunk
# rather than from a forked copy of it.
go_top_n <- if (nzchar(Sys.getenv("GO_TOP_N"))) {
  as.integer(Sys.getenv("GO_TOP_N"))
} else {
  NULL
}
go_tag <- if (is.null(go_top_n)) "" else sprintf("_top%d", go_top_n)

# cos_mat (8f) is regions x labels. It carries label colnames but NO rownames
# -- 8f keeps the region names beside it in emb$region_names and joins by
# column -- so index it through an explicit name -> row lookup rather than by
# character subscript, which silently fails with "subscript out of bounds".
cos_row_of <- if (is.null(go_top_n)) NULL else {
  stats::setNames(seq_along(emb$region_names), emb$region_names)
}

# Rank each cluster's members by cosine to THAT cluster's label, not to
# focus_label, so every panel is cut the same way relative to its own centre.
take_cluster_core <- function(keys, label, n) {
  if (is.null(n)) return(keys)
  stopifnot(label %in% colnames(cos_mat))
  ri <- cos_row_of[keys]
  if (anyNA(ri)) {
    stop(sum(is.na(ri)), " region(s) of cluster ", label,
         " are absent from model.tsv; cos_mat and ss_sets disagree.")
  }
  cv <- cos_mat[ri, label]
  keys[order(cv, decreasing = TRUE)][seq_len(min(n, length(keys)))]
}

cluster_go_keys <- lapply(cluster_go_labels, function(lab)
  take_cluster_core(ss_sets[[lab]], lab, go_top_n))
names(cluster_go_keys) <- cluster_go_labels
cluster_go_gr <- lapply(cluster_go_keys, keys_to_gr)

# Union of all clusters = the consensus universe. Built from ss_sets, NOT from
# cluster_go_keys: the background is the whole universe in both variants, so a
# top-N foreground is tested against everything rather than against a
# correspondingly shrunken background, and the two figures stay comparable.
cluster_go_bg <- keys_to_gr(unlist(ss_sets[cluster_go_labels], use.names = FALSE))

message(sprintf("8l foregrounds (%s): %s | background: %s regions",
                if (is.null(go_top_n)) "whole cluster"
                  else sprintf("top %d closest to own label", go_top_n),
                paste(sprintf("%s=%s", names(cluster_go_gr),
                              format(lengths(cluster_go_gr), big.mark = ",")),
                      collapse = ", "),
                format(length(cluster_go_bg), big.mark = ",")))

# Same GO_enrichment/ directory the 4_greatGOenrichment.Rmd figures live in.
go_plot_dir <- file.path(plot_dir, "GO_enrichment")
dir.create(go_plot_dir, showWarnings = FALSE, recursive = TRUE)
```

## Run local GREAT, one table per cluster

```{r}
# exclude = NULL, not the "gap" default: resolving the default fetches the mm10
# gap track from UCSC.
#
# Cost is NOT driven by the foreground size, which is the intuitive guess and
# is wrong. Measured on this data with 8 cores: three whole clusters
# (68,503 + 212,698 + 141,287 regions) took 23m47s, three top-5,000
# foregrounds took 13m18s. A 40x smaller foreground bought 1.8x, because the
# per-cluster fixed cost -- extending the TSS set, overlapping the
# 422,488-region background, and scoring 14,957 gene sets -- is paid whatever
# the foreground size. Shrinking go_top_n further is a sharper QUESTION, not a
# faster run.
#
# Cached, so the figure can be re-cut without re-running any of that. Delete
# the RDS to force a refresh after re-clustering.
cluster_go_rds <- file.path(go_plot_dir,
                            sprintf("cluster_go_bp_tables%s.rds", go_tag))

# Parallel over gene sets. Under SLURM take the allocation; interactively stay
# single-core rather than forking 40 workers on a shared login node.
great_cores <- max(1L, as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "1")))

if (file.exists(cluster_go_rds)) {
  cluster_go_tabs <- readRDS(cluster_go_rds)
  message("Loaded cached GREAT tables: ", cluster_go_rds)
} else {
  cluster_go_tabs <- lapply(names(cluster_go_gr), function(lab) {
    message("GREAT (local) for cluster ", lab, " ...")
    res <- rGREAT::great(cluster_go_gr[[lab]],
                         gene_sets   = "GO:BP",
                         tss_source  = "TxDb.Mmusculus.UCSC.mm10.knownGene",
                         background  = cluster_go_bg,
                         exclude     = NULL,
                         min_gene_set_size = 5,
                         cores       = great_cores)
    rGREAT::getEnrichmentTable(res)
  })
  names(cluster_go_tabs) <- names(cluster_go_gr)
  saveRDS(cluster_go_tabs, cluster_go_rds)
}

stopifnot(all(c("description", "fold_enrichment_hyper", "p_adjust_hyper",
                "observed_gene_hits") %in% names(cluster_go_tabs[[1]])))
```

## Plot: p-value bars + fold-enrichment line, one panel per cluster

```{r, fig.width = 3.74, fig.height = 1.77}
go_n_terms    <- 5    # terms per panel; 3 x 5 bars is what fits in 95 mm
go_fold_min   <- if (nzchar(Sys.getenv("GO_FOLD_MIN"))) {
  as.numeric(Sys.getenv("GO_FOLD_MIN"))    # 4c's threshold; see the prose on
} else {                                   # why relaxing it is usually wrong
  2
}
go_name_chars <- 20   # x labels are vertical, so this is a HEIGHT budget

# The vertical x labels are the binding constraint on a 45 mm figure, and the
# numbers here are measured off rendered output, not estimated: at 6 pt one
# character is ~1.2 mm tall, so 20 characters cost ~22 mm of the 45 and leave
# the panel ~13 mm. 22 characters already squeezes the panel to ~10 mm, which
# is not enough for a dual axis.
#
# Wrapping the labels onto two shorter lines was tried and is worse. It buys
# back ~7 mm of height, but two 6 pt lines are ~4.2 mm wide against a ~5 mm bar
# pitch, so neighbouring labels very nearly touch and it stops being obvious
# which line belongs to which bar. One truncated line per bar reads cleanly.
#
# Truncation is lossy for GO BP names, so the full untruncated names are in the
# table printed under this chunk -- read the figure with that table.

# 4c's filter, on the local-GREAT column names.
pick_top_go <- function(tb, lab, n = go_n_terms) {
  keep <- subset(tb, is.finite(p_adjust_hyper) &
                     fold_enrichment_hyper >= go_fold_min &
                     observed_gene_hits >= 3)
  if (!nrow(keep)) {
    warning(sprintf(
      "No GO term passes the 4c filter for cluster %s (fold >= %g, >= 3 gene hits). %s",
      lab, go_fold_min,
      sprintf("Max fold over %s terms was %.3f.",
              format(nrow(tb), big.mark = ","),
              max(tb$fold_enrichment_hyper, na.rm = TRUE))))
    return(NULL)
  }
  keep <- head(keep[order(keep$p_adjust_hyper), , drop = FALSE], n)
  # Foregrounds of 68k-213k regions drive adjusted p-values below double
  # precision, and p == 0 would make -log10() infinite and blow up the y
  # scale. Floor at the smallest representable double (-log10 = 307.65) and
  # say so, rather than plotting Inf or silently dropping the strongest term.
  floored <- keep$p_adjust_hyper < .Machine$double.xmin
  if (any(floored)) {
    message(sprintf("  %s: %d term(s) at p_adjust_hyper = 0, floored to %.3g",
                    lab, sum(floored), .Machine$double.xmin))
  }
  data.frame(cluster = lab,
             term    = keep$description,
             negLogP = -log10(pmax(keep$p_adjust_hyper, .Machine$double.xmin)),
             fold    = keep$fold_enrichment_hyper,
             stringsAsFactors = FALSE)
}

go_df <- do.call(rbind, lapply(names(cluster_go_tabs), function(lab)
  pick_top_go(cluster_go_tabs[[lab]], lab)))

# Fail here, with the numbers, rather than let an empty frame reach order()
# and surface as "order(integer(0), NULL, ...)" 40 minutes into a batch job.
if (is.null(go_df) || !nrow(go_df)) {
  stop("No GO term passes the filter in ANY cluster. Max fold_enrichment_hyper: ",
       paste(sprintf("%s = %.3f", names(cluster_go_tabs),
                     vapply(cluster_go_tabs,
                            function(d) max(d$fold_enrichment_hyper, na.rm = TRUE),
                            numeric(1))), collapse = ", "),
       " against go_fold_min = ", go_fold_min,
       ".\n  A whole-cluster foreground saturates the gene hits and pins the ",
       "fold near 1 -- set go_top_n (GO_TOP_N) rather than lowering the ",
       "threshold. See the prose above this chunk.")
}

go_df$cluster <- factor(go_df$cluster, levels = names(cluster_go_tabs))
go_df <- go_df[order(go_df$cluster, go_df$negLogP, decreasing = c(FALSE, TRUE),
                     method = "radix"), ]

# One x level per ROW, not per term: a term enriched in two clusters would
# otherwise share a level and be drawn in both panels by facet_wrap.
go_df$x <- factor(seq_len(nrow(go_df)))
go_labels <- ifelse(nchar(go_df$term) > go_name_chars,
                    paste0(substr(go_df$term, 1, go_name_chars - 1), "\u2026"),
                    go_df$term)
names(go_labels) <- go_df$x

# ggplot has ONE y scale, so the secondary axis needs a single global factor.
# That is the better choice anyway: it keeps fold enrichment comparable across
# the three panels instead of rescaling each one independently.
go_scale <- max(go_df$negLogP) / max(go_df$fold)

p_cluster_go <- ggplot(go_df, aes(x = x)) +
  geom_col(aes(y = negLogP), fill = "grey75", colour = NA, width = 0.75) +
  geom_line(aes(y = fold * go_scale, group = 1),
            colour = "black", linewidth = 0.25) +
  geom_point(aes(y = fold * go_scale), shape = 21, colour = "black",
             fill = "white", size = 0.8, stroke = 0.3) +
  facet_wrap(~ cluster, nrow = 1, scales = "free_x") +
  scale_x_discrete(labels = go_labels) +
  scale_y_continuous(
    name     = expression(-log[10] * "(FDR)"),
    expand   = expansion(mult = c(0, 0.08)),
    # Two lines, not one: a 15-character axis title at 8 pt is ~23 mm tall
    # against a ~13 mm panel, so "fold enrichment" on one line has its top
    # clipped off by the device edge. Splitting it spends horizontal space,
    # which this figure has.
    sec.axis = sec_axis(~ . / go_scale, name = "fold\nenrichment")) +
  theme_bw(base_size = 8) +
  theme(
    text             = element_text(size = 8, colour = "black"),
    axis.text.x      = element_text(angle = 90, vjust = 0.5, hjust = 1,
                                    size = 6, colour = "black"),
    axis.text.y      = element_text(size = 6, colour = "black"),
    axis.title.x     = element_blank(),
    axis.title.y     = element_text(size = 8, colour = "black"),
    axis.ticks       = element_line(linewidth = 0.25, colour = "black"),
    axis.ticks.length = unit(0.6, "mm"),
    strip.text       = element_text(size = 8, colour = "black",
                                    margin = margin(0.3, 0.8, 0.3, 0.8, "mm")),
    strip.background = element_rect(fill = NA, colour = "black",
                                    linewidth = 0.25),
    panel.grid       = element_blank(),
    panel.border     = element_rect(colour = "black", linewidth = 0.25),
    panel.spacing    = unit(1.5, "mm"),
    plot.margin      = margin(1, 1, 1, 1, "mm"),
    legend.position  = "none")

# Named apart from 4c's top_enriched_go_*.svg, which are the DESeq2 sets.
cluster_go_svg <- file.path(
  go_plot_dir, sprintf("top_enriched_go_starspace_clusters%s.svg", go_tag))
ggsave(cluster_go_svg, p_cluster_go, width = 95, height = 45, units = "mm")
message("Saved: ", cluster_go_svg)

knitr::kable(go_df[, c("cluster", "term", "negLogP", "fold")], digits = 3)

p_cluster_go
```
