---
title: "8. DESeq2 vs. StarSpace differentially-accessible regions"
author: "xue"
date: "2026-05-22"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

This document 1) builds region-class assignments directly from a trained
StarSpace model via k-means clustering, 2) compares them to DESeq2
differentially-accessible region sets, and 3) plots tornado signal
profiles of motif-bound regions stratified by both axes.

# 8a. Setup environment

## Load packages

```{r}
library(genomation)
library(GenomicDistributions)
library(GenomicDistributionsData)
library(UpSetR)
library(tidyverse)
library(venn)
library(EnrichedHeatmap)
library(rtracklayer)
library(stringr)
library(ggplot2)
```

## Set working directory

```{r, setup, include=FALSE}
wd = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/differential/geniml_univ_cc"
knitr::opts_knit$set(root.dir = wd)
```

## Define file paths

### DESeq2 input

```{r}
primary_dir = "primary"
tumoral_dir = "tumoral"

# DESeq2 differential-accessibility reports
primary_normal_file = "deseq/deseq_normal_primary_report_shrink.csv"
as_normal_file      = "deseq/deseq_normal_tumoral_report_shrink.csv"
primary_high_file   = "deseq/deseq_high_primary_report_shrink.csv"
```

### StarSpace input

```{r}
# Trained StarSpace model
starspace_model_path = "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline/reninness_score_output_new/universe_cc/d50_e10_m0.1/model_d50_e10_m0.1.tsv"
```

### Universe

```{r}
universe_file = "/home/bx2ur/code/reninness_score/data/universe/universe_cc.bed"
```

### Output

```{r}
bed_dir = "bed"
plot_dir = "plot"
```

### Metadata

```{r}
meta = "~/code/renin_atac/metadata/renin_encode_atac.csv"
```

# 8b. Build StarSpace region clusters from trained model

`cluster_starspace_regions()` reads the trained model's word/
embeddings, k-means-clusters the region rows (skipping the last 2
trained label rows), then maps each cluster to `Renin` / `Non-renin` by
cosine similarity between the cluster centroid and the trained label
vectors. K-means is O(n·k·i) --- much faster than DBSCAN's O(n²) at
typical region counts (10k--1M) and produces a clean 2-way split that
mirrors the model's two trained classes.

## Define functions

```{r}
cluster_starspace_regions <- function(
    model_path,
    n_clusters = 2,
    method = c("spherical_kmeans", "cosine_nearest", "hdbscan", "umap_hdbscan"),
    seed = 42,
    renin_label = "Renin",
    non_renin_label = "Non_Renin",
    output_renin = "Renin",
    output_non_renin = "Non-renin",
    output_other = "other",
    margin_threshold = 0,  # |cos_renin - cos_non_renin| < this -> output_other
                           # Applies in cosine_nearest method AND in the post-hoc
                           # cosine_override step (so HDBSCAN/k-means results
                           # also get a "other" band when margin > 0).
    minPts = 50,          # smaller = more sub-clusters along elongated streaks
    hdb_max_n = 20000,    # subsample cap for HDBSCAN (O(n²) memory at scale)
    hdb_pca_dim = 15,     # PCA-reduce to this dim before HDBSCAN (NULL = skip)
    label_method = c("cluster_cosine_nearest", "one_per_label"),
    init_at_labels = FALSE, # spherical_kmeans: TRUE to seed centroids at label vectors
                           # (only useful when labels sit near data centers — for the
                           # typical hinge-loss case where labels land at cluster tips,
                           # this biases the partition. Keep FALSE.)
    cosine_override = TRUE, # post-hoc: flip a region's label if its per-region cosine
                            # to the OTHER label vector exceeds cosine to the assigned
                            # one. Targets streak-tip / boundary mis-assignments from
                            # centroid propagation. Set FALSE to disable.
    spatial_smooth = TRUE,           # grow "other" via k-NN: regions whose neighbors
                                     # are mostly "other" also become "other". Captures
                                     # the geometric extent of the sub-cluster even when
                                     # individual cosine values exceed margin_threshold.
                                     # Requires margin_threshold > 0 and FNN package.
    spatial_smooth_k = 15,           # neighbors per region for propagation
    spatial_smooth_thresh = 0.5,     # fraction of neighbors that must be "other"
    spatial_smooth_max_iter = 5,     # iteration cap (label propagation converges fast)
    umap_subcluster = c("hdbscan", "kmeans") # only used by method = "umap_hdbscan".
                                             # "kmeans" forces n_clusters groups in the
                                             # 2D UMAP space (useful when sub-clusters
                                             # are density-connected so HDBSCAN merges
                                             # them).
) {
  umap_subcluster <- match.arg(umap_subcluster)
  label_method <- match.arg(label_method)
  method <- match.arg(method)
  # StarSpace's `model.tsv` is one row per token, no header:
  #   <token>\t<v1>\t...\t<vd>
  # Tokens are in vocabulary order, which depends on training config —
  # label rows ("__label__Renin", "__label__Non_Renin") are NOT
  # guaranteed to sit at the bottom. Identify labels by the "__label__"
  # prefix instead of by position (matches geniml's evaluate.py).
  embed <- read.table(
    model_path, sep = "\t", header = FALSE,
    comment.char = "", quote = "", stringsAsFactors = FALSE,
    fill = TRUE
  )
  embed <- embed[!is.na(embed[, 1]) & embed[, 1] != "", , drop = FALSE]

  is_label    <- grepl("^__label__", embed[, 1])
  label_rows  <- embed[ is_label, , drop = FALSE]
  region_rows <- embed[!is_label, , drop = FALSE]

  if (nrow(label_rows) < 2) {
    stop(sprintf(
      "model.tsv has only %d label row(s) (prefix '__label__'); expected at least 2.",
      nrow(label_rows)
    ))
  }
  if (nrow(region_rows) < n_clusters) {
    stop(sprintf("Only %d region rows in model.tsv; need at least n_clusters = %d.",
                 nrow(region_rows), n_clusters))
  }

  region_names <- region_rows[, 1]
  region_emb   <- as.matrix(region_rows[, -1])
  storage.mode(region_emb) <- "double"

  label_names <- label_rows[, 1]
  label_emb   <- as.matrix(label_rows[, -1])
  storage.mode(label_emb) <- "double"

  # Locate the two label rows by name (after stripping the "__label__" prefix).
  stripped      <- gsub("^__label__", "", label_names)
  renin_idx     <- which(stripped == renin_label)
  non_renin_idx <- which(stripped == non_renin_label)
  if (length(renin_idx) != 1 || length(non_renin_idx) != 1) {
    stop(sprintf(
      "Could not locate both '%s' and '%s' label rows; found label rows: %s",
      renin_label, non_renin_label, paste(stripped, collapse = ", ")
    ))
  }
  renin_vec     <- label_emb[renin_idx,     ]
  non_renin_vec <- label_emb[non_renin_idx, ]

  # Per-region cosine similarity to each label vector — reused by both
  # methods below and for the optional margin filter.
  cs_to <- function(M, v) {
    n <- sqrt(rowSums(M * M)) * sqrt(sum(v * v))
    as.numeric((M %*% v) / pmax(n, 1e-12))
  }
  cos_renin     <- cs_to(region_emb, renin_vec)
  cos_non_renin <- cs_to(region_emb, non_renin_vec)

  if (method == "cosine_nearest") {
    # Per-region direct assignment: each region goes to its nearest label
    # vector by cosine. This is the cosine bisector partition — by
    # construction, it agrees pointwise with the cosine-metric UMAP and
    # with the cos_renin/cos_non_renin axes used in 8f. No k-means, no
    # global optimization that can disagree with local structure.
    #
    # Only meaningful for n_clusters = 2 (one cluster per label). For
    # n_clusters > 2, fall back to spherical k-means.
    if (n_clusters != 2) {
      warning("method='cosine_nearest' assumes n_clusters=2; falling back to spherical_kmeans.")
      method <- "spherical_kmeans"
    } else {
      title_vec <- ifelse(cos_renin > cos_non_renin,
                          output_renin, output_non_renin)

      # Optional margin filter: regions where the two cos-sims are too
      # close get marked output_other instead of forced into one cluster.
      # margin_threshold = 0 keeps the original "nearest" behavior.
      if (margin_threshold > 0) {
        ambig <- abs(cos_renin - cos_non_renin) < margin_threshold
        title_vec[ambig] <- output_other
        message(sprintf(
          "  margin_threshold=%.3f marked %s/%s regions as '%s'",
          margin_threshold,
          format(sum(ambig), big.mark = ","),
          format(length(ambig), big.mark = ","),
          output_other
        ))
      }
    }
  }

  if (method == "spherical_kmeans") {
    # Normalize each region to unit norm before k-means. Makes Euclidean
    # distance equivalent to cosine distance on the sphere, so the
    # k-means boundary aligns with cosine — usually agrees with UMAP
    # except for a few boundary regions.
    region_emb_norm <- region_emb / sqrt(rowSums(region_emb * region_emb))
    region_emb_norm[!is.finite(region_emb_norm)] <- 0

    # init_at_labels: seed the first two centroids at the unit-normalized
    # label vectors. K-means then refines from there — first iteration's
    # assignment is exactly "nearest label by cosine", subsequent
    # iterations let centroids drift toward data density without losing
    # the label-anchored partition. Disables nstart (no random restarts).
    set.seed(seed)
    if (init_at_labels) {
      norm_unit <- function(v) v / max(sqrt(sum(v * v)), 1e-12)
      init_centers <- rbind(norm_unit(renin_vec), norm_unit(non_renin_vec))
      if (n_clusters > 2) {
        # Pick the remaining (n_clusters - 2) initial centers from the
        # region embeddings, far from the two label-anchored seeds.
        d_to_seeds <- pmin(
          sqrt(rowSums(sweep(region_emb_norm, 2, init_centers[1, ], `-`)^2)),
          sqrt(rowSums(sweep(region_emb_norm, 2, init_centers[2, ], `-`)^2))
        )
        extra_idx <- order(d_to_seeds, decreasing = TRUE)[seq_len(n_clusters - 2)]
        init_centers <- rbind(init_centers, region_emb_norm[extra_idx, , drop = FALSE])
      }
      km <- kmeans(region_emb_norm, centers = init_centers, iter.max = 50)
    } else {
      km <- kmeans(region_emb_norm, centers = n_clusters,
                   nstart = 10, iter.max = 50)
    }

    centroids  <- km$centers
    sims_renin <- apply(centroids, 1, function(c) {
      s <- sqrt(sum(c*c)) * sqrt(sum(renin_vec * renin_vec))
      if (s == 0) 0 else sum(c * renin_vec) / s
    })
    sims_nr <- apply(centroids, 1, function(c) {
      s <- sqrt(sum(c*c)) * sqrt(sum(non_renin_vec * non_renin_vec))
      if (s == 0) 0 else sum(c * non_renin_vec) / s
    })

    if (label_method == "cluster_cosine_nearest") {
      # Every cluster commits to Renin or Non-renin based on its centroid's
      # cosine sim. With n_clusters > 2, multiple clusters can carry the
      # same label — no "other" bucket. Useful when you want a strictly
      # binary partition but also want k-means to find finer sub-structure.
      cluster_to_label <- ifelse(sims_renin >= sims_nr,
                                  output_renin, output_non_renin)
    } else {
      # Original "one per label": assign the closest cluster to each
      # label, rest as "other". Surfaces sub-structure that doesn't
      # align with either label.
      cluster_to_label <- rep(output_other, nrow(centroids))
      cluster_to_label[which.max(sims_renin)] <- output_renin
      for (idx in order(sims_nr, decreasing = TRUE)) {
        if (cluster_to_label[idx] == output_other) {
          cluster_to_label[idx] <- output_non_renin
          break
        }
      }
    }
    title_vec <- cluster_to_label[km$cluster]
  }

  if (method == "hdbscan" || method == "umap_hdbscan") {
    # Shared pipeline for both density-based methods. Both run the same
    # pre-clustering steps and differ ONLY in where HDBSCAN gets its input:
    #
    #   (1) normalize HD embeddings
    #   (2) subsample hdb_max_n ("y") regions
    #   (3) PCA-fit on the y subsample → hdb_pca_dim ("x") dims; project
    #       full 423k regions and label vectors through the same rotation
    #   (4) fit UMAP on the y × x PCA subsample → 2D coords + model
    #   (5) HDBSCAN runs on either:
    #         - "umap_hdbscan": the 2D UMAP coords (visible-blob clustering)
    #         - "hdbscan":      the xD PCA subsample (original-geometry)
    #   (6) label each cluster by centroid cosine sim in HD space
    #   (7) propagate to all 423k regions via k-NN majority vote in PCA space
    #
    # Diagnostic intent: at the same (x, y) the saved UMAP layout is
    # IDENTICAL across the two methods, so any topology/coloring diff is
    # attributable purely to the HDBSCAN input space. Sweep x to test the
    # PCA-reduction effect; sweep y to test the UMAP-training-set effect.
    pkgs <- c("uwot", "dbscan", "RANN")
    missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
    if (length(missing) > 0) {
      stop(sprintf("hdbscan/umap_hdbscan need: %s. install.packages('%s', lib = .libPaths()[1])",
                   paste(missing, collapse = ", "),
                   paste(missing, collapse = "', '")))
    }

    # ---- (1) normalize ---------------------------------------------------
    region_emb_norm <- region_emb / sqrt(rowSums(region_emb * region_emb))
    region_emb_norm[!is.finite(region_emb_norm)] <- 0
    n_total <- nrow(region_emb_norm)

    label_emb_norm <- label_emb / sqrt(rowSums(label_emb * label_emb))
    label_emb_norm[!is.finite(label_emb_norm)] <- 0

    # ---- (2) subsample y regions ----------------------------------------
    set.seed(seed)
    if (n_total > hdb_max_n) {
      sub_idx <- sort(sample.int(n_total, hdb_max_n))
      message(sprintf("  Subsampling %s of %s regions (hdb_max_n=%d)",
                      format(hdb_max_n, big.mark = ","),
                      format(n_total,   big.mark = ","), hdb_max_n))
    } else {
      sub_idx <- seq_len(n_total)
    }
    emb_sub <- region_emb_norm[sub_idx, , drop = FALSE]   # HD subsample

    # ---- (3) PCA on y-subsample, project full 423k via rotation ---------
    # PCA fit on the subsample (fast, numerically very close to a full-data
    # fit at y ≥ 20k); both branches share one rotation so the comparison
    # isolates the HDBSCAN input space. Set hdb_pca_dim to ncol(region_emb)
    # (e.g. 50) to skip PCA entirely — used as the "no PCA reduction"
    # configuration in the x-sweep.
    if (!is.null(hdb_pca_dim) && hdb_pca_dim < ncol(emb_sub)) {
      message(sprintf("  PCA: %d → %d dims (fit on %s subsample, project full data + labels)",
                      ncol(emb_sub), hdb_pca_dim,
                      format(nrow(emb_sub), big.mark = ",")))
      pca <- prcomp(emb_sub, center = TRUE, scale. = FALSE, rank. = hdb_pca_dim)
      emb_sub_pca <- pca$x
      region_emb_pca_all <- scale(region_emb_norm, center = pca$center,
                                  scale = FALSE) %*%
                            pca$rotation[, seq_len(hdb_pca_dim), drop = FALSE]
      label_emb_pca <- scale(label_emb_norm, center = pca$center,
                             scale = FALSE) %*%
                       pca$rotation[, seq_len(hdb_pca_dim), drop = FALSE]
    } else {
      message(sprintf("  PCA: skipped (hdb_pca_dim=%s ≥ input dim=%d)",
                      ifelse(is.null(hdb_pca_dim), "NULL",
                             as.character(hdb_pca_dim)),
                      ncol(emb_sub)))
      pca <- NULL
      emb_sub_pca <- emb_sub
      region_emb_pca_all <- region_emb_norm
      label_emb_pca <- label_emb_norm
    }

    # ---- (4) UMAP on y × x PCA subsample --------------------------------
    # ret_model = TRUE so plot_starspace_clusters can reuse this exact
    # layout (saved as attributes below) and so label vectors can be
    # projected in. Identical across both methods at the same (x, y).
    message(sprintf("Running UMAP on %s × %d to 2D (cosine metric)...",
                    format(nrow(emb_sub_pca), big.mark = ","),
                    ncol(emb_sub_pca)))
    umap_model  <- uwot::umap(emb_sub_pca,
                              n_neighbors = 15, min_dist = 0.1,
                              metric = "cosine", n_components = 2,
                              verbose = FALSE, ret_model = TRUE)
    umap_coords <- umap_model$embedding
    label_umap_coords <- uwot::umap_transform(label_emb_pca, umap_model)

    # ---- (5) HDBSCAN — the ONE step that differs across methods ---------
    if (method == "umap_hdbscan") {
      if (umap_subcluster == "kmeans") {
        message(sprintf("Running k-means (k=%d) on 2D UMAP coords...", n_clusters))
        set.seed(seed)
        km <- kmeans(umap_coords, centers = n_clusters,
                     nstart = 10, iter.max = 50)
        sub_cluster_ids <- km$cluster
        K <- n_clusters
        message(sprintf("  Forced %d cluster(s)", K))
      } else {
        message(sprintf("Running HDBSCAN on 2D UMAP coords (minPts=%d)...", minPts))
        hdb <- dbscan::hdbscan(umap_coords, minPts = minPts)
        sub_cluster_ids <- hdb$cluster
        K <- max(sub_cluster_ids)
        if (K < 1) stop("HDBSCAN on UMAP coords found 0 clusters — lower minPts.")
        message(sprintf("  Found %d UMAP cluster(s); %s subsample points are noise",
                        K, format(sum(sub_cluster_ids == 0), big.mark = ",")))
      }
    } else {
      # method == "hdbscan": run HDBSCAN on the xD PCA subsample directly.
      message(sprintf("Running HDBSCAN on %d-D PCA subsample (minPts=%d)...",
                      ncol(emb_sub_pca), minPts))
      hdb <- dbscan::hdbscan(emb_sub_pca, minPts = minPts)
      sub_cluster_ids <- hdb$cluster
      K <- max(sub_cluster_ids)
      if (K < 1) stop("HDBSCAN found 0 clusters — try lowering minPts.")
      message(sprintf("  Found %d cluster(s); %s subsample regions are noise",
                      K, format(sum(sub_cluster_ids == 0), big.mark = ",")))
    }

    # ---- (6) centroids in HD space, label by cosine to label vectors ----
    # Centroids computed in original normalized (HD) space so cosine sim
    # to renin_vec / non_renin_vec is meaningful (PCA-projected centroids
    # would distort the cosine geometry).
    centroids_hd <- vapply(seq_len(K), function(k) {
      colMeans(emb_sub[sub_cluster_ids == k, , drop = FALSE])
    }, numeric(ncol(emb_sub)))
    centroids_hd <- t(centroids_hd)

    centroid_norms <- sqrt(rowSums(centroids_hd * centroids_hd))
    sims_renin <- as.numeric(centroids_hd %*% renin_vec /
      pmax(centroid_norms * sqrt(sum(renin_vec * renin_vec)), 1e-12))
    sims_nr <- as.numeric(centroids_hd %*% non_renin_vec /
      pmax(centroid_norms * sqrt(sum(non_renin_vec * non_renin_vec)), 1e-12))

    cos_diff <- sims_renin - sims_nr
    cluster_to_label <- ifelse(
      cos_diff >  margin_threshold, output_renin,
      ifelse(cos_diff < -margin_threshold, output_non_renin, output_other)
    )

    message(sprintf("  Cluster -> label (subsample n): %s",
      paste(sprintf("c%d=%s (n=%s, cos_r=%.2f, cos_nr=%.2f, diff=%.2f)",
                    seq_len(K), cluster_to_label,
                    format(tabulate(sub_cluster_ids, K), big.mark = ","),
                    sims_renin, sims_nr, cos_diff),
            collapse = "; ")))

    # ---- (7) propagate to all 423k via k-NN majority vote in PCA space --
    propagation_k <- 5
    message(sprintf("Propagating to %s regions via %d-NN majority vote in %d-D PCA...",
                    format(n_total, big.mark = ","),
                    propagation_k, ncol(emb_sub_pca)))
    nn <- RANN::nn2(emb_sub_pca, query = region_emb_pca_all, k = propagation_k)
    neighbor_clusters <- matrix(sub_cluster_ids[nn$nn.idx],
                                 nrow = n_total, ncol = propagation_k)
    cluster_ids <- apply(neighbor_clusters, 1, function(row) {
      tab <- tabulate(row, nbins = K)
      which.max(tab)
    })

    # Noise (nearest subsample neighbor was HDBSCAN-noise) → fall back to
    # per-region cosine to the two label vectors.
    noise_mask <- cluster_ids == 0L
    title_vec <- character(length(cluster_ids))
    if (any(!noise_mask)) {
      title_vec[!noise_mask] <- cluster_to_label[cluster_ids[!noise_mask]]
    }
    if (any(noise_mask)) {
      title_vec[noise_mask] <- ifelse(
        cos_renin[noise_mask] > cos_non_renin[noise_mask],
        output_renin, output_non_renin
      )
    }

    full_counts <- table(title_vec)
    message(sprintf("  Full-data counts after propagation: %s (total = %s)",
                    paste(sprintf("%s=%s", names(full_counts),
                                  format(as.numeric(full_counts), big.mark = ",")),
                          collapse = ", "),
                    format(length(title_vec), big.mark = ",")))
  }

  # ---- post-hoc cosine override -----------------------------------------
  # Two corrections, applied in order:
  #
  #   1. Margin band → "other": regions where |cos_renin - cos_non_renin|
  #      is below `margin_threshold` are too ambiguous to commit to either
  #      label. They become a third group, output_other. Set
  #      margin_threshold = 0 to disable (gives the original binary output).
  #
  #   2. Flip: of the remaining "confident" regions, any whose currently-
  #      assigned label disagrees with their cosine-nearest label get
  #      flipped. Catches streak-tip mislabels from centroid propagation.
  #
  # Together these turn HDBSCAN/k-means output into a Renin / Non-renin /
  # other partition that follows per-region cosine geometry, while keeping
  # the original method's clustering as the "structural prior" outside the
  # ambiguous band.
  if (cosine_override) {
    cos_diff <- cos_renin - cos_non_renin

    n_other_added <- 0
    if (margin_threshold > 0) {
      ambig <- abs(cos_diff) < margin_threshold & title_vec != output_other
      n_other_added <- sum(ambig)
      title_vec[ambig] <- output_other
    }

    flip_to_nr <- title_vec == output_renin     & cos_diff < 0
    flip_to_r  <- title_vec == output_non_renin & cos_diff > 0
    n_flip <- sum(flip_to_nr) + sum(flip_to_r)
    title_vec[flip_to_nr] <- output_non_renin
    title_vec[flip_to_r]  <- output_renin

    message(sprintf(
      "  Post-hoc cosine override: margin=%.3f -> %s regions to '%s'; flipped %s %s<->%s pairs (%.2f%% of regions affected total)",
      margin_threshold,
      format(n_other_added, big.mark = ","), output_other,
      format(n_flip,        big.mark = ","), output_renin, output_non_renin,
      100 * (n_other_added + n_flip) / length(title_vec)
    ))

    # ---- spatial smoothing: grow "other" via k-NN propagation ----------
    # Each region currently labeled Renin/Non-renin checks its k nearest
    # neighbors (in PCA-reduced embedding space). If at least
    # spatial_smooth_thresh of them are "other", this region flips to
    # "other" too. Iterates until convergence (no new flips) or max iters.
    # Captures the full geometric sub-cluster, including interior regions
    # whose individual cos_diff exceeds margin_threshold.
    if (spatial_smooth && margin_threshold > 0 &&
        sum(title_vec == output_other) > 10) {
      has_rann <- requireNamespace("RANN", quietly = TRUE)
      has_fnn  <- requireNamespace("FNN",  quietly = TRUE)
      if (!has_rann && !has_fnn) {
        warning("spatial_smooth requires 'RANN' or 'FNN' — skipping (install.packages('RANN'))")
      } else {
        # PCA-reduce for fast k-NN. Three strategies, fastest first:
        #   (a) Reuse HDBSCAN's PCA basis (free — already computed in the
        #       hdbscan block) and project full data via matrix multiply.
        #   (b) irlba::prcomp_irlba — partial SVD, O(n·m·k), 5-20x faster
        #       than prcomp at this scale.
        #   (c) prcomp fallback — full SVD, slow but available everywhere.
        smooth_dim <- min(15, ncol(region_emb_norm))
        if (exists("region_emb_pca_all") && exists("pca") && !is.null(pca)) {
          # Reuse the PCA basis from the shared hdbscan/umap_hdbscan pipeline
          # — the full-data projection is already in memory.
          message("  Spatial smoothing: reusing pre-computed PCA basis (no recompute)")
          coords <- region_emb_pca_all[, seq_len(min(smooth_dim,
                                                     ncol(region_emb_pca_all))),
                                       drop = FALSE]
        } else if (requireNamespace("irlba", quietly = TRUE)) {
          message("  Spatial smoothing: running irlba PCA (partial SVD)...")
          smooth_pca <- irlba::prcomp_irlba(region_emb_norm, n = smooth_dim,
                                             center = TRUE, scale. = FALSE)
          coords <- smooth_pca$x
        } else {
          message("  Spatial smoothing: running prcomp PCA (slow — install 'irlba' for 5-20x speedup)...")
          smooth_pca <- prcomp(region_emb_norm, center = TRUE, scale. = FALSE,
                               rank. = smooth_dim)
          coords <- smooth_pca$x
        }
        knn_backend <- if (has_rann) "RANN" else "FNN"
        message(sprintf(
          "  Spatial smoothing: starting at %s 'other' regions, propagating via k-NN (k=%d, thresh=%.2f, backend=%s)...",
          format(sum(title_vec == output_other), big.mark = ","),
          spatial_smooth_k, spatial_smooth_thresh, knn_backend
        ))
        n_thresh <- ceiling(spatial_smooth_k * spatial_smooth_thresh)

        # RANN::nn2 is 5-10x faster than FNN at this scale. Returns same
        # nn.idx matrix. Asking for k+1 because RANN includes the query
        # point itself as the first neighbor — strip column 1.
        if (has_rann) {
          nn_raw <- RANN::nn2(coords, k = spatial_smooth_k + 1)
          nn <- list(nn.index = nn_raw$nn.idx[, -1, drop = FALSE])
        } else {
          nn <- FNN::get.knn(coords, k = spatial_smooth_k)
        }

        for (iter in seq_len(spatial_smooth_max_iter)) {
          neighbor_titles <- matrix(title_vec[nn$nn.index],
                                     nrow = nrow(coords),
                                     ncol = spatial_smooth_k)
          n_other_nbrs <- rowSums(neighbor_titles == output_other)
          new_other <- n_other_nbrs >= n_thresh & title_vec != output_other
          n_added <- sum(new_other)
          if (n_added == 0) {
            message(sprintf("    iter %d: converged (no new flips)", iter))
            break
          }
          title_vec[new_other] <- output_other
          message(sprintf("    iter %d: +%s regions (total 'other' = %s)",
                          iter, format(n_added, big.mark = ","),
                          format(sum(title_vec == output_other),
                                 big.mark = ",")))
        }
      }
    }
  }

  # Parse "chr_start_end" -> (id, start, end). Anchor on the last two
  # numeric blocks so chromosomes with underscores (e.g. "chr1_random") parse.
  parsed <- str_match(region_names, "^(.+)_([0-9]+)_([0-9]+)$")
  ok <- !is.na(parsed[, 1])
  if (any(!ok)) {
    warning(sprintf(
      "%d region name(s) did not match 'chr_start_end' pattern; dropped.",
      sum(!ok)
    ))
  }

  result <- data.frame(
    title = title_vec[ok],
    id    = parsed[ok, 2],
    start = as.integer(parsed[ok, 3]),
    end   = as.integer(parsed[ok, 4]),
    stringsAsFactors = FALSE
  )

  # Attach UMAP coords (subsample only) + label coords as attributes so
  # plot_starspace_clusters can reuse them — avoids a second UMAP run with
  # different layout. Both density-based methods now share the same UMAP
  # pre-training step, so the saved layout is identical across them at the
  # same (hdb_pca_dim, hdb_max_n).
  if (method %in% c("hdbscan", "umap_hdbscan") && exists("umap_coords")) {
    attr(result, "umap_coords")       <- umap_coords
    attr(result, "umap_region_keys")  <- region_names[sub_idx]
    attr(result, "umap_label_coords") <- label_umap_coords
    attr(result, "umap_label_names")  <- gsub("^__label__", "", label_names)
  }

  result
}
```

## Generate `starspace_regions`

```{r}
starspace_regions <- cluster_starspace_regions(
  starspace_model_path,
  method = "hdbscan",
  minPts = 50,
  hdb_max_n   = 20000,
  hdb_pca_dim = 5,
  cosine_override         = TRUE,
  margin_threshold        = 0.1,
  spatial_smooth          = FALSE
  )

write.csv(
  starspace_regions,
  file.path(bed_dir, "cluster_regions.csv"),
  row.names = FALSE
)


```

## Visualize the embedding colored by cluster

2D projection of the region embedding, colored by the Renin / Non-renin
label that `cluster_starspace_regions()` assigned. The two trained label
vectors (`__label__Renin`, `__label__Non_Renin`) are drawn as triangles
so you can see whether each cluster sits near its named label and
roughly how far apart the two labels ended up in embedding space.

Defaults to **PCA** (instant, linear). Set `method = "umap"` for a
non-linear UMAP projection — slower (≈30–120s on 100k regions) and needs
`uwot` installed in your R library.

```{r, fig.width = 8, fig.height = 6}
plot_starspace_clusters <- function(model_path, starspace_regions,
                                    method          = c("umap", "pca", "cos_axes"),
                                    max_regions     = 5000,
                                    n_neighbors     = 15,
                                    min_dist        = 0.1,
                                    color_regions   = TRUE,
                                    alpha_renin     = 0.5,
                                    alpha_non_renin = 0.5,
                                    alpha_other     = 0.5,
                                    seed            = 42,
                                    force_refit_umap = FALSE) {
  # force_refit_umap = TRUE → ignore any saved UMAP attrs and re-fit a
  # fresh UMAP on `max_regions` regions sampled from the full set in raw
  # high-D space (cosine metric, no PCA). Use this to reproduce the
  # pre-refactor plot topology, where the clustering UMAP and the plot
  # UMAP were independent.
  method <- match.arg(method)

  # ---- Check for saved UMAP coords from cluster_starspace_regions -------
  # If present and method = "umap", reuse them so the plot shows exactly
  # the layout that was clustered on. Avoids the UMAP-vs-UMAP mismatch
  # (different subsample, different layout) that causes the appearance
  # of "wrong color" dots at cluster tips in the plot.
  saved_umap_coords  <- attr(starspace_regions, "umap_coords")
  saved_region_keys  <- attr(starspace_regions, "umap_region_keys")
  saved_label_coords <- attr(starspace_regions, "umap_label_coords")
  saved_label_names  <- attr(starspace_regions, "umap_label_names")
  use_saved <- method == "umap" &&
               !force_refit_umap &&
               !is.null(saved_umap_coords) &&
               !is.null(saved_region_keys) &&
               !is.null(saved_label_coords) &&
               !is.null(saved_label_names)

  if (use_saved) {
    message("Reusing UMAP coords from cluster_starspace_regions (no UMAP re-fit)")

    set.seed(seed)
    n_regions <- nrow(saved_umap_coords)
    idx <- if (n_regions > max_regions) sort(sample.int(n_regions, max_regions))
           else seq_len(n_regions)

    region_coords    <- saved_umap_coords[idx, , drop = FALSE]
    region_names_sub <- saved_region_keys[idx]

    ss_key          <- paste(starspace_regions$id, starspace_regions$start,
                             starspace_regions$end, sep = "_")
    region_to_label <- setNames(starspace_regions$title, ss_key)
    region_labels   <- region_to_label[region_names_sub]

    label_coords <- saved_label_coords
    label_names  <- saved_label_names
    axis_x       <- "UMAP 1"
    axis_y       <- "UMAP 2"
  } else {
    # ---- Original path: parse model.tsv, subsample, run dim-reduction ----
    # Reparse model.tsv — identify labels by "__label__" prefix, not by position.
    embed <- read.table(model_path, sep = "\t", header = FALSE,
                        comment.char = "", quote = "", stringsAsFactors = FALSE,
                        fill = TRUE)
    embed <- embed[!is.na(embed[, 1]) & embed[, 1] != "", , drop = FALSE]
    is_label    <- grepl("^__label__", embed[, 1])
    label_rows  <- embed[ is_label, , drop = FALSE]
    region_rows <- embed[!is_label, , drop = FALSE]

    region_emb   <- as.matrix(region_rows[, -1]); storage.mode(region_emb) <- "double"
    label_emb    <- as.matrix(label_rows[,  -1]); storage.mode(label_emb)  <- "double"
    region_names <- region_rows[, 1]
    label_names  <- gsub("^__label__", "", label_rows[, 1])

    # Subsample to max_regions (matches geniml's create_umap_visualization()
    # default of 5000). UMAP at 5k is ~10s; at 50k it can take minutes.
    set.seed(seed)
    n_regions <- nrow(region_emb)
    idx <- if (n_regions > max_regions) sort(sample.int(n_regions, max_regions))
           else seq_len(n_regions)
    region_emb_sub   <- region_emb[idx, ]
    region_names_sub <- region_names[idx]

    # Per-region cluster label from starspace_regions, joined by "chr_start_end".
    ss_key          <- paste(starspace_regions$id, starspace_regions$start,
                             starspace_regions$end, sep = "_")
    region_to_label <- setNames(starspace_regions$title, ss_key)
    region_labels   <- region_to_label[region_names_sub]

    # Joint dim-reduction so region points and label triangles end up in
    # the same coordinate system.
    combined <- rbind(region_emb_sub, label_emb)
    if (method == "umap") {
      if (!requireNamespace("uwot", quietly = TRUE)) {
        stop("Install 'uwot' for UMAP: install.packages('uwot', lib = .libPaths()[1])")
      }
      message(sprintf("Running UMAP on %s points (n_neighbors=%d, min_dist=%.2f, cosine)...",
                      format(nrow(combined), big.mark = ","), n_neighbors, min_dist))
      coords <- uwot::umap(combined,
                           n_neighbors = min(n_neighbors, nrow(combined) - 1),
                           min_dist    = min_dist,
                           metric      = "cosine",
                           n_components = 2,
                           verbose     = TRUE)
      axis_x <- "UMAP 1"; axis_y <- "UMAP 2"
    } else if (method == "pca") {
      pc     <- prcomp(combined, scale. = FALSE)
      coords <- pc$x[, 1:2]
      axis_x <- "PC1"; axis_y <- "PC2"
    } else if (method == "cos_axes") {
      renin_idx     <- which(label_names == "Renin")
      non_renin_idx <- which(label_names == "Non_Renin")
      if (length(renin_idx) != 1 || length(non_renin_idx) != 1) {
        stop("cos_axes needs both 'Renin' and 'Non_Renin' label rows in model.tsv.")
      }
      cs <- function(M, v) {
        n <- sqrt(rowSums(M * M)) * sqrt(sum(v * v))
        as.numeric((M %*% v) / pmax(n, 1e-12))
      }
      coords <- cbind(
        cs(combined, label_emb[renin_idx,     ]),
        cs(combined, label_emb[non_renin_idx, ])
      )
      axis_x <- "cos sim to Renin label"
      axis_y <- "cos sim to Non-renin label"
    }

    # Split coordinates back out (matches geniml's split: region_coords first,
    # label_coords second).
    region_coords <- coords[seq_along(idx), , drop = FALSE]
    label_coords  <- coords[(length(idx) + 1):nrow(coords), , drop = FALSE]
  }

  region_df <- data.frame(
    x       = region_coords[, 1],
    y       = region_coords[, 2],
    cluster = region_labels,
    stringsAsFactors = FALSE
  )
  label_df <- data.frame(
    x     = label_coords[, 1],
    y     = label_coords[, 2],
    name  = label_names,
    stringsAsFactors = FALSE
  )

  # geniml's label palette: darkorange for Renin, royalblue for Non_Renin.
  label_palette <- vapply(label_df$name, function(n) {
    nl <- tolower(n)
    if (nl == "renin")                              "darkorange"
    else if (grepl("non", nl) && grepl("renin", nl)) "royalblue"
    else                                            "grey40"
  }, character(1))
  names(label_palette) <- label_df$name

  cluster_palette <- c(Renin = "darkorange", `Non-renin` = "royalblue",
                       other = "grey60")

  # Region layers: split by cluster so each color has its own alpha.
  # Setting alpha_renin = 0 hides orange points to reveal blue dots that
  # might be underneath (and vice versa). When color_regions = FALSE, all
  # regions render uniform grey at alpha_other (geniml's style).
  rasterise_if_avail <- function(layer) {
    if (requireNamespace("ggrastr", quietly = TRUE))
      ggrastr::rasterise(layer, dpi = 200) else layer
  }

  region_layers <- if (color_regions) {
    df_r  <- region_df[region_df$cluster == "Renin",     ]
    df_nr <- region_df[region_df$cluster == "Non-renin", ]
    df_o  <- region_df[!region_df$cluster %in% c("Renin", "Non-renin"), ]
    list(
      rasterise_if_avail(
        geom_point(data = df_nr, aes(x, y, colour = cluster),
                   size = 1, alpha = alpha_non_renin)),
      rasterise_if_avail(
        geom_point(data = df_o, aes(x, y, colour = cluster),
                   size = 1, alpha = alpha_other)),
      rasterise_if_avail(
        geom_point(data = df_r, aes(x, y, colour = cluster),
                   size = 1, alpha = alpha_renin))
      # Order: blue first, then other, then orange on top — so when both
      # alphas are non-zero, the minority color (usually orange Renin
      # cluster) is visible above the dense blue background.
    )
  } else {
    list(rasterise_if_avail(
      geom_point(data = region_df, aes(x, y),
                 colour = "lightgrey", size = 1, alpha = alpha_other)))
  }

  ggplot() +
    region_layers +
    geom_point(data = label_df, aes(x, y, fill = name),
               size = 6, shape = 24, colour = "black", stroke = 1.5,
               show.legend = TRUE) +
    geom_text(data = label_df, aes(x, y, label = name),
              hjust = -0.15, vjust = -0.3,
              fontface = "bold", size = 4, colour = "black") +
    scale_colour_manual(values = cluster_palette, name = "k-means cluster") +
    scale_fill_manual(values   = label_palette,   name = "Label embedding") +
    labs(title    = sprintf("StarSpace embedding — %s",
                            switch(method,
                                   pca      = "PCA",
                                   umap     = "UMAP (n_neighbors=15, min_dist=0.1, cosine)",
                                   cos_axes = "cos-sim to label vectors")),
         subtitle = sprintf("n = %s of %s regions",
                            format(length(idx), big.mark = ","),
                            format(n_regions,   big.mark = ",")),
         x        = axis_x,
         y        = axis_y) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          legend.position  = "right")
}
```

```{r}
p_embed_clusters <- plot_starspace_clusters(starspace_model_path,
                                            starspace_regions,
                                            method        = "umap",
                                            max_regions   = 5000,
                                            n_neighbors   = 15,
                                            min_dist      = 0.1,
                                            alpha_renin     = 0.3,
                                            alpha_non_renin = 0.3,
                                            alpha_other     = 0.3,
                                            color_regions = TRUE,
                                            force_refit_umap = TRUE)

ggsave(file.path("plot", "starspace_embedding_clusters_force_refit.svg"),
       p_embed_clusters, width = 10, height = 8, units = "in", dpi = 200)

p_embed_clusters <- plot_starspace_clusters(starspace_model_path,
                                            starspace_regions,
                                            method        = "umap",
                                            max_regions   = 5000,
                                            n_neighbors   = 15,
                                            min_dist      = 0.1,
                                            alpha_renin     = 0.3,
                                            alpha_non_renin = 0.3,
                                            alpha_other     = 0.3,
                                            color_regions = TRUE,
                                            force_refit_umap = FALSE)

ggsave(file.path("plot", "starspace_embedding_clusters.svg"),
       p_embed_clusters, width = 10, height = 8, units = "in", dpi = 200)
```

# 8c. Process DESeq2 differential regions

## Load DESeq2 reports

```{r}
primary_normal <- readGeneric(primary_normal_file, chr = 1, start = 2, end = 3,
                              keep.all.metadata = TRUE, zero.based = FALSE,
                              remove.unusual = FALSE, header = TRUE,
                              skip = 0, sep = ",")
primary_high <- readGeneric(primary_high_file, chr = 1, start = 2, end = 3,
                            keep.all.metadata = TRUE, zero.based = FALSE,
                            remove.unusual = FALSE, header = TRUE,
                            skip = 0, sep = ",")
as_normal <- readGeneric(as_normal_file, chr = 1, start = 2, end = 3,
                         keep.all.metadata = TRUE, zero.based = FALSE,
                         remove.unusual = FALSE, header = TRUE,
                         skip = 0, sep = ",")
absFoldChange = 2
fdr = 0.01
```

## Load universe

```{r}
universe <- read.table(universe_file, sep = "\t", header = FALSE,
                       stringsAsFactors = FALSE)
universe_keys <- paste(universe[, 1], universe[, 2], universe[, 3], sep = "_")
length(universe_keys)
```

## Filter into up / down sets

```{r}
primary_normal <- subset(primary_normal, pvalue < fdr)
primary_normal_up <- subset(primary_normal, log2FoldChange >=  absFoldChange)
primary_normal_down <- subset(primary_normal, log2FoldChange <= -absFoldChange)

primary_high <- subset(primary_high, padj < fdr)
primary_high_up <- subset(primary_high, log2FoldChange >=  absFoldChange)
primary_high_down <- subset(primary_high, log2FoldChange <= -absFoldChange)

as_normal <- subset(as_normal, padj < fdr)
as_normal_up <- subset(as_normal, log2FoldChange >=  absFoldChange)
as_normal_down <- subset(as_normal, log2FoldChange <= -absFoldChange)
```

## Build cross-contrast intersection sets

```{r}
lst_up <- list(
  paste(as.data.frame(as_normal_up)$seqnames,        as.data.frame(as_normal_up)$start,        as.data.frame(as_normal_up)$end,        sep = "_"),
  paste(as.data.frame(primary_normal_up)$seqnames,   as.data.frame(primary_normal_up)$start,   as.data.frame(primary_normal_up)$end,   sep = "_"),
  paste(as.data.frame(primary_high_up)$seqnames,     as.data.frame(primary_high_up)$start,     as.data.frame(primary_high_up)$end,     sep = "_")
)

lst_down <- list(
  paste(as.data.frame(as_normal_down)$seqnames,      as.data.frame(as_normal_down)$start,      as.data.frame(as_normal_down)$end,      sep = "_"),
  paste(as.data.frame(primary_normal_down)$seqnames, as.data.frame(primary_normal_down)$start, as.data.frame(primary_normal_down)$end, sep = "_"),
  paste(as.data.frame(primary_high_down)$seqnames,   as.data.frame(primary_high_down)$start,   as.data.frame(primary_high_down)$end,   sep = "_")
)

counts_up   <- table(unlist(lst_up))
counts_down <- table(unlist(lst_down))

inc <- names(counts_up)  [counts_up   >= 2]
dec <- names(counts_down)[counts_down >= 2]

```

# 8d. Overlap analysis between DESeq2 and StarSpace

## Build region categories

StarSpace clusters (`r`, `nr`) + DESeq2 calls (`inc`, `dec`, `nd`). `nd`
is just `universe - inc - dec`, drawn from the BED loaded in 8c.

```{r}
ss_up   <- subset(starspace_regions, title == "Renin")
ss_down <- subset(starspace_regions, title == "Non-renin")

r  <- paste(ss_up$id,   ss_up$start,   ss_up$end,   sep = "_")
nr <- paste(ss_down$id, ss_down$start, ss_down$end, sep = "_")

nd <- setdiff(c(r, nr), c(inc, dec)) 
```

## UpSet plot

```{r}
listInput_all <- list(
  Increase         = inc,
  Decrease         = dec,
  Non_differential = nd,
  Renin            = r,
  Non_renin        = nr
)

svg(file.path(plot_dir, "ss_deseq_upset.svg"), width = 10, height = 6, onefile = FALSE)
upset(fromList(listInput_all),
      mainbar.y.label = "overlapped regions",
      sets.x.label = "regions per group",
      sets = c("Increase", "Decrease", "Non_differential", "Renin", "Non_renin"),
      keep.order = TRUE, order.by = "freq",
      show.numbers = "yes",
      set_size.show = TRUE,
      set_size.scale_max = max(lengths(listInput_all)) * 1.2,
      text.scale = c(1.5,1.5,1.5,1.5,1.75,1.5))
dev.off()

sets <- list(
  `Increase:Renin`             = intersect(inc, r),
  `Increase:Non_renin`         = intersect(inc, nr),
  `Decrease:Renin`             = intersect(dec, r),
  `Decrease:Non_renin`         = intersect(dec, nr),
  `Non_differential:Renin`     = intersect(nd, r),
  `Non_differential:Non_renin` = intersect(nd, nr)
)
lengths(sets)
```

# 8e. Write intersection BED files

```{r}
write_set_bed <- function(peaks, out_path) {
  df <- data.frame(peaks = peaks)
  df <- df %>%
    extract(peaks, c("seqname", "start", "end"), regex = "(.*)_([^_]+)_([^_]+)$")
  df <- df[order(df$seqname, df$start), ]
  write.table(df, out_path, row.names = FALSE, quote = FALSE,
              sep = "\t", col.names = FALSE)
  GenomicRanges::makeGRangesFromDataFrame(df)
}

nd_r  <- write_set_bed(sets$`Non_differential:Renin`,    file.path(bed_dir, "nd_r.bed"))
nd_nr <- write_set_bed(sets$`Non_differential:Non_renin`,file.path(bed_dir, "nd_nr.bed"))
inc_r <- write_set_bed(sets$`Increase:Renin`,            file.path(bed_dir, "inc_r.bed"))
inc_nr<- write_set_bed(sets$`Increase:Non_renin`,        file.path(bed_dir, "inc_nr.bed"))
dec_r <- write_set_bed(sets$`Decrease:Renin`,            file.path(bed_dir, "dec_r.bed"))
dec_nr<- write_set_bed(sets$`Decrease:Non_renin`,        file.path(bed_dir, "dec_nr.bed"))
```

# 8f. StarSpace embedding vs DESeq2 differential accessibility

Two diagnostics tying the trained StarSpace embedding back to the DESeq2
calls. This section runs before the tornado plots (8g–8i) so the
embedding ↔ DE relationship is validated first, and the cos-sim values
are available to inform tornado feature-set selection downstream.

1.  **Scatter** — cosine similarity of each region embedding to the
    trained `Renin` label embedding vs DESeq2 log2 fold change, faceted
    by contrast. If the StarSpace model learned "renin-like
    accessibility," cos-sim should track log2FC; per-facet Spearman ρ
    quantifies that.
2.  **Binned line plot** — % of regions called differential by DESeq2 as
    a function of cos-sim, in quantile bins. A clean knee identifies a
    cos-sim threshold for defining renin-specific regions without
    relying on the DESeq2 FDR/log2FC cutoff. Pooled-across-contrasts and
    per-contrast views are both produced.

Cost: cos-sim is one matrix–vector product (O(n·d), \<1 s for n=10⁶,
d=50). The bottleneck is re-reading the three DESeq2 CSVs (8c
subset-overwrote the unfiltered objects). The merged region-level table
is built once and reused for both panels.

## Read embeddings, compute cos-sim to the Renin label

```{r}
read_starspace_embeddings <- function(model_path) {
  embed <- read.table(model_path, sep = "\t", header = FALSE, skip = 1,
                      comment.char = "", quote = "", stringsAsFactors = FALSE,
                      fill = TRUE)
  embed <- embed[!is.na(embed[, 1]) & embed[, 1] != "", , drop = FALSE]
  n_rows      <- nrow(embed)
  region_rows <- embed[1:(n_rows - 2), , drop = FALSE]
  label_rows  <- embed[(n_rows - 1):n_rows, , drop = FALSE]

  region_emb <- as.matrix(region_rows[, -1]); storage.mode(region_emb) <- "double"
  label_emb  <- as.matrix(label_rows[,  -1]); storage.mode(label_emb)  <- "double"

  list(
    region_names = region_rows[, 1],
    region_emb   = region_emb,
    label_emb    = label_emb,
    label_names  = gsub("^__label__", "", label_rows[, 1])
  )
}

emb           <- read_starspace_embeddings(starspace_model_path)
renin_idx     <- which(emb$label_names == "Renin")
non_renin_idx <- which(emb$label_names == "Non_Renin")
stopifnot(length(renin_idx) == 1, length(non_renin_idx) == 1)
renin_vec     <- emb$label_emb[renin_idx, ]
non_renin_vec <- emb$label_emb[non_renin_idx, ]

# Cosine sim of every region row to each label vector. One BLAS call each.
region_norms   <- sqrt(rowSums(emb$region_emb * emb$region_emb))
renin_norm     <- sqrt(sum(renin_vec * renin_vec))
non_renin_norm <- sqrt(sum(non_renin_vec * non_renin_vec))
cos_renin      <- as.numeric((emb$region_emb %*% renin_vec) /
                             (region_norms * renin_norm))
cos_non_renin  <- as.numeric((emb$region_emb %*% non_renin_vec) /
                             (region_norms * non_renin_norm))

emb_df <- data.frame(
  region        = emb$region_names,
  cos_renin     = cos_renin,
  cos_non_renin = cos_non_renin,
  stringsAsFactors = FALSE
)
```

## Join with the 8b k-means cluster labels

```{r}
cluster_csv <- file.path(bed_dir, "cluster_regions_2.csv")
if (file.exists(cluster_csv)) {
  starspace_regions <- read.csv(cluster_csv, stringsAsFactors = FALSE)
  message("Loaded cached starspace_regions from ", cluster_csv,
          " (", nrow(starspace_regions), " regions)")
} else {
  if (!exists("cluster_starspace_regions")) {
    stop("cluster_regions.csv not found and cluster_starspace_regions() ",
         "is not defined. Run 8b first (or source its function chunk).")
  }
  starspace_regions <- cluster_starspace_regions(starspace_model_path)
  write.csv(starspace_regions, cluster_csv, row.names = FALSE)
  message("Recomputed and cached starspace_regions to ", cluster_csv,
          " (", nrow(starspace_regions), " regions)")
}

ss_dt <- data.frame(
  region  = paste(starspace_regions$id, starspace_regions$start,
                  starspace_regions$end, sep = "_"),
  cluster = starspace_regions$title,
  stringsAsFactors = FALSE
)
emb_df <- merge(emb_df, ss_dt, by = "region", all.x = TRUE)

```

## Re-read DESeq2 reports (unfiltered) and build the long-format table

```{r}
# Use data.table::fread if available (~5× faster on 100k+ rows); else read.csv.
read_deseq <- function(path) {
  df <- if (requireNamespace("data.table", quietly = TRUE)) {
    as.data.frame(data.table::fread(path))
  } else {
    read.csv(path, stringsAsFactors = FALSE)
  }
  df$region <- paste(df$seqnames, df$start, df$end, sep = "_")
  df
}

dq_pn <- read_deseq(primary_normal_file); dq_pn$contrast <- "primary_normal"
dq_ph <- read_deseq(primary_high_file);   dq_ph$contrast <- "primary_high"
dq_an <- read_deseq(as_normal_file);      dq_an$contrast <- "as_normal"

# Per-contrast differential flag — same filter logic as 8c (primary_normal
# uses raw pvalue; the other two use padj).
dq_pn$is_diff <- !is.na(dq_pn$pvalue) & dq_pn$pvalue < fdr &
                 abs(dq_pn$log2FoldChange) >= absFoldChange
dq_ph$is_diff <- !is.na(dq_ph$padj)   & dq_ph$padj   < fdr &
                 abs(dq_ph$log2FoldChange) >= absFoldChange
dq_an$is_diff <- !is.na(dq_an$padj)   & dq_an$padj   < fdr &
                 abs(dq_an$log2FoldChange) >= absFoldChange

keep_cols   <- c("region", "log2FoldChange", "contrast", "is_diff")
all_dq      <- rbind(dq_pn[, keep_cols], dq_ph[, keep_cols], dq_an[, keep_cols])
scatter_df  <- merge(all_dq, emb_df, by = "region")
scatter_df$contrast <- factor(scatter_df$contrast,
                              levels = c("primary_normal", "primary_high", "as_normal"))

# Per-contrast Spearman correlation, computed on ALL rows (not the
# plotting subsample). Annotated onto each facet below.
cor_df <- scatter_df %>%
  group_by(contrast) %>%
  summarise(rho = cor(log2FoldChange, cos_renin,
                      method = "spearman", use = "complete.obs"),
            n   = n(),
            .groups = "drop") %>%
  mutate(label = sprintf("rho = %.2f  (n = %s)",
                         rho, format(n, big.mark = ",")))
print(cor_df)
```

## (1) Scatter: cosine similarity to Renin label vs log2 fold change

```{r, fig.width = 12, fig.height = 5}
plot_cos_vs_log2fc <- function(df, non_diff_cap = 20000, seed = 42) {
  has_ggrastr <- requireNamespace("ggrastr", quietly = TRUE)
  if (!has_ggrastr) {
    set.seed(seed)
    diff_df <- df[df$is_diff, ]
    nd_df   <- df[!df$is_diff, ]
    nd_df   <- nd_df[
      if (nrow(nd_df) > non_diff_cap) sort(sample.int(nrow(nd_df), non_diff_cap))
      else seq_len(nrow(nd_df)),
    ]
    df <- rbind(diff_df, nd_df)
    message(sprintf(
      "ggrastr not installed; plotting %s differential + %s non-differential points",
      format(nrow(diff_df), big.mark = ","),
      format(nrow(nd_df),   big.mark = ",")
    ))
  }

  # Status: Increase / Decrease / Non-differential. Levels ordered so grey
  # draws first, colored on top.
  df$status <- factor(
    ifelse(df$is_diff & df$log2FoldChange > 0, "Increase",
    ifelse(df$is_diff & df$log2FoldChange < 0, "Decrease", "Non-differential")),
    levels = c("Non-differential", "Decrease", "Increase")
  )
  df <- df[order(df$status), ]   # draw grey first, colored on top

  # Pivot to long: each region appears twice — once with cos_sim = cos_renin
  # (top row), once with cos_sim = cos_non_renin (bottom row). Keep only the
  # columns the plot needs to bound memory at ~half scatter_df's footprint.
  df_long <- bind_rows(
    df %>% transmute(contrast, log2FoldChange, status,
                     cos_sim = cos_renin,     label_target = "Renin"),
    df %>% transmute(contrast, log2FoldChange, status,
                     cos_sim = cos_non_renin, label_target = "Non-renin")
  ) %>%
    mutate(label_target = factor(label_target, levels = c("Renin", "Non-renin")))

  pt <- if (has_ggrastr) {
    ggrastr::rasterise(geom_point(alpha = 0.3, size = 0.3), dpi = 200)
  } else {
    geom_point(alpha = 0.3, size = 0.3)
  }

  ggplot(df_long, aes(log2FoldChange, cos_sim, colour = status)) +
    pt +
    facet_grid(label_target ~ contrast, switch = "y") +
    geom_hline(yintercept = 0, colour = "black", linewidth = 0.3) +
    geom_vline(xintercept = c(-absFoldChange, absFoldChange),
               colour = "black", linewidth = 0.3, linetype = "dashed") +
    geom_text(
      data = cor_df, aes(x = -Inf, y = Inf, label = label),
      hjust = -0.05, vjust = 1.3, size = 3.2, inherit.aes = FALSE
    ) +
    scale_colour_manual(
      values = c(Increase           = "#FF7F0E",   # tab:orange
                 Decrease           = "#1F77B4",   # tab:blue
                 `Non-differential` = "grey70"),
      breaks = c("Increase", "Decrease", "Non-differential"),
      name   = NULL
    ) +
    labs(x = "DESeq2 log2 fold change",
         y = "cosine similarity to label") +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          legend.position  = "bottom",
          strip.placement  = "outside")
}

p_scatter <- plot_cos_vs_log2fc(scatter_df)

ggsave(file.path(plot_dir, "starspace_cos_vs_log2fc.svg"),
       p_scatter, width = 12, height = 9, units = "in", dpi = 200)

```

## (2) P(differential) vs cosine similarity — binned line plot

Replaces the cluster stacked bar. The k-means split is a hard binary cut
that throws away the continuous cos-sim information, which is precisely
what we want to inspect here. Binning regions by cos-sim quantile and
plotting the per-bin DE rate directly shows whether embedding distance
and DESeq2 calls co-vary smoothly (linear slope → no natural cutoff,
keep using DESeq2 FDR/log2FC) or whether there's a knee where DE rate
drops sharply (use cos-sim as the threshold instead).

Bin edges come from quantiles of the full pooled cos-sim distribution,
so the same edges apply to every panel — the averaged and faceted views
are directly comparable.

```{r}
plot_pde_vs_cos <- function(df,
                            facet_by_contrast = TRUE,
                            n_bins = 20) {
  # Bin edges from quantiles of the full cos_renin distribution. Applying
  # the same edges to every contrast (and to the pooled "avg" view) keeps
  # panels directly comparable.
  brks <- quantile(df$cos_renin,
                   probs = seq(0, 1, length.out = n_bins + 1),
                   na.rm = TRUE)
  df$cos_bin <- cut(df$cos_renin, breaks = brks, include.lowest = TRUE)

  group_cols <- if (facet_by_contrast) c("contrast", "cos_bin") else "cos_bin"

  bin_summary <- df %>%
    group_by(across(all_of(group_cols))) %>%
    summarise(
      n       = n(),
      cos_mid = median(cos_renin),
      pct_inc = 100 * mean(is_diff & log2FoldChange > 0),
      pct_dec = 100 * mean(is_diff & log2FoldChange < 0),
      .groups = "drop"
    ) %>%
    tidyr::pivot_longer(c(pct_inc, pct_dec),
                        names_to  = "direction",
                        values_to = "pct") %>%
    mutate(direction = recode(direction,
                              pct_inc = "Increase",
                              pct_dec = "Decrease"))

  p <- ggplot(bin_summary, aes(cos_mid, pct, colour = direction)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    scale_colour_manual(values = c(Increase = "#FF7F0E", Decrease = "#1F77B4"),
                        name = "DESeq2 direction") +
    labs(x = "cosine similarity to Renin label (bin median)",
         y = "% of regions called differential",
         caption = sprintf(
           "Quantile-binned, %d bins, n_rows = %s%s",
           n_bins, format(nrow(df), big.mark = ","),
           if (facet_by_contrast) " (per-contrast)" else " (pooled across contrasts)"
         )) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())

  if (facet_by_contrast) {
    p <- p + facet_wrap(~ contrast)
  }
  p
}
```

### (2a) Pooled across contrasts — dual-axis with knee detection

Each direction gets its own cos-sim axis:

- **Increase (orange)** is binned by `cos_renin` and read against the
  **bottom** x-axis (cos sim to the Renin label).
- **Decrease (blue)** is binned by `cos_non_renin` and read against the
  **top** x-axis (cos sim to the Non-renin label).

Both axes share the same numerical scale (cos-sim is bounded in [-1, 1]
either way), they just measure distance to different label vectors. The
two binnings are independent — each line is in its own native axis
space.

Dashed vertical lines mark the **knee** for each direction: the smallest
cos-sim at which the per-bin DE rate first crosses `de_threshold_pct`.
Those two thresholds are returned in the result and reused in 8g–8i to
subset the tornado feature sets.

```{r, fig.width = 7, fig.height = 4.5}
plot_pde_vs_cos_dual_axis <- function(df,
                                      n_bins           = 20,
                                      knee_method      = c("kneedle", "abs"),
                                      de_threshold_pct = 50) {
  knee_method <- match.arg(knee_method)
  # ---- bin Increase by cos_renin ----
  brks_r <- quantile(df$cos_renin,
                     probs = seq(0, 1, length.out = n_bins + 1),
                     na.rm = TRUE)
  bin_inc <- df %>%
    mutate(cos_bin = cut(cos_renin, breaks = brks_r, include.lowest = TRUE)) %>%
    group_by(cos_bin) %>%
    summarise(cos_mid = median(cos_renin),
              pct     = 100 * mean(is_diff & log2FoldChange > 0),
              n       = n(),
              .groups = "drop") %>%
    mutate(direction = "Increase")

  # ---- bin Decrease by cos_non_renin ----
  brks_nr <- quantile(df$cos_non_renin,
                      probs = seq(0, 1, length.out = n_bins + 1),
                      na.rm = TRUE)
  bin_dec <- df %>%
    mutate(cos_bin = cut(cos_non_renin, breaks = brks_nr, include.lowest = TRUE)) %>%
    group_by(cos_bin) %>%
    summarise(cos_mid = median(cos_non_renin),
              pct     = 100 * mean(is_diff & log2FoldChange < 0),
              n       = n(),
              .groups = "drop") %>%
    mutate(direction = "Decrease")

  binned <- bind_rows(
    bin_inc %>% select(cos_mid, pct, n, direction),
    bin_dec %>% select(cos_mid, pct, n, direction)
  ) %>%
    mutate(direction = factor(direction, levels = c("Increase", "Decrease")))

  # ---- knee detection ----
  # Two methods, controlled by `knee_method`:
  #
  # "kneedle" (default) — finds the point where the curve STARTS rising
  #   exponentially. For a convex-increasing curve (flat tail → steep
  #   rise), the data sits BELOW the chord from first to last bin. The
  #   knee is the point of maximum downward deviation from that chord
  #   (after normalising x and y to [0, 1]). This catches the inflection
  #   where %DE first lifts off background, not where it has already
  #   climbed past some absolute level.
  #
  # "abs" — fallback: smallest cos_mid where per-bin pct first crosses
  #   `de_threshold_pct`. Use this if the curve is roughly linear and
  #   Kneedle is picking a degenerate point.
  detect_knee <- function(bd) {
    bd <- bd %>% arrange(cos_mid) %>% dplyr::filter(!is.na(pct))
    if (nrow(bd) < 3) return(NA_real_)

    if (knee_method == "kneedle") {
      x_range <- range(bd$cos_mid)
      y_range <- range(bd$pct)
      if (diff(x_range) == 0 || diff(y_range) == 0) return(NA_real_)
      xn <- (bd$cos_mid - x_range[1]) / diff(x_range)
      yn <- (bd$pct    - y_range[1]) / diff(y_range)
      # Max downward deviation from chord (0,0) → (1,1).
      return(bd$cos_mid[which.min(yn - xn)])
    }

    # "abs" method
    above <- which(bd$pct >= de_threshold_pct)
    if (length(above) == 0) return(NA_real_)
    bd$cos_mid[min(above)]
  }
  knee_r  <- detect_knee(bin_inc)
  knee_nr <- detect_knee(bin_dec)

  knees_df <- data.frame(
    direction = factor(c("Increase", "Decrease"),
                       levels = c("Increase", "Decrease")),
    knee      = c(knee_r, knee_nr),
    label     = c(sprintf("Renin knee = %.3f", knee_r),
                  sprintf("Non-renin knee = %.3f", knee_nr))
  )

  p <- ggplot(binned, aes(cos_mid, pct, colour = direction)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.5) +
    geom_vline(data = knees_df,
               aes(xintercept = knee, colour = direction),
               linetype = "dashed", linewidth = 0.5, na.rm = TRUE,
               show.legend = FALSE) +
    geom_text(data = knees_df,
              aes(x = knee, y = Inf, label = label, colour = direction),
              hjust = -0.05, vjust = 1.5, size = 3, na.rm = TRUE,
              show.legend = FALSE) +
    scale_x_continuous(
      name     = "cos sim to Renin label  (Increase line, bottom axis)",
      sec.axis = dup_axis(
        name = "cos sim to Non-renin label  (Decrease line, top axis)"
      )
    ) +
    scale_colour_manual(values = c(Increase = "#FF7F0E", Decrease = "#1F77B4"),
                        name = "DESeq2 direction") +
    labs(y = "% of regions called differential",
         caption = sprintf(
           "Quantile-binned (%d bins each). Dashed verticals = knee where %%DE starts rising exponentially (%s%s). n_rows = %s (pooled across contrasts)",
           n_bins,
           if (knee_method == "kneedle") "Kneedle: max downward deviation from chord"
           else sprintf("abs: first bin with %%DE >= %d%%", de_threshold_pct),
           "",
           format(nrow(df), big.mark = ",")
         )) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())

  list(plot           = p,
       knee_renin     = knee_r,
       knee_non_renin = knee_nr,
       bin_inc        = bin_inc,
       bin_dec        = bin_dec)
}

# Default: kneedle inflection detection — picks the cos-sim where the
# %DE curve starts rising exponentially, not where it has already climbed
# past a fixed level. To override and use a hard %DE threshold instead
# (e.g. for a high-purity cutoff), pass knee_method = "abs" and tune
# de_threshold_pct.
pde_avg_res <- plot_pde_vs_cos_dual_axis(
  scatter_df,
  n_bins      = 20,
  knee_method = "kneedle"
  # , de_threshold_pct = 50   # only used when knee_method = "abs"
)

# Export the two knee thresholds for downstream tornado subsetting (8g–8i).
cos_renin_threshold     <- pde_avg_res$knee_renin
cos_non_renin_threshold <- pde_avg_res$knee_non_renin
message(sprintf(
  "Knee thresholds: cos_renin >= %.3f, cos_non_renin >= %.3f",
  cos_renin_threshold, cos_non_renin_threshold
))

ggsave(file.path(plot_dir, "starspace_pde_vs_cos_avg.svg"),
       pde_avg_res$plot, width = 7, height = 4.5, units = "in")
```

### (2a-bed) Subset regions by knee threshold for tornado feature sets

These two GRanges are the embedding-defined "renin-specific" and
"non-renin-specific" sets, ready to be slotted into the `feats` list in
8g. They use the knee thresholds detected above, so they're tighter (or
looser) than the DESeq2-derived sets exactly to the degree that the
embedding agrees with DESeq2.

**Manual override:** the Non-renin curve doesn't have a clean knee in
this run, so its auto-detected `knee_non_renin` is noisy. Set both
thresholds to the same well-defined value (the clear Renin-side knee at
0.217) so the dec_nr / nd_nr filter on `cos_non_renin` mirrors the inc_r
/ nd_r filter on `cos_renin`. Comment out the override block to fall
back to the auto-detected per-direction knees.

```{r}
# ---- threshold override --------------------------------------------------
# Set to NULL to use the auto-detected knee from pde_avg_res. Set to a
# numeric to override (e.g. when Kneedle returns an unstable point because
# the curve has no clear inflection).
manual_cos_renin_threshold     <- 0.217   # NULL to use pde_avg_res$knee_renin
manual_cos_non_renin_threshold <- 0.217   # NULL to use pde_avg_res$knee_non_renin

cos_renin_threshold <- if (!is.null(manual_cos_renin_threshold))
  manual_cos_renin_threshold else pde_avg_res$knee_renin
cos_non_renin_threshold <- if (!is.null(manual_cos_non_renin_threshold))
  manual_cos_non_renin_threshold else pde_avg_res$knee_non_renin

message(sprintf(
  "Using thresholds: cos_renin >= %.3f (%s), cos_non_renin >= %.3f (%s)",
  cos_renin_threshold,
  if (!is.null(manual_cos_renin_threshold)) "manual" else "auto/kneedle",
  cos_non_renin_threshold,
  if (!is.null(manual_cos_non_renin_threshold)) "manual" else "auto/kneedle"
))
# --------------------------------------------------------------------------

names_to_granges <- function(names) {
  parsed <- str_match(names, "^(.+)_([0-9]+)_([0-9]+)$")
  ok <- !is.na(parsed[, 1])
  GenomicRanges::makeGRangesFromDataFrame(data.frame(
    seqnames = parsed[ok, 2],
    start    = as.integer(parsed[ok, 3]),
    end      = as.integer(parsed[ok, 4])
  ))
}

renin_emb_set <- emb_df %>%
  dplyr::filter(!is.na(cos_renin), cos_renin >= cos_renin_threshold) %>%
  dplyr::pull(region) %>%
  names_to_granges()

non_renin_emb_set <- emb_df %>%
  dplyr::filter(!is.na(cos_non_renin), cos_non_renin >= cos_non_renin_threshold) %>%
  dplyr::pull(region) %>%
  names_to_granges()

message(sprintf(
  "Embedding-defined sets: renin_emb_set = %s regions, non_renin_emb_set = %s regions",
  format(length(renin_emb_set),     big.mark = ","),
  format(length(non_renin_emb_set), big.mark = ",")
))

# Write to disk so 8g can reload them without re-running 8f.
writeLines(
  paste(GenomicRanges::seqnames(renin_emb_set),
        GenomicRanges::start(renin_emb_set),
        GenomicRanges::end(renin_emb_set), sep = "\t"),
  file.path(bed_dir, "renin_emb_set.bed")
)
writeLines(
  paste(GenomicRanges::seqnames(non_renin_emb_set),
        GenomicRanges::start(non_renin_emb_set),
        GenomicRanges::end(non_renin_emb_set), sep = "\t"),
  file.path(bed_dir, "non_renin_emb_set.bed")
)
```

### (2b) Faceted by contrast — primary_normal / primary_high / as_normal

```{r, fig.width = 12, fig.height = 4}
p_pde_facet <- plot_pde_vs_cos(scatter_df, facet_by_contrast = TRUE)

ggsave(file.path(plot_dir, "starspace_pde_vs_cos_by_contrast.svg"),
       p_pde_facet, width = 12, height = 4, units = "in")
```

# 8g. Tornado signal profiles

## Reload intersection BEDs (for resumable execution)

```{r}
load_bed <- function(path) {
  readGeneric(path, chr = 1, start = 2, end = 3, strand = NULL,
              meta.cols = NULL, keep.all.metadata = FALSE, zero.based = FALSE,
              remove.unusual = FALSE, header = FALSE, skip = 0, sep = "\t")
}

dec_nr <- load_bed(file.path(bed_dir,"dec_nr.bed"))
dec_r  <- load_bed(file.path(bed_dir,"dec_r.bed"))
inc_nr <- load_bed(file.path(bed_dir,"inc_nr.bed"))
inc_r  <- load_bed(file.path(bed_dir,"inc_r.bed"))
nd_nr  <- load_bed(file.path(bed_dir,"nd_nr.bed"))
nd_r   <- load_bed(file.path(bed_dir,"nd_r.bed"))
```

## Subset feature sets by top-N cos-sim ranking

Replace the knee-based threshold with a fixed top-N count per subset.
Each of the four DESeq×StarSpace subsets stays — only the *filter rule*
changes:

- `inc_r`  → keep top N by `cos_renin`     (renin-confidence rank)
- `nd_r`   → keep top N by `cos_renin`
- `dec_nr` → keep top N by `cos_non_renin` (non-renin-confidence rank)
- `nd_nr`  → keep top N by `cos_non_renin`

Why: top-N gives **deterministic, balanced row counts per subset**
(every subset is at most `top_n`), so the heatmaps stop being dominated
by the huge `nd_*` blocks. The DESeq2 categorization is preserved.

Needs `emb_df` from 8f (cos_renin / cos_non_renin per region).

```{r}
top_n <- 1000

if (!exists("emb_df") || is.null(emb_df$cos_renin) ||
    is.null(emb_df$cos_non_renin)) {
  stop("emb_df with cos_renin / cos_non_renin not in scope. ",
       "Run 8f first or rebuild it here.")
}

# Key lookups: "chr_start_end" -> per-region cosine sim
key_cos_renin     <- setNames(emb_df$cos_renin,     emb_df$region)
key_cos_non_renin <- setNames(emb_df$cos_non_renin, emb_df$region)

gr_to_key <- function(gr) {
  paste(GenomicRanges::seqnames(gr),
        GenomicRanges::start(gr),
        GenomicRanges::end(gr), sep = "_")
}

# Filter a GRanges to its top-N regions by an arbitrary cos-sim lookup.
filter_top_by_cos <- function(gr, cos_lookup, top_n) {
  keys     <- gr_to_key(gr)
  cos_vals <- cos_lookup[keys]
  # Drop NAs (regions missing from emb_df). Then take top-N by cos.
  finite_ix <- which(is.finite(cos_vals))
  ord <- finite_ix[order(cos_vals[finite_ix], decreasing = TRUE)]
  gr[head(ord, top_n)]
}

sizes_before <- vapply(
  list(inc_r = inc_r, dec_nr = dec_nr, nd_r = nd_r, nd_nr = nd_nr),
  length, integer(1)
)

# Renin-side filter (inc_r, nd_r) -> top by cos_renin
inc_r <- filter_top_by_cos(inc_r, key_cos_renin,     top_n)
nd_r  <- filter_top_by_cos(nd_r,  key_cos_renin,     top_n)
# Non-renin-side filter (dec_nr, nd_nr) -> top by cos_non_renin
dec_nr <- filter_top_by_cos(dec_nr, key_cos_non_renin, top_n)
nd_nr  <- filter_top_by_cos(nd_nr,  key_cos_non_renin, top_n)

sizes_after <- c(
  inc_r  = length(inc_r),
  dec_nr = length(dec_nr),
  nd_r   = length(nd_r),
  nd_nr  = length(nd_nr)
)

message(sprintf("Top-%d subset:", top_n))
for (nm in names(sizes_before)) {
  message(sprintf("  %-7s %s -> %s",
                  paste0(nm, ":"),
                  format(sizes_before[[nm]], big.mark = ","),
                  format(sizes_after[[nm]],  big.mark = ",")))
}
```

## Subset feature sets by cos-sim knee from 8f

Restricts each feature set to regions on the "correct" side of the knee
detected in 8f section (2a):

- `_r` sets (Renin cluster) → keep only regions with
  `cos_renin >= cos_renin_threshold`
- `_nr` sets (Non-renin cluster) → keep only regions with
  `cos_non_renin >= cos_non_renin_threshold`

Why: regions that pass the embedding cluster's k-means filter but still
sit below the knee are the embedding's lukewarm calls — they don't show
the renin/non-renin signal strongly. Dropping them sharpens the tornado
average, drops peak memory by 5–20× depending on where the knee lands,
and is the same operation visually summarized by the dashed knee lines
in 8f (2a).

The two embedding BED files were written at the end of 8f by the
(2a-bed) chunk. Falls back to the unfiltered sets with a clear message
if those BEDs aren't on disk yet (e.g. you're running 8g before 8f).

```{r}
emb_bed_paths <- c(
  renin     = "bed/renin_emb_set.bed",
  non_renin = "bed/non_renin_emb_set.bed"
)

if (all(file.exists(emb_bed_paths))) {
  renin_emb_set     <- load_bed(emb_bed_paths[["renin"]])
  non_renin_emb_set <- load_bed(emb_bed_paths[["non_renin"]])

  # "chr_start_end" key matching — matches the convention used in 8c/8e.
  gr_keys <- function(gr) {
    paste(GenomicRanges::seqnames(gr),
          GenomicRanges::start(gr),
          GenomicRanges::end(gr), sep = "_")
  }
  renin_emb_keys     <- gr_keys(renin_emb_set)
  non_renin_emb_keys <- gr_keys(non_renin_emb_set)

  subset_gr <- function(gr, keys) gr[gr_keys(gr) %in% keys]

  sizes_before <- vapply(
    list(inc_r = inc_r, dec_nr = dec_nr, nd_r = nd_r, nd_nr = nd_nr),
    length, integer(1)
  )

  inc_r  <- subset_gr(inc_r,  renin_emb_keys)
  nd_r   <- subset_gr(nd_r,   renin_emb_keys)
  dec_nr <- subset_gr(dec_nr, non_renin_emb_keys)
  nd_nr  <- subset_gr(nd_nr,  non_renin_emb_keys)

  sizes_after <- c(
    inc_r  = length(inc_r),
    dec_nr = length(dec_nr),
    nd_r   = length(nd_r),
    nd_nr  = length(nd_nr)
  )

  message("Subset feature sets by cos-sim knee:")
  for (nm in names(sizes_before)) {
    message(sprintf(
      "  %-7s %s -> %s  (%.1f%% kept)",
      paste0(nm, ":"),
      format(sizes_before[[nm]], big.mark = ","),
      format(sizes_after[[nm]],  big.mark = ","),
      100 * sizes_after[[nm]] / max(sizes_before[[nm]], 1)
    ))
  }
} else {
  missing <- emb_bed_paths[!file.exists(emb_bed_paths)]
  message(sprintf(
    "Embedding BED(s) missing — tornado feature sets stay un-subset.\n  Missing: %s\n  Working dir: %s\n  Run 8f section (2a-bed) and stage the BEDs to this dir to enable subsetting.",
    paste(missing, collapse = ", "),
    getwd()
  ))
}
```

## Load bigWig signal files

Two bigwig configurations — one for each tornado plot we'll produce:

1.  **`bigwigs_avg`**: four group averages (Non-renin, WT, Recruited,
    Tumoral).
2.  **`bigwigs_samples`**: hand-picked individual replicate bigwigs.

`EnrichedHeatmap`'s `normalizeToMatrix` takes a GRanges of signal
coverage, which we import on the fly via `rtracklayer::import.bw()`.

PEPATAC output layout (renin + ENCODE pipelines):
`<base>/<sample_id>/aligned_mm10/<sample_id>_smooth.bw`

```{r}
# PEPATAC output roots. Renin and ENCODE samples live in sibling pipeline
# dirs under the same brickyard processed root.
renin_bw_base  <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/renin_ATAC/results_pipeline"
encode_bw_base <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/encode_ATAC/results_pipeline"
susztak_bw_base <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/susztak_ATAC/results_pipeline"

pepatac_bw <- function(sample_id, base = renin_bw_base) {
  file.path(base, sample_id, "aligned_mm10",
            paste0(sample_id, "_smooth.bw"))
}

# (1) Group-averaged bigwigs — one per sample group, for the high-level
# tornado. The PEPATAC `data/` sibling dir holds NR/WT/PH/PN/TN_average.bw.
# Order here = display order in legends and as columns in the heatmap.
avg_bw_dir <- "/project/shefflab/brickyard/results_pipeline/gomez_atac/results_pipeline_PEPATAC/data"

bigwigs_avg <- c(
  WT = file.path(avg_bw_dir, "PN_average.bw"),   # wild-type renin cells (subset of Renin_Normal)
  PH = file.path(avg_bw_dir, "PH_average.bw"),   # primary high (recruited)
  TN = file.path(avg_bw_dir, "TN_average.bw"),   # tumoral (As4.1)
  NR = file.path(avg_bw_dir, "NR_average.bw")    # non-renin (ENCODE controls)
)
stopifnot(all(file.exists(bigwigs_avg)))

# (2) Selected per-replicate bigwigs — for the per-sample tornado.
# Display name on the left, PEPATAC sample id (= dir name) on the right.
bigwigs_samples <- c(
  "WT_1"                  = pepatac_bw("WT_1"),
  "WT_2"                  = pepatac_bw("WT_2"),
  "Ren1c-YFP_1"           = pepatac_bw("Ren1c-YFP_1"),
  "Ren1c-YFP_2"           = pepatac_bw("Ren1c-YFP_2"),
  "Ren1c-YFP_Low_Na+Cap_1"= pepatac_bw("Ren1c-YFP_Low_Na+Cap_1"),
  "Ren1c-YFP_Low_Na+Cap_2"= pepatac_bw("Ren1c-YFP_Low_Na+Cap_2"),
  "Ren1c--_Ren1c-YFP_1"   = pepatac_bw("Ren1c--_Ren1c-YFP_1"),
  "Ren1c--_Ren1c-YFP_2"   = pepatac_bw("Ren1c--_Ren1c-YFP_2"),
  "Recruited_1"           = pepatac_bw("Recruited_1"),
  "Recruited_2"           = pepatac_bw("Recruited_2"),
  "Recruited_GFP"         = pepatac_bw("Recruited_GFP"),
  "ASKO_1"                = pepatac_bw("ASKO_1"),
  "ASKO_2"                = pepatac_bw("ASKO_2"),
  "As4.1_2"               = pepatac_bw("As4.1_2"),
  "As4.1_H89_Vehicle_1"   = pepatac_bw("As4.1_H89_Vehicle_1"),
  "As4.1_H89_Vehicle_2"   = pepatac_bw("As4.1_H89_Vehicle_2"),
  "As4.1_Jq1_-_1"         = pepatac_bw("As4.1_Jq1_-_1"),
  "As4.1_Jq1_-_2"         = pepatac_bw("As4.1_Jq1_-_2"),
  "As4.1_A485_Vehicle_2"  = pepatac_bw("As4.1_A485_Vehicle_2"),
  # ENCODE / Susztak control samples — separate PEPATAC base
  "P0_1"                  = pepatac_bw("P0_mouse_kidney_bulk_ATAC-seq_sample_1", susztak_bw_base),
  "P0_2"                  = pepatac_bw("P0_mouse_kidney_bulk_ATAC-seq_sample_2", susztak_bw_base),
  "3W_1"                  = pepatac_bw("3-week_mouse_kidney_bulk_ATAC-seq_sample_1", susztak_bw_base),
  "3W_2"                  = pepatac_bw("3-week_mouse_kidney_bulk_ATAC-seq_sample_2", susztak_bw_base),
  "8W_1"                  = pepatac_bw("8-week_mouse_kidney_bulk_ATAC-seq_sample_1", susztak_bw_base),
  "8W_2"                  = pepatac_bw("8-week_mouse_kidney_bulk_ATAC-seq_sample_2", susztak_bw_base),
  "ENCSR023QZX_1"         = pepatac_bw("ENCSR023QZX_1", encode_bw_base),
  "ENCSR023QZX_2"         = pepatac_bw("ENCSR023QZX_2", encode_bw_base),
  "ENCSR389CLN_1"         = pepatac_bw("ENCSR389CLN_1", encode_bw_base),
  "ENCSR389CLN_2"         = pepatac_bw("ENCSR389CLN_2", encode_bw_base),
  "ENCSR732OTZ_1"         = pepatac_bw("ENCSR732OTZ_1", encode_bw_base),
  "ENCSR732OTZ_2"         = pepatac_bw("ENCSR732OTZ_2", encode_bw_base),
  "ENCSR758IRM_1"         = pepatac_bw("ENCSR758IRM_1", encode_bw_base),
  "ENCSR758IRM_2"         = pepatac_bw("ENCSR758IRM_2", encode_bw_base)
)
```

## Define feature set + signal-binning helper

For each (feature, bigwig) pair, resize features to their midpoint, bin
the bigwig coverage into `w`-bp windows spanning ±`extend` around each
midpoint, then average across features per position. Output is a long-
format df with (`feature_set, sample_name, position, mean, sd, n`).

Defaults: **extend = 1000 bp, w = 20 bp** → 100 bins per row (50 per side).

```{r}
flatten_features_eh <- function(features, bigwigs, extend = 1000, w = 20,
                                verbose = TRUE) {
  ts  <- function() format(Sys.time(), "%H:%M:%S")
  log <- function(...) if (verbose) message(sprintf("[%s] %s", ts(), paste0(...)))

  pos_vec <- c(
    seq(-extend + w / 2, -w / 2, by = w),
    seq( w / 2,  extend - w / 2, by = w)
  )
  out <- vector("list", length(features) * length(bigwigs))
  k <- 0L

  n_s <- length(bigwigs)
  n_f <- length(features)
  log(sprintf(">> flatten_features_eh: %d sample(s) × %d feature set(s) = %d calls; ",
              n_s, n_f, n_s * n_f),
      sprintf("extend=%d w=%d", extend, w))

  total_t0 <- Sys.time()
  for (si in seq_along(bigwigs)) {
    sample_name <- names(bigwigs)[si]
    bw_path     <- bigwigs[[sample_name]]

    log(sprintf("[%d/%d] importing %s ...", si, n_s, basename(bw_path)))
    t0 <- Sys.time()
    signal <- rtracklayer::import(bw_path)
    log(sprintf("    imported in %.1fs; %s ranges, %.0f MB in memory",
                as.numeric(Sys.time() - t0, units = "secs"),
                format(length(signal), big.mark = ","),
                object.size(signal) / 1e6))

    for (fi in seq_along(features)) {
      feature_set <- names(features)[fi]
      feat_gr     <- features[[feature_set]]
      if (length(feat_gr) == 0) {
        log(sprintf("    [%d/%d] %s × %s: skipped (0 features)",
                    fi, n_f, sample_name, feature_set))
        next
      }
      log(sprintf("    [%d/%d] %s × %s: %s features, normalizeToMatrix ...",
                  fi, n_f, sample_name, feature_set,
                  format(length(feat_gr), big.mark = ",")))
      t1 <- Sys.time()
      mid <- GenomicRanges::resize(feat_gr, width = 1, fix = "center")
      mat <- EnrichedHeatmap::normalizeToMatrix(
        signal, target = mid,
        value_column = "score",
        extend = extend, mean_mode = "w0", w = w
      )
      log(sprintf("        done in %.1fs; mat dim %d × %d",
                  as.numeric(Sys.time() - t1, units = "secs"),
                  nrow(mat), ncol(mat)))

      k <- k + 1L
      out[[k]] <- data.frame(
        feature_set = feature_set,
        sample_name = sample_name,
        position    = pos_vec,
        mean        = colMeans(mat, na.rm = TRUE),
        sd          = matrixStats::colSds(as.matrix(mat), na.rm = TRUE),
        n           = nrow(mat),
        stringsAsFactors = FALSE
      )
    }
    rm(signal); gc(verbose = FALSE)
  }
  log(sprintf(">> total wall time: %.1f min",
              as.numeric(Sys.time() - total_t0, units = "mins")))
  do.call(rbind, out[seq_len(k)])
}

# DESeq2 × StarSpace intersection feature sets for tornado + heatmap:
#   inc_r  = Increase × Renin     (top-N by cos_renin within inc_r)
#   dec_nr = Decrease × Non_renin (top-N by cos_non_renin within dec_nr)
#   nd_r   = Non_diff × Renin     (top-N by cos_renin within nd_r)
#   nd_nr  = Non_diff × Non_renin (top-N by cos_non_renin within nd_nr)
feats <- GRangesList(
  "inc_r"  = inc_r,
  "dec_nr" = dec_nr,
  "nd_r"   = nd_r,
  "nd_nr"  = nd_nr
)
```

# 8h. Tornado line plots

±1000 bp around each region's midpoint, w = 20 bp bins (100 bins total).
Lines are per-bin mean across regions; **ribbons show mean ± SEM**
(standard error of the per-region values at that bin). Color palette is
shared with the heatmap top-annotation lines in 8i.

## Helpers

```{r}
# Shared discrete color palette for the 4 feature groups. Same colors
# in tornado_facets() lines and build_tornado_heatmap_independent()
# top-annotation curves so the two figures cross-reference visually.
feature_group_colors <- c(
  inc_r  = "#F8766D",   # red/pink
  dec_nr = "#7CAE00",   # green
  nd_r   = "#00BFC4",   # cyan
  nd_nr  = "#C77CFF"    # purple
)

# Line + ribbon plot, faceted by sample. `ribbon` controls the band:
#   "sem"  = mean ± SEM (default — narrow at high n; shows mean reliability)
#   "sd"   = mean ± SD (wider — shows region-to-region spread)
#   "ci95" = mean ± 1.96 × SEM
#   "none" = line only
tornado_facets <- function(df, title,
                           y_label      = "mean signal",
                           ribbon       = c("sem", "sd", "ci95", "none"),
                           ribbon_alpha = 0.2,
                           hline_zero   = FALSE,
                           group_colors = feature_group_colors) {
  ribbon <- match.arg(ribbon)

  if (ribbon != "none") {
    df <- df %>%
      dplyr::mutate(
        .half = dplyr::case_when(
          ribbon == "sem"  ~ sd / sqrt(pmax(n, 1)),
          ribbon == "ci95" ~ 1.96 * sd / sqrt(pmax(n, 1)),
          ribbon == "sd"   ~ sd,
          TRUE             ~ 0
        ),
        .ymin = mean - .half,
        .ymax = mean + .half
      )
  }

  p <- ggplot(df, aes(position))
  if (ribbon != "none") {
    p <- p +
      geom_ribbon(aes(ymin = .ymin, ymax = .ymax, fill = feature_set),
                  alpha = ribbon_alpha, colour = NA) +
      scale_fill_manual(values = group_colors, guide = "none")
  }
  p <- p +
    geom_line(aes(y = mean, colour = feature_set), linewidth = 0.5) +
    scale_colour_manual(values = group_colors) +
    facet_wrap(~ sample_name) +
    labs(x = "position relative to region midpoint",
         y = y_label, colour = NULL, title = title) +
    theme_bw(base_size = 11) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          legend.position = "right")
  if (hline_zero) {
    p <- p + geom_hline(yintercept = 0, linetype = "dashed",
                        linewidth = 0.3, colour = "grey50")
  }
  p
}
```

## (1) Group-averaged tornado

```{r}
df_avg_rds <- file.path(plot_dir, "df_avg_1k_w20.rds")
df_avg <- if (file.exists(df_avg_rds)) {
  message("Loading cached df_avg from ", df_avg_rds)
  readRDS(df_avg_rds)
} else {
  message("Computing df_avg (4 bigwig imports + normalizeToMatrix)...")
  out <- flatten_features_eh(feats, bigwigs_avg, extend = 1000, w = 20)
  saveRDS(out, df_avg_rds)
  out
}

p_avg <- tornado_facets(
  df_avg,
  "ATAC signal at region midpoints (group averages, mean ± SEM)",
  ribbon = "sem"
)
ggsave(file.path(plot_dir, "tornado_group_averages_1k_w20.svg"),
       p_avg, width = 12, height = 8, units = "in", dpi = 300)
```

## (2) Per-sample tornado

To keep memory bounded for the 33-sample run, subsample each feature set
to at most 5,000 regions before normalizeToMatrix. The mean signal curve
is essentially indistinguishable from the full set, since each bin still
averages over thousands of regions.

```{r}
subsample_feats <- function(gr_list, max_n = 5000, seed = 42) {
  set.seed(seed)
  GRangesList(lapply(gr_list, function(g) {
    if (length(g) <= max_n) g
    else g[sort(sample.int(length(g), max_n))]
  }))
}
subsample_n <- 5000   # set to NULL to disable

feats_for_samples <- if (is.null(subsample_n)) {
  feats
} else {
  subsample_feats(feats, max_n = subsample_n)
}
vapply(feats_for_samples, length, integer(1))

df_samples_rds <- file.path(plot_dir, "df_samples_1k_w20.rds")
df_samples <- if (file.exists(df_samples_rds)) {
  message("Loading cached df_samples from ", df_samples_rds)
  readRDS(df_samples_rds)
} else {
  message("Computing df_samples (33 bigwig imports + normalizeToMatrix)...")
  out <- flatten_features_eh(feats_for_samples, bigwigs_samples,
                              extend = 1000, w = 20)
  saveRDS(out, df_samples_rds)
  out
}

p_samples <- tornado_facets(
  df_samples,
  "ATAC signal at region midpoints (selected samples, mean ± SEM)",
  ribbon       = "sem",
  ribbon_alpha = 0.15
) +
  theme(legend.text     = element_text(size = 7),
        legend.key.size = unit(0.6, "lines"))
ggsave(file.path(plot_dir, "tornado_per_sample_1k_w20.svg"),
       p_samples, width = 16, height = 12, units = "in", dpi = 300)
```

# 8i. Per-region tornado heatmaps

±1000 bp / w = 20 bp matrices, same feature sets as 8h. Each row is a
region; columns are bin positions. Color = signal (raw) or fraction of
total per-(sample, bin) (sum-to-one normalization).

## Helpers

```{r}
compute_tornado_matrices <- function(features, bigwigs, extend = 1000, w = 20,
                                     verbose = TRUE) {
  ts  <- function() format(Sys.time(), "%H:%M:%S")
  log <- function(...) if (verbose) message(sprintf("[%s] %s", ts(), paste0(...)))

  group_names <- names(features)
  combined <- do.call(c, lapply(group_names, function(fn) {
    g <- features[[fn]]
    if (length(g) == 0) return(g)
    mcols(g) <- S4Vectors::DataFrame(.group = rep(fn, length(g)))
    g
  }))
  group_factor <- factor(combined$.group, levels = group_names)
  target_mid   <- GenomicRanges::resize(combined, width = 1, fix = "center")
  log(sprintf(">> compute_tornado_matrices: %d sample(s) × %d feature group(s) → ",
              length(bigwigs), length(features)),
      sprintf("%s total regions, %d bins", format(length(target_mid), big.mark = ","),
              2 * extend / w))

  mats <- vector("list", length(bigwigs))
  names(mats) <- names(bigwigs)
  for (i in seq_along(bigwigs)) {
    sn <- names(bigwigs)[i]
    log(sprintf("  [%d/%d] %s: import + normalizeToMatrix ...",
                i, length(bigwigs), sn))
    t0 <- Sys.time()
    sig <- rtracklayer::import(bigwigs[[i]])
    mats[[sn]] <- EnrichedHeatmap::normalizeToMatrix(
      sig, target_mid,
      value_column = "score",
      extend = extend, mean_mode = "w0", w = w
    )
    log(sprintf("      done in %.1fs",
                as.numeric(Sys.time() - t0, units = "secs")))
    rm(sig); gc(verbose = FALSE)
  }

  list(
    mats         = mats,
    group_factor = group_factor,
    extend       = extend,
    w            = w
  )
}

# Apply (or skip) sum_to_one across feature_sets, then compute one shared
# color function spanning every (sample, group, bin) cell. Lets multiple
# build_tornado_heatmap_independent() calls (e.g. diff vs nd splits)
# share an exact color scale.
compute_unified_col_fun <- function(tornado_data,
                                    normalize = c("none", "sum_to_one"),
                                    col_pal = c("white", "darkblue"),
                                    cap_quantile = 0.99) {
  normalize    <- match.arg(normalize)
  mats         <- tornado_data$mats
  group_factor <- tornado_data$group_factor

  if (normalize == "sum_to_one") {
    for (sn in names(mats)) {
      mat <- mats[[sn]]
      group_bin_means <- vapply(levels(group_factor), function(g) {
        colMeans(mat[group_factor == g, , drop = FALSE], na.rm = TRUE)
      }, numeric(ncol(mat)))
      bin_totals <- rowSums(group_bin_means, na.rm = TRUE)
      mats[[sn]] <- sweep(mat, 2, pmax(bin_totals, 1e-12), "/")
    }
  }

  cap <- max(vapply(mats,
                    function(m) stats::quantile(m, cap_quantile, na.rm = TRUE),
                    numeric(1)))
  message(sprintf("compute_unified_col_fun: %d sample(s), %d group(s), cap = %.4f",
                  length(mats), length(levels(group_factor)), cap))
  circlize::colorRamp2(c(0, cap), col_pal)
}

# Per-sample INDEPENDENT row sorting: each sample's heatmap is built
# standalone with its own per-group row order, sorted by that sample's
# enriched_score. Returns a NAMED LIST of EnrichedHeatmap objects.
# Render the list with draw_independent_heatmaps() below.
build_tornado_heatmap_independent <- function(tornado_data,
                                              display_groups = NULL,
                                              normalize = c("none", "sum_to_one"),
                                              col_pal = c("white", "darkblue"),
                                              cap_quantile = 0.99,
                                              col_fun = NULL,
                                              use_raster = TRUE,
                                              group_colors = feature_group_colors,
                                              verbose = TRUE) {
  normalize    <- match.arg(normalize)
  mats         <- tornado_data$mats
  group_factor <- tornado_data$group_factor

  if (normalize == "sum_to_one") {
    for (sn in names(mats)) {
      mat <- mats[[sn]]
      group_bin_means <- vapply(levels(group_factor), function(g) {
        colMeans(mat[group_factor == g, , drop = FALSE], na.rm = TRUE)
      }, numeric(ncol(mat)))
      bin_totals <- rowSums(group_bin_means, na.rm = TRUE)
      mats[[sn]] <- sweep(mat, 2, pmax(bin_totals, 1e-12), "/")
    }
  }

  if (!is.null(display_groups)) {
    keep_rows <- group_factor %in% display_groups
    mats <- lapply(mats, function(m) m[keep_rows, , drop = FALSE])
    group_factor <- factor(group_factor[keep_rows], levels = display_groups)
  }

  if (is.null(col_fun)) {
    cap <- max(vapply(mats,
                      function(m) stats::quantile(m, cap_quantile, na.rm = TRUE),
                      numeric(1)))
    col_fun <- circlize::colorRamp2(c(0, cap), col_pal)
  } else {
    cap <- environment(col_fun)$breaks[length(environment(col_fun)$breaks)]
  }
  legend_title <- if (normalize == "sum_to_one") "fraction" else "signal"

  line_cols <- group_colors[levels(group_factor)]
  if (any(is.na(line_cols))) {
    fallback <- scales::hue_pal()(length(levels(group_factor)))
    line_cols[is.na(line_cols)] <- fallback[is.na(line_cols)]
  }

  if (verbose) {
    message(sprintf("  independent sort: %d sample heatmap(s), %d feature groups, color cap = %.4f",
                    length(mats), length(levels(group_factor)), cap))
  }

  heatmaps <- lapply(seq_along(mats), function(i) {
    sn  <- names(mats)[i]
    mat <- mats[[sn]]

    new_order <- unlist(lapply(levels(group_factor), function(g) {
      gi <- which(group_factor == g)
      gi[order(EnrichedHeatmap::enriched_score(mat[gi, , drop = FALSE]),
               decreasing = TRUE)]
    }), use.names = FALSE)
    mat_sorted <- mat[new_order, , drop = FALSE]
    gf_sorted  <- group_factor[new_order]

    # Row-block titles only on the leftmost panel — keeps right panels compact.
    row_title <- if (i == 1L)
                   levels(group_factor)
                 else
                   rep("", length(levels(group_factor)))

    EnrichedHeatmap::EnrichedHeatmap(
      mat_sorted,
      col          = col_fun,
      name         = sn,
      column_title = sn,
      row_split    = gf_sorted,
      row_title    = row_title,
      cluster_rows = FALSE,
      use_raster   = use_raster,
      show_heatmap_legend  = (i == 1L),
      heatmap_legend_param = list(title = legend_title),
      top_annotation = ComplexHeatmap::HeatmapAnnotation(
        enriched = EnrichedHeatmap::anno_enriched(
          gp = grid::gpar(col = line_cols, lty = 1, lwd = 1.2)
        )
      )
    )
  })
  names(heatmaps) <- names(mats)
  heatmaps
}

# Render a list of standalone heatmaps side-by-side in one SVG. Each
# heatmap is drawn in its own grid viewport, so per-sample row orders
# are preserved (no shared alignment from the `+` operator).
draw_independent_heatmaps <- function(heatmaps,
                                       svg_path,
                                       column_title = NULL,
                                       panel_width_in = 2.5,
                                       height_in = 10) {
  n <- length(heatmaps)
  total_width <- n * panel_width_in + 1.5   # reserve space for legend

  svg(svg_path, width = total_width, height = height_in)
  grid::grid.newpage()

  if (!is.null(column_title)) {
    grid::grid.text(column_title, x = 0.5, y = 0.97,
                    gp = grid::gpar(fontsize = 12, fontface = "bold"))
    grid::pushViewport(grid::viewport(y = 0.48, height = 0.92))
  }

  grid::pushViewport(grid::viewport(
    layout = grid::grid.layout(nrow = 1, ncol = n)
  ))

  for (i in seq_along(heatmaps)) {
    grid::pushViewport(grid::viewport(layout.pos.col = i))
    ComplexHeatmap::draw(
      heatmaps[[i]],
      newpage = FALSE,
      show_heatmap_legend = (i == 1L),
      show_annotation_legend = (i == 1L)
    )
    grid::popViewport()
  }
  grid::popViewport()
  if (!is.null(column_title)) grid::popViewport()
  dev.off()
}
```

## (1b) Group-averaged tornado heatmap (4 columns: WT, PH, TN, NR)

```{r}
# Normalization: "sum_to_one" or "none"
heatmap_norm <- "none"

td_avg_rds <- file.path(plot_dir, "td_avg_1k_w20.rds")
td_avg <- if (file.exists(td_avg_rds)) {
  message("Loading cached td_avg from ", td_avg_rds)
  readRDS(td_avg_rds)
} else {
  message("Computing td_avg (4 bigwig imports + normalizeToMatrix)...")
  out <- compute_tornado_matrices(feats, bigwigs_avg, extend = 1000, w = 20)
  saveRDS(out, td_avg_rds)
  out
}

# Unified color scale across BOTH splits (diff and nd) of the avg heatmap.
avg_col_fun <- compute_unified_col_fun(td_avg, normalize = heatmap_norm)

# Split 1: differential (inc_r + dec_nr)
draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_avg, display_groups = c("inc_r", "dec_nr"),
    normalize = heatmap_norm, col_fun = avg_col_fun
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_averages_diff_%s_1k_w20.svg",
                                heatmap_norm)),
  column_title = sprintf("Differential (inc_r, dec_nr) — group averages [%s]",
                          heatmap_norm),
  panel_width_in = 2.5,
  height_in      = 8
)

# Split 2: non-differential (nd_r + nd_nr)
draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_avg, display_groups = c("nd_r", "nd_nr"),
    normalize = heatmap_norm, col_fun = avg_col_fun
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_averages_nd_%s_1k_w20.svg",
                                heatmap_norm)),
  column_title = sprintf("Non-differential (nd_r, nd_nr) — group averages [%s]",
                          heatmap_norm),
  panel_width_in = 2.5,
  height_in      = 8
)
```

## (2b) Per-sample tornado heatmap (33 columns, grouped by WT/PH/TN/NR)

```{r}
td_samples_rds <- file.path(plot_dir, "td_samples_1k_w20.rds")
td_samples <- if (file.exists(td_samples_rds)) {
  message("Loading cached td_samples from ", td_samples_rds)
  readRDS(td_samples_rds)
} else {
  message("Computing td_samples (33 bigwig imports + normalizeToMatrix)...")
  out <- compute_tornado_matrices(feats_for_samples, bigwigs_samples,
                                   extend = 1000, w = 20)
  saveRDS(out, td_samples_rds)
  out
}

samples_col_fun <- compute_unified_col_fun(td_samples, normalize = heatmap_norm)

draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_samples, display_groups = c("inc_r", "dec_nr"),
    normalize = heatmap_norm, col_fun = samples_col_fun
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_per_sample_diff_%s_1k_w20.svg",
                                heatmap_norm)),
  column_title = sprintf("Differential (inc_r, dec_nr) — per-sample [%s]",
                          heatmap_norm),
  panel_width_in = 0.7,
  height_in      = 8
)

draw_independent_heatmaps(
  build_tornado_heatmap_independent(
    td_samples, display_groups = c("nd_r", "nd_nr"),
    normalize = heatmap_norm, col_fun = samples_col_fun
  ),
  svg_path = file.path(plot_dir,
                       sprintf("tornado_heatmap_per_sample_nd_%s_1k_w20.svg",
                                heatmap_norm)),
  column_title = sprintf("Non-differential (nd_r, nd_nr) — per-sample [%s]",
                          heatmap_norm),
  panel_width_in = 0.7,
  height_in      = 8
)
```
# 8j. GO enrichment weighted-rank plot — refined Renin vs Non-renin regions

Same GREAT enrichment + renin-term weighted-rank procedure as
`4_greatGOenrichment.Rmd` sections 4d/4e, but applied to the refined
region classification that combines both DESeq2 and StarSpace evidence:

- **Renin** = `inc_r` — regions that (1) showed increased accessibility
  in renin samples in the DESeq2 analysis AND (2) fell in the StarSpace
  Renin cluster from 8b.
- **Non-renin** = everything else in the StarSpace cluster universe,
  i.e. `inc_nr ∪ dec_r ∪ dec_nr ∪ nd_r ∪ nd_nr`. These are regions that
  failed at least one of the two criteria above.

Background: genome-wide null (no explicit `bg` arg to GREAT). Using the
cluster union as background here would push foreground ≈ background for
the Non-renin set (it covers ~99% of the universe), collapsing every
hypergeometric p-value toward 1 and emptying the post-filter table. The
genome-wide null lets both sets produce real enrichment tables that the
weighted-rank step can score against each other.

## Load packages

```{r}
library(rGREAT)
```

## Build refined Renin / Non-renin GRanges from 8e intersection BEDs

Reload from disk so this chunk is independent of whether the in-memory
`inc_r` / `dec_nr` / etc. variables have been overwritten by the
top-N or knee-based subsetting in 8g.

```{r}
load_bed_8j <- function(path) {
  readGeneric(path, chr = 1, start = 2, end = 3, strand = NULL,
              meta.cols = NULL, keep.all.metadata = FALSE,
              zero.based = FALSE, remove.unusual = FALSE,
              header = FALSE, skip = 0, sep = "\t")
}

inc_r_full  <- load_bed_8j(file.path(bed_dir, "inc_r.bed"))
inc_nr_full <- load_bed_8j(file.path(bed_dir, "inc_nr.bed"))
dec_r_full  <- load_bed_8j(file.path(bed_dir, "dec_r.bed"))
dec_nr_full <- load_bed_8j(file.path(bed_dir, "dec_nr.bed"))
nd_r_full   <- load_bed_8j(file.path(bed_dir, "nd_r.bed"))
nd_nr_full  <- load_bed_8j(file.path(bed_dir, "nd_nr.bed"))

renin_refined_gr <- inc_r_full
non_renin_gr     <- c(inc_nr_full, dec_r_full, dec_nr_full,
                      nd_r_full,  nd_nr_full)

message(sprintf(
  "Refined sets: Renin = %s regions (inc_r), Non-renin = %s regions (everything else)",
  format(length(renin_refined_gr), big.mark = ","),
  format(length(non_renin_gr),     big.mark = ",")
))

ss_query <- GenomicRanges::GRangesList(
  Renin     = renin_refined_gr,
  Non_renin = non_renin_gr
)
```

## Run GREAT GO Biological Process enrichment

```{r}
ss_getBioEnrichment <- function(gr) {
  # No explicit background: when the Non-renin set covers ~99% of the
  # cluster universe, foreground ≈ background and every hypergeometric
  # p-value collapses toward 1, leaving the filter step with zero rows.
  # Testing each set against the genome-wide null lets both produce real
  # enrichment tables that the weighted-rank step can compare.
  job <- rGREAT::submitGreatJob(gr, species = "mm10")
  tb  <- rGREAT::getEnrichmentTables(job)
  tb$`GO Biological Process`
}

ss_enrichment <- lapply(names(ss_query), function(s) {
  ss_getBioEnrichment(ss_query[[s]])
})
names(ss_enrichment) <- names(ss_query)
```

## Weighted rank of renin-related GO terms

```{r}
ss_filterEnrichment <- function(df) {
  if (nrow(df) == 0) return(df)
  # rGREAT 1.x column for foreground gene hits is `Hyper_Observed_Gene_Hits`,
  # not `Hyper_Foreground_Gene_Hits`. If your rGREAT version differs, the
  # message below shows the actual columns so you can swap names.
  message("Enrichment columns: ", paste(colnames(df), collapse = ", "))
  keep <- df$Hyper_Adjp_BH <= 0.1 & df$Hyper_Observed_Gene_Hits >= 3
  go <- df[which(keep), , drop = FALSE]
  rownames(go) <- seq_len(nrow(go))
  go
}

ss_getReninGOTerms <- function(dfs) {
  keywords <- c(
    "notch", "cAMP",
    "angiotensin", "aldosterone", "renin",
    "vasculogenesis", "blood vessel",
    "arteriole", "blood pressure",
    "smooth muscle cell",
    "cell fate",
    "kidney", "renal", "nephron"
  )
  pattern <- paste(keywords, collapse = "|")
  matched <- unlist(lapply(dfs, function(df) {
    df$name[grepl(pattern, df$name, ignore.case = TRUE)]
  }))
  unique(matched)
}

ss_getRank <- function(df, group) {
  df <- df[order(df$Hyper_Adjp_BH), ]
  df$rank_pvalue <- seq_len(nrow(df))
  df <- df[order(df$Hyper_Fold_Enrichment), ]
  df$rank_fold <- seq_len(nrow(df))
  df$Rank <- rowMeans(df[, c("rank_pvalue", "rank_fold")])
  df$Group <- group
  df
}

ss_getRankWeight <- function(dfs, my_go_terms) {
  total <- max(sapply(dfs, nrow))
  for (i in seq_along(dfs)) {
    dfs[[i]]$rank_pvalue_weight <- (total - dfs[[i]]$rank_fold) / total
    dfs[[i]] <- subset(dfs[[i]], name %in% my_go_terms)
  }
  for (t in my_go_terms) {
    term_sum <- sum(sapply(dfs, function(df) {
      w <- df$rank_pvalue_weight[df$name == t]
      if (length(w) == 0) 0 else w
    }))
    for (i in seq_along(dfs)) {
      idx <- dfs[[i]]$name == t
      if (any(idx)) {
        dfs[[i]]$rank_weight[idx] <- dfs[[i]]$rank_pvalue_weight[idx] / term_sum
      }
    }
  }
  dfs
}
```

```{r}
ss_go        <- lapply(ss_enrichment, ss_filterEnrichment)

# Guard: empty enrichment table after filtering usually means foreground ≈
# background (collapsing hypergeometric p-values toward 1). Print sizes so
# this is easy to diagnose if it happens.
n_rows_per_set <- sapply(ss_go, nrow)
message("Filtered enrichment table sizes: ",
        paste(sprintf("%s = %d", names(n_rows_per_set), n_rows_per_set),
              collapse = ", "))
if (any(n_rows_per_set == 0)) {
  stop("Empty enrichment table(s): ",
       paste(names(n_rows_per_set)[n_rows_per_set == 0], collapse = ", "),
       ". Check the foreground vs background sizes in ss_getBioEnrichment().")
}

ss_renin_terms <- ss_getReninGOTerms(ss_go)

ss_rank_dfs <- lapply(names(ss_go), function(s) ss_getRank(ss_go[[s]], group = s))
names(ss_rank_dfs) <- names(ss_go)
ss_rank_dfs <- ss_getRankWeight(ss_rank_dfs, ss_renin_terms)

ss_plot_df <- do.call(rbind, ss_rank_dfs)
ss_plot_df <- ss_plot_df[!is.na(ss_plot_df$name), ]
rownames(ss_plot_df) <- NULL

ss_plot_df <- ss_plot_df %>%
  select(name, Group,
         Rank_weight      = rank_weight,
         Pval_rank        = rank_pvalue,
         Pval_rank_weight = rank_pvalue_weight)

ss_plot_df$Group <- factor(ss_plot_df$Group, levels = c("Renin", "Non_renin"))

ss_term_order <- ss_plot_df %>%
  filter(Group == "Renin") %>%
  arrange(desc(Rank_weight)) %>%
  mutate(labels = factor(name)) %>%
  head(50)

ss_plot_df <- filter(ss_plot_df, name %in% ss_term_order$name)
ss_new_order <- ss_plot_df %>%
  filter(Group == "Renin") %>%
  arrange(desc(Rank_weight)) %>%
  mutate(labels = factor(name))
```

## Plot stacked bar of weighted rank

```{r, fig.width = 8, fig.height = 7}
output_file <- file.path(plot_dir, "refined_renin_go_rank_weight_stacked_bar.svg")
dir.create(dirname(output_file), showWarnings = FALSE, recursive = TRUE)

n_terms <- length(unique(ss_plot_df$name))

p_ss_go <- ss_plot_df %>%
  mutate(labels = factor(name, levels = ss_new_order$labels)) %>%
  ggplot(aes(fill = Group, y = Rank_weight, x = labels)) +
  geom_bar(position = "fill", stat = "identity", width = 0.9) +
  scale_fill_manual(values = c(Renin = "#e1812c", Non_renin = "lightgrey")) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 6, color = "black"),
    axis.text.y = element_text(size = 6, color = "black"),
    axis.title  = element_text(size = 8, color = "black"),
    panel.grid  = element_blank()
  ) +
  labs(y = "Enrichment Weight", x = "GO Terms",
       title = "Refined Renin (inc_r) vs Non-renin (everything else) — renin-related GO terms")

ggsave(output_file, plot = p_ss_go,
       width = max(4, n_terms * 0.1), height = 7, units = "in")

p_ss_go
```
